### 2023-08-07 papers/Doubletiltconcept/doubletiltconcept.tex
Add first draft of double tilt concept paper

%% Beginning of file 'sample631.tex'
%%
%% Modified 2021 March
%%
%% This is a sample manuscript marked up using the
%% AASTeX v6.31 LaTeX 2e macros.
%%
%%
%% using aastex version 6.3
\documentclass[linenumbers]{aastex631}

%% The default is a single spaced, 10 point font, single spaced article.
%% There are 5 other style options available via an optional argument. They
%% can be invoked like this:
%%
%% \documentclass[arguments]{aastex631}
%% 
%% where the layout options are:
%%
%%  twocolumn   : two text columns, 10 point font, single spaced article.
%%                This is the most compact and represent the final published
%%                derived PDF copy of the accepted manuscript from the publisher
%%  manuscript  : one text column, 12 point font, double spaced article.
%%  preprint    : one text column, 12 point font, single spaced article.  
%%  preprint2   : two text columns, 12 point font, single spaced article.
%%  modern      : a stylish, single text column, 12 point font, article with
%% 		  wider left and right margins. This uses the Daniel
%% 		  Foreman-Mackey and David Hogg design.
%%  RNAAS       : Supresses an abstract. Originally for RNAAS manuscripts 
%%                but now that abstracts are required this is obsolete for
%%                AAS Journals. Authors might need it for other reasons. DO NOT
%%                use \begin{abstract} and \end{abstract} with this style.
%%
%% Note that you can submit to the AAS Journals in any of these 6 styles.
%%
%% There are other optional arguments one can invoke to allow other stylistic
%% actions. The available options are:
%%
%%   astrosymb    : Loads Astrosymb font and define \astrocommands. 
%%   tighten      : Makes baselineskip slightly smaller, only works with 
%%                  the twocolumn substyle.
%%   times        : uses times font instead of the default
%%   linenumbers  : turn on lineno package.
%%   trackchanges : required to see the revision mark up and print its output
%%   longauthor   : Do not use the more compressed footnote style (default) for 
%%                  the author/collaboration/affiliations. Instead print all
%%                  affiliation information after each name. Creates a much 
%%                  longer author list but may be desirable for short 
%%                  author papers.
%% twocolappendix : make 2 column appendix.
%%   anonymous    : Do not show the authors, affiliations and acknowledgments 
%%                  for dual anonymous review.
%%
%% these can be used in any combination, e.g.
%%
%% \documentclass[twocolumn,linenumbers,trackchanges]{aastex631}
%%
%% AASTeX v6.* now includes \hyperref support. While we have built in specific
%% defaults into the classfile you can manually override them with the
%% \hypersetup command. For example,
%%
%% \hypersetup{linkcolor=red,citecolor=green,filecolor=cyan,urlcolor=magenta}
%%
%% will change the color of the internal links to red, the links to the
%% bibliography to green, the file links to cyan, and the external links to
%% magenta. Additional information on \hyperref options can be found here:
%% https://www.tug.org/applications/hyperref/manual.html#x1-40003
%%
%% Note that in v6.3 "bookmarks" has been changed to "true" in hyperref
%% to improve the accessibility of the compiled pdf file.
%%
%% If you want to create your own macros, you can do so
%% using \newcommand. Your macros should appear before
%% the \begin{document} command.
%%
\newcommand{\vdag}{(v)^\dagger}
\newcommand\aastex{AAS\TeX}
\newcommand\latex{La\TeX}

%% Reintroduced the \received and \accepted commands from AASTeX v5.2
%\received{March 1, 2021}
%\revised{April 1, 2021}
%\accepted{\today}

%% Command to document which AAS Journal the manuscript was submitted to.
%% Adds "Submitted to " the argument.
%\submitjournal{PSJ}

%% For manuscript that include authors in collaborations, AASTeX v6.31
%% builds on the \collaboration command to allow greater freedom to 
%% keep the traditional author+affiliation information but only show
%% subsets. The \collaboration command now must appear AFTER the group
%% of authors in the collaboration and it takes TWO arguments. The last
%% is still the collaboration identifier. The text given in this
%% argument is what will be shown in the manuscript. The first argument
%% is the number of author above the \collaboration command to show with
%% the collaboration text. If there are authors that are not part of any
%% collaboration the \nocollaboration command is used. This command takes
%% one argument which is also the number of authors above to show. A
%% dashed line is shown to indicate no collaboration. This example manuscript
%% shows how these commands work to display specific set of authors 
%% on the front page.
%%
%% For manuscript without any need to use \collaboration the 
%% \AuthorCollaborationLimit command from v6.2 can still be used to 
%% show a subset of authors.
%
%\AuthorCollaborationLimit=2
%
%% will only show Schwarz & Muench on the front page of the manuscript
%% (assuming the \collaboration and \nocollaboration commands are
%% commented out).
%%
%% Note that all of the author will be shown in the published article.
%% This feature is meant to be used prior to acceptance to make the
%% front end of a long author article more manageable. Please do not use
%% this functionality for manuscripts with less than 20 authors. Conversely,
%% please do use this when the number of authors exceeds 40.
%%
%% Use \allauthors at the manuscript end to show the full author list.
%% This command should only be used with \AuthorCollaborationLimit is used.

%% The following command can be used to set the latex table counters.  It
%% is needed in this document because it uses a mix of latex tabular and
%% AASTeX deluxetables.  In general it should not be needed.
%\setcounter{table}{1}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%
%% The following section outlines numerous optional output that
%% can be displayed in the front matter or as running meta-data.
%%
%% If you wish, you may supply running head information, although
%% this information may be modified by the editorial offices.
\shorttitle{Double Tilted Rowland Spectrograph}
\shortauthors{G\"unther et al.}
%%
%% You can add a light gray and diagonal water-mark to the first page 
%% with this command:
%% \watermark{text}
%% where "text", e.g. DRAFT, is the text to appear.  If the text is 
%% long you can control the water-mark size with:
%% \setwatermarkfontsize{dimension}
%% where dimension is any recognized LaTeX dimension, e.g. pt, in, etc.
%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\graphicspath{{./}{figures/}}
%% This is the end of the preamble.  Indicate the beginning of the
%% manuscript itself with \begin{document}.

\begin{document}

\title{Concept of a Double Tilted Rowland Spectrograph for X-rays}

%% The new \altaffiliation can be used to indicate some secondary information
%% such as fellowships. This command produces a non-numeric footnote that is
%% set away from the numeric \affiliation footnotes.  NOTE that if an
%% \altaffiliation command is used it must come BEFORE the \affiliation call,
%% right after the \author command, in order to place the footnotes in
%% the proper location.
%%
%% Use \email to set provide email addresses. Each \email will appear on its
%% own line so you can put multiple email address in one \email call. A new
%% \correspondingauthor command is available in V6.31 to identify the
%% corresponding author of the manuscript. It is the author's responsibility
%% to make sure this name is also in the author list.
%%

\correspondingauthor{Hans Moritz G\"unther}
\email{hgunther@mit.edu}

\author[0000-0003-4243-2840]{Hans Moritz G{\"u}nther}
\affiliation{MIT Kavli Institute for Astrophysics and Space Research, 77 Massachusetts Avenue, Cambridge, MA 02139, USA}

\author{Add your name behind this, will sort later}
\affiliation{}

\begin{abstract}

This example manuscript is intended to serve as a tutorial and template for
authors to use when writing their own AAS Journal articles. The manuscript
includes a history of \aastex\ and documents the new features in the
previous versions as well as the bug fixes in version 6.31. This
manuscript includes many figure and table examples to illustrate these new
features.  Information on features not explicitly mentioned in the article
can be viewed in the manuscript comments or more extensive online
documentation. Authors are welcome replace the text, tables, figures, and
bibliography with their own and submit the resulting manuscript to the AAS
Journals peer review system.  The first lesson in the tutorial is to remind
authors that the AAS Journals, the Astrophysical Journal (ApJ), the
Astrophysical Journal Letters (ApJL), the Astronomical Journal (AJ), and
the Planetary Science Journal (PSJ) all have a 250 word limit for the 
abstract\footnote{Abstracts for Research Notes of the American Astronomical 
Society (RNAAS) are limited to 150 words}.  If you exceed this length the
Editorial office will ask you to shorten it. This abstract has 182 words.

\end{abstract}

%% Keywords should appear after the \end{abstract} command. 
%% The AAS Journals now uses Unified Astronomy Thesaurus concepts:
%% https://astrothesaurus.org
%% You will be asked to selected these concepts during the submission process
%% but this old "keyword" functionality is maintained in case authors want
%% to include these concepts in their preprints.
\keywords{Classical Novae (251) --- Ultraviolet astronomy(1736) --- History of astronomy(1868) --- Interdisciplinary astronomy(804)}

%% From the front matter, we move on to the body of the paper.
%% Sections are demarcated by \section and \subsection, respectively.
%% Observe the use of the LaTeX \label
%% command after the \subsection to give a symbolic KEY to the
%% subsection for cross-referencing in a \ref command.
%% You can use LaTeX's \ref and \label commands to keep track of
%% cross-references to sections, equations, tables, and figures.
%% That way, if you change the order of any elements, LaTeX will
%% automatically renumber them.
%%
%% We recommend that authors also use the natbib \citep
%% and \citet commands to identify citations.  The citations are
%% tied to the reference list via symbolic KEYs. The KEY corresponds
%% to the KEY in the \bibitem in the reference list below.

\section{Introduction} \label{sec:intro}
X-ray spectroscopy can deliver many important insights into the astrophysics of the hot and energetic universe that are not otherwise observable. In particular, high-resolution spectroscopy (with resolving power $> 500$ or better) can resolve closely-spaced emission lines which offer diagnostics such as density and temperature, reveal the near-edge fine structure in absorption lines, or allow us to detect the precense of faint and narrow absorption lines caused by the warm-hot intergalactic medium in front of bright background quasars to give just a few examples. In the X-ray band, such high resolving powers can be achieved with either microcalorimaters or dispersive elements such as crystals or diffraction gratings. Microcalorimeters offer a fixed absolute energy resolution $\Delta E$, which gives them good resolving power $R=E/\Delta E$ at large energies $E$ and consequently microcalorimeter detectors are planned for the imaging cameras for several of the next X-ray observatories\cite{charm,athena,lynx papers for detectors). However, diffraction gratings have a distinct advantage for softer X-rays below about 2~keV.

The two largest X-ray observatories in history both field diffraction gratings: The Reflection Grating Spectrometer (RGS) on XMM-Newton\cite{RGS} and both the High-Energy Transmission Grating Spectrograph (HETGS) and the Low-Energy Transmission Spectrograph (LETGS) on Chandra\cite{Carnizares2005,LETG}.
**Are each of these called spectrograph or spectrometer?**
Unlike microcalorimeters, diffraction gratings require an additional dispersive element (a transmission or a reflection grating) in the optical path of the instrument and dedicated detectors to capture the dispersed signal. One common layout for a diffraction spectrometer is the Rowlandtorus\cite{Beuermann:78}. In the plan defined bt the optical axis and the dispersion direction, the gratings are located at one side of a circle, the detectors on the other. This circle is commonly called the ``Rowland circle''. In three dimensions, the surface that defines the best grating positions, make up the surface of a torus, thus the name ``Rowland torus'' for this geometry (see Sect.~\ref{sect:onetorus} for a figure).

In this article, we present an optical layout for an instrument that stacks two or four different channels with distinct optical axes in such a way that each channel follows the Rowland torus prescription, yet at the same time all channels are imaged onto a common set of detectors. The general concept could be applied for a large range of focal lengths, grating constants, and other hardware properties. Thus, we discuss advantanges and challenges of this layout without constraint to any specific mission concept. However, we note that Arcus\cite{}, an X-ray spectroscopy mission selected for phase A study in a NASA midex (medium explorer) call in 2018 applied this layout. Specific details of the  Arcus performace, such as predicted effective area and resolving power are shown \citenum{}, but the general concept of a double tilted Rowland spectrograph with its advantages and drawbacks has not been discussed -- a void we attempt to fill with this article.

The layout of this article is as follows: First, we discuss the properties on the optical elements that are most influential for the design of a DTRS, namely mirror properties, gratings, and detectors in Sect.~\ref{sect:elements}. Then, the illustrate the optical layout, modifying the design of a traditional single-channel Rowland torus spectrograph to a DTRS. In section~\ref{sect:discussion}, we discuss advantages and disadvantages of our DTRS concept. We end with a short summart in section \ref{sect:summary}.

\section{Elements of an X-ray grating spectrograph}
\label{sect:elements}

\subsection{Mirror characteristics and sub-aperturing}

\subsection{Dispersive elements}
In prinicle works for any transmission grating, but makes most sense for CAT gratings, so discuss those

trasmission vs diffraction gratings - Casey 2020

Because X-ray can penetrate most material commonly used to manufacture gratings, X-ray instruments are designed to use the gratings either in transmission or, e.g.\ in the case of the RGS, at very small reflection angles close to the angle of total reflection. Thus, transmission and reflection gratings would be located at essentially the same position in the optical path.

\subsection{Detectors}
The dispersed spectrum is spread out far in the dispersion direction. Thus, the ideal detector is much longer in one direction than in the other. Individual detector elements (e.g.\ CCDs) can be tiled, but to avoid chip gaps, it is beneficial to keep the number of elements low. In our DTRS design several spectral traces are positioned in parallel (section~\ref{sect:opticallayout), so the detector has to be position sensitive in 2D to separate individual channels.  Because ir covers a large area than in direct imaging observations, a detector with moderate cost and power requirements per surface area is well-suited for this design. The detector needs to be slightly curved to follow the Rowland torus. If it is made up of smaller individually flat elements, those elements can be arranged tangential to a circle.

The dispersed spectrum contains fewer photons than direct imaging with the same collecting area and spreads this signal out over a large physical region. Thus, the counts rates are lower. Detectors that might by at risk of pile-up in direct observations can still be used for the dispersed signal. On the other hand, the larger regions needed to extract the signal means that background events (internal background and diffuse astrophysical emission) are more important than in direct imaging observations. Similarly, fast timing resolution is typically not needed, because the lower signal means that the detected photons need to be binned up into longer time bins for analysis than a direct imaging observation. Intrinsic energy resolution is required to both suppress the background by filtering out events that are incompatible with wavelengths of the dispersed signal at any particular location and to separate grating orders, where photons of different orders are diffracted to the same position (e.g.\ the sixth order for the O~{\sc vii} forbidden line of the He-like triplet at 2.21~nm and the seventh order of 1.94~mn photons, close to the O~{\sc viii} Lyman$\alpha$ line).

In our DTRS design, the direct images of all channels are positioned on detectors to provide an accurate wavelength calibration, but also because the high-enery part of the spectrum, which is not efficiently dispersed by gratings can provide valuable context for the scientific interpretation of the high-resolution grating spectrum. Thus, the dector type chosen should also be able to handle direct imaging, albeit at a much lower count rate than in a direct imaging instrument without dispersive elements.

**Ask Eric is he knows a paper to cite for comparison of detector types. Lynx?**

\section{Optical Layout}
In this section, we describe the optical layout for a DTRS. We start from the commonly used layout with a single spectrsocopy channel as it is employed in Chandra/LETGS or XMM/RGS. We then show how this design can be modified step-by-step and what considerations are important when chosing the specific parameters for a design. We illustrastrate the concept with ray-trace simulations. Those simulations are performed using a real ray-trace with the marxs code, but the dispersion angle (which is set by the energy of the photons and the grating constant of the grating) is chosen to be larger than feasible for a real instrument. This is done purely to separate the direct light (zero order) and the dispersed light (positive and negative orders) better in the images, so that the concept can be explained more clearly. Static images are embedded in the article, but we strongly encourage the reader explore the interactive 3D models embedded in the electronic version of the publication (might require Acrobat reader).

f, d, sketches

\subsection{One Rowland Torus with central optical axis}
\label{sect:onetorus}
chip gaps chandra, dither

%\begin{figure}
%  \begin{interactive}{js}{file_wth_arll_components.tar/zip}
%  figure call (e.g.\plotone, \includegraphics, etc.)
%  \end{interactive}
%  \caption{Description of the example figure and the interactive portion which includes how to use the interactive functionality.}
%\end{figure}

\subsection{One tilted Rowland Torus}

This torus is tilted Ralfs tild paper
\cite{doi:10.1117/12.856482}

\subsection{Double tilted Rowland Torus}
For R > r signal dips below plane, fish form, two points, symmetry

two opticla axes,
offset chip gaps
spectroscopic vs imganing focus unless I;ve said that above.

\subsection{Channel offsets}

\subsubsection{Channel offsets in the cross-dispersion direction}

\subsubsection{Channel offsets in the dispersion direction}

\section{Discussion}

\subsection{Subaperturing space considerations, and resolving power}
lower mirror requirements, thus chepaer cost for same resolving power, but can use wdge only, stack channels to get square aperture and get back effective area

\subsection{Number of CCDs}
2 camerase, different directions. Need to get zeroth orders anyway (high-energy, wavelength scale), but more CCDs than just one directions (Lynx), make good use of rea-estate

\subsection{Robustness}
relaxes alignment because multiple channels (thus lower cost and technical risk), resilience because independent channels on different CCDS (see XMM loss of RGS chip), check for systemics in calibration because independent channels, chip gaps

wavelength self-calibration (does not need DTRS, but works there, too). is order sorting and several orders ad one energy

\subsection{Other advantages}

Zero order seen, for direct observation (hard X-rays) and calibration

\subsection{Drawbacks}
increased background due ot larger CCD region, less source-free region for background.instrum analysis, coplexity of analysis (numberof spectra!), sprectra coupled (cross-dispersion, read-out streaks for traditional CCDs, wings and energy resolution), does not work for crowded fields, fields with extended emission (galactic center)

\subsection{Similar concepts}

The Far Ultraviolett Spectroscopic Explorer (FUSE) \cite{2000ApJ...538L...1M} faced a design challenge similar to what we discuss here for X-ray instruments. FUSE also has four spectroscopic channels with gratings arranged in a Rowland geometry. Unlike in X-rays, UV gratings can operate in reflection close to normal incidence. The reflection gratings are thus positioned perpendicular to the path of the rays, which means that are almost tangential to the surface of the Rowland torus. Thus, a small number of gratings (just one per channel in the extreme case) is sufficient and, even if the grating itself is flat, it never deviates much from the surface of the Rowland torus. Since UV light has a longer wavelength, it can be dispersed to larger angles increasing the spectral resolving power. In FUSE, detectors just cover the dispersed signal, and not the direct beam (zeroth order) and detectors for the different channels are independent of each other.

Our design of a DTRS places the direct beam for all channels on a detector. This way, the position of the pointing can be constantly monitored and determined from the observed data. If the pointing position drifts and the optical channels are not fully aligned, this can be seen in the science data. In contrast, FUSE has performed observations where the target was not in the field-of-view of all channels, but the amount of the exposure time lost is unclear \cite{2000ApJ...538L...1M}.

\section{Summary}

A specialized layout for a mission with point source spectroscopy goal, not a general observatory design

Arcus (need ot bring is up finally - or have a section on "Application" obove?)

\begin{acknowledgements}
Support for this work was provided in part through NASA grant NNX17AG43G and Smithsonian Astrophysical Observatory (SAO)
contract SV3-73016 to MIT for support of the {\em Chandra} X-Ray Center (CXC),
which is operated by SAO for and on behalf of NASA under contract NAS8-03060.

\end{acknowledgements}

\software{marxs\cite{}, AstroPy \citep{2013A&A...558A..33A,2018AJ....156..123A}, NumPy \citep{van2011numpy,harris2020array}, Matplotlib \citep{Hunter:2007},IPython\cite{IPython}}

\bibliography{bib}{}
\bibliographystyle{aasjournal}

%% Include this line if you are using the \added, \replaced, \deleted
%% commands to see a summary list of all changes at the end of the article.
%\listofchanges

\end{document}

### 2023-08-07 notebooks/production/grid2designtorus.py

@@ -1,128 +1,0 @@
-''' '''
-import configparser
-import os
-import time
-import sys
-
-import numpy as np
-import astropy.units as u
-from astropy.table import Table
-from openpyxl import load_workbook
-
-from marxs.missions.arcus.defaults import DefaultSource, DefaultPointing
-from marxs.missions.arcus.generate_rowland import make_rowland_from_d_BF_R_f
-from marxs.missions.arcus.arcus import defaultconf
-from marxs.missions.arcus.ralfgrating import RegularGrid
-from marxs.missions.arcus.spo import spos_large, spos_large_pos4d
-from marxs.missions.arcus.arcus import PerfectArcus
-from marxs.missions.mitsnl.catgrating import InterpolateEfficiencyTable
-
-
-cfgpath = [os.path.join(os.path.dirname(sys.modules[__name__].__file__), '..', 'site.cfg')]
-'Path list to search for configuration files.'
-
-def get_path(name):
-    '''Get path name info from site.cfg file in root directory.
-
-    If a path does not exist, it will be created.
-    '''
-    conf = configparser.ConfigParser()
-    cfgfile = conf.read(cfgpath)
-    if not cfgfile:
-        raise Exception("No config file with path specifications found. File must be called 'site.py' and be located in one of the following directories: {}".format(cfgpath))
-    path = conf.get("Path", name)
-    if not os.path.exists(path):
-        os.makedirs(path)
-    return path
-
-
-conf = defaultconf.copy()
-conf['spo_geom'] = spos_large
-conf['spo_pos4d'] = spos_large_pos4d
-
-class BigSPOArcus(PerfectArcus):
-    gratings_class = RegularGrid
-
-
-def sheet_to_table(sheet):
-    # construct column names
-    colnames = [sheet['B8'].value, sheet['C8'].value] + [cell.value for cell in sheet[9][3:]]
-    # read data
-    tab = Table([a for a in sheet.iter_cols(min_row=10, min_col=2, values_only=True)],
-                names=colnames)
-    # add units
-    tab['wavelength'].unit = u.Unit(sheet['B9'].value.replace('[', '').replace(']', ''))
-    tab['angle'].unit = u.Unit(sheet['C9'].value.replace('[', '').replace(']', ''))
-    # Read meta data, which is saved at fixed location
-    for i in [2,3,4,5]:
[…]

### 2023-08-07 production/paralleljobs/RAeff_perfect_slurm.sh
Move production of R/Aeff grid to slurm

@@ -1,0 +1,7 @@
+#!/bin/bash
+#SBATCH --job-name=RAeff
+#SBATCH --ntasks=1
+#SBATCH --cpus-per-task=1
+#SBATCH --array=0-389
+conda activate arcus
+python RAeff_single_simulation.py /work/submit/hgunther/ARCUS/raysperfectRAeff/ ${SLURM_ARRAY_TASK_ID} --perfect

### 2023-08-07 production/paralleljobs/RAeff_single_simulation.py

@@ -1,0 +1,128 @@
+# Licensed under GPL version 3 - see LICENSE.rst
+'''Run a single monoenenergetic simulation of Arcus.
+
+This is indented for running grids of with different energies
+to build up effective area curves or ARF/RMF. Such a calculation
+is "embarrassingly parallel" and can easily be run as a number of
+different Python processes. To facilitate that use, this is written
+as a Python script, not an importable module.
+
+Further Optimization
+====================
+Further optimization of this script is possible to reduce file sizes
+and memory needs during a run. However, at this point it seems that
+that is not worth the effort, so ideas are just kept below as notes
+should this become important in the future.
+
+Do no generate local coordinate names
+-------------------------------------
+The easiest way to achieve that is probably to change the defaults in
+the classes before creating the instrument objects:
+
+   >>> from marxs.optics.scatter import RadialMirrorScatter
+   >>> RadialMirrorScatter.inplanescattercol = None
+   >>> RadialMirrorScatter.perpplanescattercol = None
+
+And similarly for `` ScatterPerChannel.loc_coos_name = None``,
+``PerfectLens.loc_coos_name = None``,
+``FlatGrating.loc_coos_name = None``
+``FlatGrating.blaze_name = None``,
+``L1.blaze_name = None``.
+
+Have fewer detectors
+--------------------
+`marxs.missions.arcus.arcus.Arcus` has several detectors for analysis,
+but we probably don't need all of them here. Could save some time by
+not having all of them.
+'''
+
+import sys
+from os.path import join, exists
+import time
+import argparse
+import numpy as np
+import astropy.units as u
+from marxs.missions.arcus.arcus import (Arcus, PerfectArcus,
+                                        defaultconf as conf)
+from marxs.missions.arcus.defaults import DefaultSource, DefaultPointing
+from marxs.optics import RectangleAperture
+from marxs.missions.arcus.utils import id_num_offset
+
+parser = argparse.ArgumentParser(
+                    description = 'Run a single simulation in a wavelength grid that can later we used to generate R/Aeff curves or even ARFs and RMFs.\n' + 
+                       'The interface is specifically designed to make is easy to iterate over this by changing just the "i" parameter' +
+                       'which helps with scheduling systems list slurm job arrays.')
[…]

### 2023-08-07 production/paralleljobs/RAeff_slurm.sh

@@ -1,0 +1,6 @@
+#!/bin/bash
+#SBATCH --job-name=RAeff
+#SBATCH --ntasks=1
+#SBATCH --cpus-per-task=1
+#SBATCH --array=0-389
+python RAeff_single_simulation.py /work/submit/hgunther/ARCUS/raysRAeff/ ${SLURM_ARRAY_TASK_ID}

### 2023-08-07 production/paralleljobs/arf_rmf_slurm.sh

@@ -1,0 +1,8 @@
+#!/bin/bash
+#SBATCH --job-name=arf_rmf
+#SBATCH --ntasks=1
+#SBATCH --cpus-per-task=1
+#SBATCH --array=0-58
+python ../RAeff_single_simulation.py /work/submit/hgunther/ARCUS/arfrmf/ ${SLURM_ARRAY_TASK_ID} \
+  --wave_lo=1.5 --wave_hi=60. --wave_step=1.0 \
+  -n 500000 --channels 1

### 2023-08-07 production/simulations.py

@@ -1,46 +1,0 @@
-from os.path import join
-import time
-import numpy as np
-import astropy.units as u
-from arcus.arcus import (Arcus, PerfectArcus, defaultconf)
-from arcus.defaults import DefaultSource, DefaultPointing
-from arcus.analysis.grid import aeffRfromraygrid, csv_per_order
-from utils import get_path
-
-n_photons = 400000
-wave = np.arange(1.5, 50., 0.15) * u.Angstrom
-
-energies = wave.to(u.keV, equivalencies=u.spectral()).value
-
-mypointing = DefaultPointing()
-
-for instrum, path in zip([#Arcus(),
-                          PerfectArcus()
-                      ],
-                         [#'raygrid',
-                          'raygrid-perfect'
-                      ]):
-    outpath = get_path(path)
-
-    for i in range(len(wave)):
-        print('{0}/{1} = {2}'.format(i + 1, len(energies), time.ctime()))
-        mysource = DefaultSource(energy=energies[i])
-
-        photons = mysource.generate_photons(n_photons)
-        photons = mypointing(photons)
-        photons = instrum(photons)
-
-        photons.write(join(outpath,
-                            'wave{0:05.2f}.fits'.format(wave.value[i])),
-                      overwrite=True)
-    aeffRfromraygrid(outpath, instrum.elements[0], defaultconf,
-                     join(get_path('arcus'), path + 'RAeff.fits'))
-
-csv_per_order(join(get_path('arcus'), 'raygridRAeff.fits'), 'Aeff4',
-              join(get_path('figures'), 'Aeff.csv'))
-csv_per_order(join(get_path('arcus'), 'raygridRAeff.fits'), 'R4',
-              join(get_path('figures'), 'R.csv'))
-csv_per_order(join(get_path('arcus'), 'raygrid-perfectRAeff.fits'), 'Aeff4',
-              join(get_path('figures'), 'Aeff-perfect.csv'))
-csv_per_order(join(get_path('arcus'), 'raygrid-perfectRAeff.fits'), 'R4',
-              join(get_path('figures'), 'R-perfect.csv'))

### 2023-08-07 production/single_RAeff_simulation.py

@@ -1,58 +1,0 @@
-from os.path import join
-import time
-import argparse
-import numpy as np
-import astropy.units as u
-from marxs.missions.arcus.arcus import (Arcus, PerfectArcus)
-from marxs.missions.arcus.defaults import DefaultSource, DefaultPointing
-#from arcus.analysis.grid import aeffRfromraygrid, csv_per_order
-
-parser = argparse.ArgumentParser(
-                    description = 'Run a single simulation in a wavelength grid that can later we used to generate R/Aeff curves or even ARFs and RMFs.\n' + 
-                       'The interface is specifically designed to make is easy to iterate over this by changing just the "i" parameter' +
-                       'which helps with scheduling systems list slurm job arrays.',
-                    epilog = 'Text at the bottom of help')
-parser.add_argument('outpath', help='Path to write simulation output') 
-parser.add_argument('i', type=int, help='index in wavelength array to select wavelength for this simulation') 
-parser.add_argument('-n', '--n_photons', default=100000, type=int,
-                    help='number of photons')
-parser.add_argument('--wave_lo', default=1.5, type=float,
-                    help='lower end of wavelength range in Ang')
-parser.add_argument('--wave_hi', default=50., type=float,
-                    help='upper end of wavelength range in Ang')
-parser.add_argument('--wave_step', default=0.15, type=float,
-                    help='step size')
-parser.add_argument('--perfect',
-                    action='store_true',
-                    help='Use perfect Arcus without default misalignments') 
-args = parser.parse_args()
-
-wave = np.arange(args.wave_lo, args.wave_hi, args.wave_step) * u.Angstrom
-energies = wave.to(u.keV, equivalencies=u.spectral())
-pointing = DefaultPointing()
-
-instrument = PerfectArcus() if args.perfect else Arcus()
-print(f'{time.ctime()} - begin simulation for {wave[args.i]:05.2f} Ang')
-source = DefaultSource(energy=energies[args.i])
-photons = source.generate_photons(args.n_photons * u.s)
-photons = pointing(photons)
-photons = instrument(photons)
-
-# There is no way we need all those columns. Waste of space.
-photons.keep_columns(
- [
- 'energy',   # Could turn into a keyword since it's the same for each row
- 'probability',
- 'xou',
- 'facet',  # Probably don't need, but keep for now
- 'order',
- 'order_L1',
[…]

### 2024-02-26 production/RAeff_analyse_grid.py
Changes to R/Aeff production scripts

@@ -1,0 +1,15 @@
+import numpy as np
+from marxs.missions.arcus.analyze_grid import aeffRfromraygrid
+from marxs.missions.arcus.arcus import defaultconf as conf
+
+out = aeffRfromraygrid('raysRAeffperfect/',
+                       conf,
+                       orders=np.arange(-20, 5),
+                      allow_inconsistent_meta=False)
+out.write('raygrid-perfectRAeff.fits')
+
+out = aeffRfromraygrid('raysRAeff/',
+                       conf,
+                       orders=np.arange(-20, 5),
+                      allow_inconsistent_meta=False)
+out.write('raygridRAeff.fits')

### 2024-02-26 papers/Doubletiltconcept/doubletiltconcept.tex
Doubletilt paper to first round of Co-authors

@@
 %%
-\newcommand{\vdag}{(v)^\dagger}
-\newcommand\aastex{AAS\TeX}
-\newcommand\latex{La\TeX}
 
@@
 
+\usepackage{amsmath}
+
 \begin{document}
@@
 \begin{abstract}
-
-This example manuscript is intended to serve as a tutorial and template for
-authors to use when writing their own AAS Journal articles. The manuscript
-includes a history of \aastex\ and documents the new features in the
-previous versions as well as the bug fixes in version 6.31. This
-manuscript includes many figure and table examples to illustrate these new
-features.  Information on features not explicitly mentioned in the article
-can be viewed in the manuscript comments or more extensive online
-documentation. Authors are welcome replace the text, tables, figures, and
-bibliography with their own and submit the resulting manuscript to the AAS
-Journals peer review system.  The first lesson in the tutorial is to remind
-authors that the AAS Journals, the Astrophysical Journal (ApJ), the
-Astrophysical Journal Letters (ApJL), the Astronomical Journal (AJ), and
-the Planetary Science Journal (PSJ) all have a 250 word limit for the 
-abstract\footnote{Abstracts for Research Notes of the American Astronomical 
-Society (RNAAS) are limited to 150 words}.  If you exceed this length the
-Editorial office will ask you to shorten it. This abstract has 182 words.
+High-resolution spectroscopy in soft X-rays ($<2$~keV) requires diffractive elements to resolve any astrophysically relevant diagnostics, such as closely spaced lines, weak absorption lines, or line profiles.
+For transmission gratings, the Rowland torus geometry describes how gratings and detectors need to be positioned to optimize the spectral resolving power.
+We describe how an on-axis Rowland geometry can be tilted to accommodate blazed gratings.
+In this geometry, two channels with separate optical axes can share the same detectors (double tilted Rowland spectrograph, DTRS).
+Small offsets between the channels can mitigate the effect of chip gaps and reduce the alignment requirements during the construction of the instrument.
+The DTRS concept is especially useful for sub-apertured mirrors, because it allows an effective use of space in the entrance aperture of a spacecraft.
+One mission that applies this concept is the Arcus probe.
 
@@
 %% to include these concepts in their preprints.
-\keywords{Classical Novae (251) --- Ultraviolet astronomy(1736) --- History of astronomy(1868) --- Interdisciplinary astronomy(804)}
+\keywords{Astronomical instrumentation (799) --- Spectrometers (1554) --- X-ray astronomy (1810)}
 
@@
 \section{Introduction} \label{sec:intro}
-X-ray spectroscopy can deliver many important insights into the astrophysics of the hot and energetic universe that are not otherwise observable.
-In particular, high-resolution spectroscopy (with resolving power $> 500$ or better) can resolve closely-spaced emission lines which offer diagnostics such as density and temperature, reveal the near-edge fine structure in absorption lines, or allow us to detect the precense of faint and narrow absorption lines caused by the warm-hot intergalactic medium in front of bright background quasars to give just a few examples.
+X-ray spectroscopy can deliver many insights into the astrophysics of the hot and energetic universe that are not otherwise observable.
+In particular, high-resolution spectroscopy (with resolving power $> 500$ or better) can resolve closely-spaced emission lines which offer diagnostics such as density and temperature, reveal the near-edge fine structure in absorption lines, or allow us to detect the presence of faint and narrow absorption lines caused by the warm-hot intergalactic medium in front of bright background quasars to give just a few examples.
 In the X-ray band, such high resolving powers can be achieved with either microcalorimaters or dispersive elements such as crystals or diffraction gratings.
-Microcalorimeters offer a fixed absolute energy resolution $\Delta E$, which gives them good resolving power $R=E/\Delta E$ at large energies $E$ and consequently microcalorimeter detectors are planned for the imaging cameras for several of the next X-ray observatories\cite{charm,athena,lynx papers for detectors).
+Microcalorimeters offer a fixed absolute energy resolution $\Delta E$, which gives them good resolving power $R=E/\Delta E$ at large energies $E$ and consequently a microcalorimeter has been launched on Astro-E/Hitomi/SXS \citep{2014SPIE.9144E..2AM} and XRISM/Resolve \citep{2018JATIS...4a1214K} and this technology is also considered for other large missions in development such as Athena \citep{2014SPIE.9144E..2LR} and Lynx \citep{2019JATIS...5b1017B}.
+\citet{2016Natur.535..117H} used the microcalorimeter data from Astro-E/Hitomi to measure the line-of-sight velocity dispersion of \ion{Fe}{26} in the Perseus cluster and found that the turbulent pressure is only a few percent of the thermodynamic pressure in that region.
 However, diffraction gratings have a distinct advantage for softer X-rays below about 2~keV.
 
-The two largest X-ray observatories in history both field diffraction gratings: The Reflection Grating Spectrometer (RGS) on XMM-Newton\cite{RGS} and both the High-Energy Transmission Grating Spectrograph (HETGS) and the Low-Energy Transmission Spectrograph (LETGS) on Chandra\cite{Carnizares2005,LETG}.
-**Are each of these called spectrograph or spectrometer?**
+The two largest X-ray observatories in history both field diffraction gratings: The Reflection Grating Spectrometer (RGS) on XMM-Newton \citep{2001A&A...365L...7D} and both the High-Energy Transmission Grating  \citep[HETG,][]{2005PASP..117.1144C} and the Low-Energy Transmission Grating \citep[LETG,][]{1997SPIE.3113..172P} on Chandra.
 Unlike microcalorimeters, diffraction gratings require an additional dispersive element (a transmission or a reflection grating) in the optical path of the instrument and dedicated detectors to capture the dispersed signal.
-One common layout for a diffraction spectrometer is the Rowlandtorus\cite{Beuermann:78}.
-In the plan defined bt the optical axis and the dispersion direction, the gratings are located at one side of a circle, the detectors on the other.
+One common layout for a diffraction spectrometer is the Rowlandtorus \citep{Beuermann:78}.
+In the plane defined by the optical axis and the dispersion direction, the gratings are located at one side of a circle, the detectors on the other.
 This circle is commonly called the ``Rowland circle''.
-In three dimensions, the surface that defines the best grating positions, make up the surface of a torus, thus the name ``Rowland torus'' for this geometry (see Sect.~\ref{sect:onetorus} for a figure).
+In three dimensions, the surface that defines the best grating positions has the shape of a torus, thus the name ``Rowland torus'' for this geometry (see Sect.~\ref{sect:onetorus} for a figure).
 
@@
 The general concept could be applied for a large range of focal lengths, grating constants, and other hardware properties.
-Thus, we discuss advantanges and challenges of this layout without constraint to any specific mission concept.
-However, we note that Arcus\cite{}, an X-ray spectroscopy mission selected for phase A study in a NASA midex (medium explorer) call in 2018 applied this layout.
-Specific details of the  Arcus performace, such as predicted effective area and resolving power are shown \citenum{}, but the general concept of a double tilted Rowland spectrograph with its advantages and drawbacks has not been discussed -- a void we attempt to fill with this article.
+Thus, we discuss advantages and challenges of this layout in general.
+However, we note that Arcus \citep{2023SPIE12678E..0ES}, an X-ray spectroscopy mission proposed for NASA midex (medium explorer) and probe calls, applies this layout.
+Specific details of the  Arcus performance, such as predicted effective area and resolving power are shown in \citet{2018SPIE10699E..6FG,2023SPIE12678E..1DG}, but the general concept of a double tilted Rowland spectrograph (DTRS) with its advantages and drawbacks has not been discussed -- a void we attempt to fill with this article.
 
-The layout of this article is as follows: First, we discuss the properties on the optical elements that are most influential for the design of a DTRS, namely mirror properties, gratings, and detectors in Sect.~\ref{sect:elements}.
-Then, the illustrate the optical layout, modifying the design of a traditional single-channel Rowland torus spectrograph to a DTRS.
-In section~\ref{sect:discussion}, we discuss advantages and disadvantages of our DTRS concept.
-We end with a short summart in section \ref{sect:summary}.
+The structure of this article is as follows: First, we discuss the properties on the optical elements that are most influential for the design of a DTRS, namely mirror properties, gratings, and detectors in Sect.~\ref{sect:elements}.
+Then, we illustrate the optical layout, modifying the design of a traditional single-channel Rowland torus spectrograph to a DTRS.
+In section~\ref{sect:discussion}, we discuss advantages and disadvantages of our DTRS concept in general and for the example of Arcus in particular (section~\ref{sect:applications}).
+We end with a short summary in section \ref{sect:summary}.
 
@@
 \label{sect:elements}
+An X-ray spectrograph using the Rowland layout consists of a focussing mirror, diffractive gratings, and a detector.
 
 \subsection{Mirror characteristics and sub-aperturing}
+The concept of a DTRS can be implemented with different mirror technologies.
+The size of the collecting area is to first order proportional to the number of photons that the spectrograph can detect, and the spectral resolving power $\Delta \lambda / \lambda$ increases when the mirror point-spread-function (PSF) decreases.
+Depending on the type of mirror chosen, the PSF can be dominated by the alignment error between mirror shells, in particular for mirrors that consist of hundreds of elements.
+In this case, X-rays are offset from the focus point both in the plane of reflection and perpendicular to it.
+On the other hand, mirrors can also have a figure error, which is the deviation of the mirror surface from the ideal shape.
+If the PSF is dominated by the figure error or scattering by particulates on the mirror surface, then scatter in the plane of reflection is typically larger than the out-of-plane scatter.
+In this case, sub-aperturing of the aperture can improve the spectral resolving power.
+This is illustrated in Fig.~\ref{fig:subaperture}.
+
+\begin{figure*}
+\plotone{subaperture.pdf}
+\caption{
+    Illustration of the concept of sub-aperturing.
+All panels show a view from the universe looking down onto the telescope.
+\emph{left}: Each dot is a photon shown as it passes through a diffraction gratings.
+Photons are colored according to their polar angle in this panel.
+\emph{middle and right:} Photons detected in the zeroth and first diffraction order on the detector, respectively.
+The dispersion direction is from left to right ($y$ coordinate), the distance from the focal point at $(y, z) = (0, 0)$ is shown.
+Each photon is shown in the same color as in the left panel.
+Photons passing mirror and grating close to $z=0$ (dark purple) are spread out along the dispersion direction ($y$ coordinate); photons passing mirror and gratings close to $y=0$ are more concentrated along the dispersion direction.
+Sub-aperturing means that the mirror and gratings do not cover a full circle, but only regions marked in yellow and green in the left panel, so that the resulting photon distributions on the detector are narrower in dispersion direction.
+Sub-aperturing thus reduces the PSF in the dispersion direction, which increases the spectral resolving power.}
+\label{fig:subaperture}
+\end{figure*}
 
 \subsection{Dispersive elements}
-In prinicle works for any transmission grating, but makes most sense for CAT gratings, so discuss those
-
-trasmission vs diffraction gratings - Casey 2020
-
-Because X-ray can penetrate most material commonly used to manufacture gratings, X-ray instruments are designed to use the gratings either in transmission or, e.g.\ in the case of the RGS, at very small reflection angles close to the angle of total reflection.
-Thus, transmission and reflection gratings would be located at essentially the same position in the optical path.
+In a Rowland torus geometry, diffraction gratings are placed on the surface of the Rowland torus.
+They diffract light according to the grating equation for photons arriving perpendicular to the grating surface:
+\begin{equation}
+n \lambda = d \sin \theta \label{eqn:diffraction}
+\end{equation}
+where $\lambda$ is the wavelength of the light, $n$ is the order of the diffraction, $d$ is the grating constant, and $\theta$ is the angle between the incoming light and the normal to the grating.
+(The equation can be rewritten for rays hitting the grating at arbitrary angles, but for the purpose of this conceptual discussion that is not needed.)
+Typically, these transmission gratings are flat and only a few cm$^2$ in size; the spacing between gratings bars $d$ is the same for all gratings and does not vary within a grating.
+While a flat grating necessarily deviates from the curved surface of a torus, in today's instruments those deviations are small enough that they do not impact the resolving power of the spectrograph significantly.
+Larger gratings would increase the effective area of the instrument because less space is needed for frames and other mounting structures that hold the gratings in place, but they would also deviate more from the surface of the torus.
+\citet{2020SPIE11444E..88G} explored bending the gratings in one dimension or varying the distance between grating bars over a grating to compensate for the deviations from the ideal shape in simulations.
+However, these concepts have not been implemented in a hardware yet.
 
@@
 Thus, the ideal detector is much longer in one direction than in the other.
-Individual detector elements (e.g.\ CCDs) can be tiled, but to avoid chip gaps, it is beneficial to keep the number of elements low.
-In our DTRS design several spectral traces are positioned in parallel (section~\ref{sect:opticallayout), so the detector has to be position sensitive in 2D to separate individual channels.
-Because ir covers a large area than in direct imaging observations, a detector with moderate cost and power requirements per surface area is well-suited for this design.
-The detector needs to be slightly curved to follow the Rowland torus.
-If it is made up of smaller individually flat elements, those elements can be arranged tangential to a circle.
+Individual detector elements (e.g.\ CCDs) can be tiled, but to avoid chip gaps where photons are lost and gaps appear in the extracted spectum, it is beneficial to keep the number of elements low.
+In our DTRS design several spectral traces are positioned in parallel (section~\ref{sect:opticallayout}), so the detector has to be position sensitive in 2D to separate individual channels.
+Because it covers a larger area than in direct imaging observations, a detector with moderate cost and power requirements per surface area is well-suited for this design.
+The detector should follow the shape of the Rowland circle.
+This can be achieved by using curved detectors or by tiling smaller flat detector elements where each element is tangential to the Rowland circle.
 
-The dispersed spectrum contains fewer photons than direct imaging with the same collecting area and spreads this signal out over a large physical region.
-Thus, the counts rates are lower.
-Detectors that might by at risk of pile-up in direct observations can still be used for the dispersed signal.
-On the other hand, the larger regions needed to extract the signal means that background events (internal background and diffuse astrophysical emission) are more important than in direct imaging observations.
-Similarly, fast timing resolution is typically not needed, because the lower signal means that the detected photons need to be binned up into longer time bins for analysis than a direct imaging observation.
-Intrinsic energy resolution is required to both suppress the background by filtering out events that are incompatible with wavelengths of the dispersed signal at any particular location and to separate grating orders, where photons of different orders are diffracted to the same position (e.g.\ the sixth order for the O~{\sc vii} forbidden line of the He-like triplet at 2.21~nm and the seventh order of 1.94~mn photons, close to the O~{\sc viii} Lyman$\alpha$ line).
+The gratings spread out the signal over a large physical region on the detector.
+Thus, the counts rates per pixel are much lower than they would be in an imaging observation.
+Detectors that might be at risk of pile-up in direct observations can still be used for the dispersed signal.
+On the other hand, the larger regions needed to extract the signal imply that background events (internal background and diffuse astrophysical emission) are more important than in direct imaging observations.
+Similarly, fast timing resolution is typically not needed, because the lower signal means that the detected photons need to be binned up into longer time bins for analysis anyway.
+Intrinsic energy resolution is required to both suppress the background by filtering out events that are incompatible with wavelengths of the dispersed signal at any particular location and to separate grating orders, where photons of different orders are diffracted to the same position (e.g.\ the sixth order for the O~{\sc vii} forbidden line of the He-like triplet at 2.21~nm and the seventh order of 1.94~mn photons, close to the O~{\sc viii} Lyman$\alpha$ line end up at almost the same $\theta$ according to equation~\ref{eqn:diffraction}).
 
-In our DTRS design, the direct images of all channels are positioned on detectors to provide an accurate wavelength calibration, but also because the high-enery part of the spectrum, which is not efficiently dispersed by gratings can provide valuable context for the scientific interpretation of the high-resolution grating spectrum.
-Thus, the dector type chosen should also be able to handle direct imaging, albeit at a much lower count rate than in a direct imaging instrument without dispersive elements.
-
-**Ask Eric is he knows a paper to cite for comparison of detector types. Lynx?**
+In our DTRS design, the direct images of all channels are positioned on detectors to provide an accurate wavelength calibration, but also because the high-energy part of the spectrum, which is not efficiently dispersed by gratings, can provide valuable context for the scientific interpretation of the high-resolution grating spectrum.
+Thus, the detector type chosen should also be able to handle direct imaging, albeit at a lower count rate than in a direct imaging instrument without dispersive elements.
 
 \section{Optical Layout}
+\label{sect:opticallayout}
 In this section, we describe the optical layout for a DTRS.
-We start from the commonly used layout with a single spectrsocopy channel as it is employed in Chandra/LETGS or XMM/RGS.
-We then show how this design can be modified step-by-step and what considerations are important when chosing the specific parameters for a design.
-We illustrastrate the concept with ray-trace simulations.
-Those simulations are performed using a real ray-trace with the marxs code, but the dispersion angle (which is set by the energy of the photons and the grating constant of the grating) is chosen to be larger than feasible for a real instrument.
-This is done purely to separate the direct light (zero order) and the dispersed light (positive and negative orders) better in the images, so that the concept can be explained more clearly.
-Static images are embedded in the article, but we strongly encourage the reader explore the interactive 3D models embedded in the electronic version of the publication (might require Acrobat reader).
+We start from a single central Rowland torus as in Chandra/HETGS and Chandra/LETGS.
+We then show how this design can be modified step-by-step and what considerations are important when choosing the specific parameters for a design.
+We illustrate the concept with ray-trace simulations.
+Those simulations are performed with the marxs code \citep{2017AJ....154..243G}.
+However, the setup is chosen to illustrate the concept of a DTRS, not to represent any particular instrument.
+For example, the position of the diffractive gratings shown leaves valuable mirror area uncovered and is done without considering the size and location of support structures that the gratings can be mounted on.
+The grating constant $d$ is chosen to be smaller than feasible for a real instrument.
+Static images of the 3D ray-traces are embedded in the article, but we encourage the reader explore the interactive 3D models embedded in the electronic version of the publication.
 
-f, d, sketches
+The points $\vec p$ on the surface of a torus can be parameterized by two angles $\varphi$ and $\theta$.
+\begin{equation}
+\vec p(\varphi, \theta) = \vec c + R \; \vec e_R(\varphi) + r \; (\vec e_y \sin \theta + \vec e_R(\varphi) \cos \theta)
+\end{equation}
+where $\vec c$ points to the center of the torus.
+The axis of symmetry is given by a unit vector $\vec e_y$.
+We define a vector $\vec e_R(\varphi) = \vec e_x \cos\varphi + \vec e_z \sin\varphi$.
+$R \vec e_R$ points from the center of the torus to the center of a circle with radius $r$.
+Sweeping that circle around the symmetry axis forms the torus.
+If $r < R$, the torus has an inner hole like a doughnut, and if $r > R$ the inner parts of the torus overlap with each other (``spindle torus'').
+
+Figure~\ref{fig:sketch} shows sketches of the geometry in the plane spanned by the optical axis and the symmetry axis of the torus. Details will be discussed in the following sub-sections.
+
+\begin{figure*}
+    \plotone{sketch}
+    \caption{\emph{left:} Sketch of the on-axis Rowland geometry in the plane of the optical axis and the symmetry axis of te torus.
+Photons arrive from the left.
+The location of the mirror is shown as blue box.
+The optical axis (solid blue line) passed through the center of the Rowland circle (blue square) and is perpendicular to the symmetry axis of the torus (blue long-dashed line).
+The two blue circles show the cut through the 3D torus in this plane; they are connected by blue dotted lines to help visualize the 3D position of the torus.
+    \emph{center:} Tilted Rowland torus.
+The optical axis of the blue channel is offset from the center of the Rowland circle, but intersects the symmetry axis of the torus in the focal point $F$.
+$\alpha$ is the angle between the optical axis and the line from the point where the optical axis intersects the Rowland torus to the center of the Rowland circle.
+$\beta$ is the angle from optical axis to the line connecting that intersection to the ``hinge'', the other point where symmetry axis of the torus and the Rowland circle intersect in this plane.
+    \emph{right:} Layout with two Rowland tori, tilted in opposite direction.
+In the plane spanned by both optical axes (which are parallel) and the symmetry axis of the torus, the two tori overlap in exactly one circle, the Rowland circle.
+        }
+    \label{fig:sketch}
+\end{figure*}
 
@@
 \label{sect:onetorus}
-chip gaps chandra, dither
-
+For a single Rowland torus with no tilt, the optical axis passes through the center of the Rowland circle and $r=R$ as shown in the left panel of figure~\ref{fig:sketch}.
+Figure~\ref{fig:3d:single} shows a 3D view of a ray-trace.
+Since the diffraction gratings are symmetric with respect to the optical axis, positive or negative diffraction orders (up or down in figure~\ref{fig:sketch}) are equally likely, and the diffracted photons will be symmetric with respect to the focal point.
+To avoid the same wavelength to fall in a chip gap on the detector for both the positive and the negative diffraction orders, it is useful to position the CCDs with an offset, so that the chip gaps are not symmetric.
+\begin{figure}
+    \plotone{ifigs/single}
+    \caption{3D image of the ray-trace for a source with a continuum spectrum over a limited bandpass.
+Rays are shown from the entrace aperture on (circle on the right).
+They pass a focussing mirror represented by a green box and get diffracted by gratings (white squares).
+They are detected by CCDs (yellow).
+Rays are colored by the diffraction order: Light purple (middle) are zero order photons.
+A section of the Rowland torus is shown as transparent gray.
+The big square at the left indicates the position of the focal plane.
+Rays are traced past the detectors to the focal plane to show how the CCDs on the Rowland circle capture the rays when the dispersion width is smallest - past the CCDs the rays widen in dispersion direction again, but still become narrower in cross-dispersion direction.
+This figure is interactive in the online version, allowing the reader to pan, zoom, and rotate to see those details and inspect the rays from all angles.
+(Use the mouse to rotate, ``ctrl'' and mouse to pan, mouse wheel to zoom, and double-click to set the center of rotation.
+On other devices, other native controls might be available, e.g.\ two-finger touch to zoom on touchscreens.)
+        }
+    \label{fig:3d:single}
+\end{figure}
 %\begin{figure}
@@
 
-This torus is tilted Ralfs tild paper
-\cite{doi:10.1117/12.856482}
+%alpha, beta, equations. notation different from Ralf.
+
+\begin{figure*}
+    \plotone{ondetector}
+    \caption{Distribution of photons on the detectors for different designs.
+Photons are colored by channel and the position of the zero(s) order(s) is highlighted with red boxes.
+Note that the dispersion and cross-dispersion directions are not to scale.
+        }
+    \label{fig:fish}
+\end{figure*}
+
+\citet{doi:10.1117/12.856482} suggested tilting the torus with respect to the optical axis when using CAT gratings.
+This tilt can be described by two angles $\alpha$ and $\beta$ as shown in the middle panel of figure~\ref{fig:sketch}.
+$\alpha$ is the angle between the optical axis and the line from the point where the optical axis intersects the Rowland torus (labelled ``A'' in the figure) and the focal point of the mirror (``F'').
+We call the distance between those two points $a$, which is chosen to be slightly shorter than the focal length $f$ of the mirror to allow for space to mount the gratings behind the mirror.
+In this configuration, the torus has $R < r$ there are two points where the torus overlaps itself in the plane spanned by the optical axis and the symmetry axis of the torus (the plane shown in figure~\ref{fig:sketch}).
+The first of those intersection points is F.
+We call the second point ``Hinge''.
+$\beta$ is the angle from optical axis to the line connecting A and the Hinge. With that we can derive the radius of the Rowland circle to
+$$r = \frac{a}{2}\cos \alpha.$$
+ If the focal point F is the origin of the coordinate system, then the center of the Rowland circle (blue squares in left blue circles in the panels in figure~\ref{fig:sketch}) is at location
+ \begin{equation}
+    \begin{pmatrix} x \\ y \\ z \end{pmatrix} =
+    \begin{pmatrix} 0 \\ \frac{a}{2} \\  \frac{a}{2}\tan\alpha \end{pmatrix}.
+ \end{equation}
+
+Since A, F, and the Hinge are by definition all on the Rowland circle, their distance to this point is $r$.
+With this symmetry, we can obtain the angle of the axis of the torus with respect to the case where the axis of the torus is perpendicular to the optical axis (left panel of figure~\ref{fig:sketch}).
+This angle is $\beta-\alpha$.
+With that, we can derive $R$, the distance between the center of the Rowland circle and the center of the torus, to
+\begin{equation}
+    R = r \sin\left(\frac{\pi}{2} - \beta\right),
+\end{equation}
+which places the center of the torus at
+\begin{equation}
+    \begin{pmatrix} x \\ y \\ z \end{pmatrix} =
+    \begin{pmatrix} 0 \\ \frac{a}{2}-R\cos{\beta-\alpha} \\ \frac{a}{2}\tan{\alpha}+R\sin{\beta-\alpha} \end{pmatrix}.
+\end{equation}
+
+CAT gratings are blazed and thus deviate more from the Rowland-torus than a grating that is placed perpendicular to the incoming rays.
+Figure~\ref{fig:3d:tilted} shows this layout in 3D, but with a very large blaze angle for clearity; for real CAT gratings typical blaze angles are 1-2~deg.
+Tilting the Rowland torus by about twice the blaze angle reduces this effect.
+Also, CAT gratings disperse predominantly in one direction.
+However, the Rowland geometry optimizes the spectral resolving power by minimizing the spread of the photons in cross-dispersion direction; detectors are located on the Rowland circle, which represents the position of minimal spread in dispersion direction.
+This is not the same location as the imaging focal plane, which is defined by minimizing the total size of the PSF.
+At the focal point, the Rowland circle and the imaging focal plane match.
+Figure~\ref{fig:fish} shows how the dispersed photons spread out in cross-dispersion direction, roughly resembling the outline of a fish.
+With a tilted Rowland torus, there is a second intersection point between the spectral and the imaging focus where the cross-dispersion width becomes small (the point between the body and the tail of the fish, about +300 in dispersion direction in the second panel of figure~\ref{fig:fish}).
+This location can be chosen to be roughly in the center of the dispersed spectrum and thus limit the cross-dispersion profile to reduce the area of the detector that contributes to the extracted signal and thus the background.
+We note that the width in cross-dispersion direction is much larger for larger diffraction angles and, in real instruments, will be smaller than shown in the figure.
+
+\begin{figure}
+    \plotone{ifigs/tilted}
+    \caption{Ray-trace like figure~\ref{fig:3d:single}, but for a tilted Rowland torus setup.
+See figure~\ref{fig:3d:single} for an explanation of the elements shown in the image and the interactivity in the online version.
+        }
+    \label{fig:3d:tilted}
+\end{figure}
+
+A tilted Rowland can also be combined with sub-aperturing to increase the spectral resolving power as discussed in figure~\ref{fig:subaperture}.
+In this case, not all of the fish shape is filled with photons.
+This can be seen by comparing the dark purple points in the second, third, and forth panel in figure~\ref{fig:fish}.
+The Rowland torus for this channel has the same tilt and dimensions in all three designs, but in the second panel, the mirror fills the full circle, in the third panel it has just two wedges, and in the forth panel just one.
+In either case, the spectral extraction region used in data analysis should take this shape into account to minimize detector background.
 
 \subsection{Double tilted Rowland Torus}
-For R > r signal dips below plane, fish form, two points, symmetry
+\begin{figure}
+    \plotone{ifigs/tilted_double}
+    \caption{Ray-trace like figure~\ref{fig:3d:tilted}, but with channels split on two separately tilted Rowland tori.
+See figure~\ref{fig:3d:single} for an explanation of the elements shown in the image and the interactivity in the online version.
+        }
+    \label{fig:3d:tilted_double}
+\end{figure}
 
-two opticla axes,
-offset chip gaps
-spectroscopic vs imganing focus unless I;ve said that above.
+Sub-aperturing reduces the geometric area of the mirror and thus the effective area of the instrument.
+This can be compensated by adding a second channel, with mirrors mounted next to the mirrors of the first channel.
+Again, gratings and detectors for this second channel need to be positioned on a Rowland torus, however the two Rowland tori for the two channels need not be identical.
+When $R> r$ the Rowland circle intersects the focal plane twice.
+We choose those two points as the focal points for the two channels of the instrument; this layout is symmetric with respect to the center of the Rowland circle (figure~\ref{fig:sketch}, right).
+By selecting $\alpha$ close to the average dispersion angle of dispersed photons, we can reach a configuration where the same detectors that detect the zeroth order of one channel, will now also detect the dispersed photons of the other channel and vice versa.
+The second channel also adds redundancy in case of detector failure, and can compensate for chip gaps if the position of the detector elements on one channels is moved along the Rowland circle a little bit with respect to the other channel.
+Figure~\ref{fig:3d:tilted_double} shows a ray-trace for this configuration.
 
 \subsection{Channel offsets}
+The next two subsections explain how the channels can be offset in both dispersion and cross-dispersion direction to reduce alignment requirements and mitigate chip gaps.
+Figure~\ref{fig:3d:tilted_double_offset} shows a 3D ray-trace image in this configuration; the resulting signal on the detector is shown in figure~\ref{fig:fish} (bottom panel).
+\begin{figure}
+    \plotone{ifigs/tilted_double_offset}
+    \caption{Ray-trace like figure~\ref{fig:3d:tilted_double}, but with offsets in dispersion and cross-dispersion direction.
+See figure~\ref{fig:3d:single} for an explanation of the elements shown in the image and the interactivity in the online version.
+        }
+    \label{fig:3d:tilted_double_offset}
+\end{figure}
+\subsubsection{Channel offsets in the cross-dispersion direction}
+In the data reduction, we need to assign a unique wavelength to each dispersed photon.
+To do that, it is necessary to know which optical channel a photon passed through and thus the setup needs to avoid overlapping photons from both channels on the same location of the detector.
+This can be accomplished by shifting the Rowland tori (and mirrors and gratings positioned on them) in the cross-dispersion direction (into or out of the plane of the plot in figure~\ref{fig:sketch}).
+The shift should be large enough to clearly separate the two channels, but keep both dispersed spectral traces on the detector with margin (figure~\ref{fig:fish}).
+Since this shift is perpendicular to the Rowland circle, it does not impact the spectral resolving power.
 
-\subsubsection{Channel offsets in the cross-dispersion direction}
+While it is possible to have a mirror that spans the entire circle for each of the two channels, the DTRS layout is particularly well-suited for sub-aperturing.
+In this case, each channel will have two opposing wedges filled with mirror elements and gratings leading to a total of four mirror and grating wedges.
+These can then be arranged closely together (see figure~\ref{fig:3d:tilted_double}) to maximize the area used given the size constraints of the spacecraft.
+In this case, it is an option to split each of the two channels into two sub-channels with their own separate offset, such that there are two parallel traces running in each direction (Figure~\ref{fig:3d:tilted_double_offset}) for a total of four traces (figure~\ref{fig:fish}, right panel).
+Without this split, the opposing mirror and grating wedges have to be carefully aligned to the same optical axis, any misalignment between them would broaden the PSF and thus reduce the spectral resolving power.
+With this split, each wedge defines its own optical axis and the exact location and thus the wavelength scale for each zeroth order can be determined from observations.
 
 \subsubsection{Channel offsets in the dispersion direction}
+When each channel is split into two sub-channels as described above and each of the sub-channels has its own Rowland torus (not shown in the figures because four overlapping tori are hard to visually separate), the two sub-channels can also be offset in the dispersion direction.
+This causes the Rowland torus that defines the grating positions to deviate from the torus that sets the position of the detectors, but if the offset is small, the impact on the spectral resolving power is negligible.
+A benefit of such an offset in dispersion direction is that the dispersed spectra of the two sub-channels do not hit chip gaps at the same wavelength, spreading out the effect of chip gaps on the extracted spectrum without the need to dither the spacecraft.
 
 \section{Discussion}
+\label{sect:discussion}
 
 \subsection{Subaperturing space considerations, and resolving power}
-lower mirror requirements, thus chepaer cost for same resolving power, but can use wdge only, stack channels to get square aperture and get back effective area
+Manufacturing mirrors with a smaller PSF is more difficult and costly than manufacturing mirrors with a larger PSF, but otherwise identical properties.
+For a spectrograph, the width of the PSF in dispersion direction is far more important than the width in cross-dispersion direction.
+If the width in dispersion direction can be reduced by sub-aperturing, that provides a higher spectral resolving power at lower cost and technical complexity compared to reducing the PSF of the mirror itself.
+On the other hand, it reduces the collecting area of the mirror, since only a fraction of the full circle can be used.
+The DTRS design allows us to compensate that by having multiple channels, ultimately delivering better resolving power at the same mirror requirements than a single-channel design.
 
-\subsection{Number of CCDs}
-2 camerase, different directions. Need to get zeroth orders anyway (high-energy, wavelength scale), but more CCDs than just one directions (Lynx), make good use of rea-estate
+The independent mirror wedges in a DTRS design can be arranged on the front-end of the spacecraft such that they take up a similar geometric area as a mirror filling the full circle in a single channel design.
+
+\subsection{Zeroth order is seen}
+Observing the direct image of all channels on the detector has several advantages.
+First, most of the high-energy photons are not diffracted by typical CAT gratings.
+So, the direct image can be used for analysis of high-energy spectrum using the intrinsic energy resolution of the detector, e.g.\ a CCD or even microcalorimeter.
+For example, the Fe line at 6.7~keV is a widely used diagnostic for many astrophysical objects.
+Second, accurate knowledge of the position of the direct image is important for the wavelength calibration of the dispersed signal.
+The wavelength of the light is determined by measuring the distance $x$ from the detected photon to the zeroth order and the known distance $a$ between the gratings and the focal plane.
+Conceptually, $\tan \theta = x/a$.
+With the known grating constant $d$, equation~\ref{eqn:diffraction} then yields $\lambda$.
+For high spectral resolving power and large values of $\theta$ it might be important to take the 3d shape of the detector into account, but either way, accurate knowledge of the zero order position is crucial.
+
+\subsection{Number of detector elements}
+An instrument that uses only a single tilted Rowland torus needs to cover a long strip of dispersed light with a detector and a small area around the zeroth order.
+On the other hand, a DTRS needs two long detector strips to cover the dispersed signal on both sides; those cameras will cover the zeroth order of all channels at the same time.
 
 \subsection{Robustness}
-relaxes alignment because multiple channels (thus lower cost and technical risk), resilience because independent channels on different CCDS (see XMM loss of RGS chip), check for systemics in calibration because independent channels, chip gaps
+The DTRS concept increases the robustness and reduces the technical risk of a mission in several ways.
+Because we can assign each sub-apertured mirror wedge its own channel (in the 4 channel design with offsets), there is no need to align the mirror for each channel to each other to better than the cm scale.
+Also, the design is robust against loosing individual detectors in a camera.
+Since half of the photons are dispersed from left-to-right and the others right-to-left, loss of a detector element in a camera does not result in a complete loss of spectral coverage for any wavelength.
+The multichannel design can also help to uncover systematics in the calibration.
 
-wavelength self-calibration (does not need DTRS, but works there, too). is order sorting and several orders ad one energy
-
-\subsection{Other advantages}
-
-Zero order seen, for direct observation (hard X-rays) and calibration
+%wavelength self-calibration (does not need DTRS, but works there, too). is order sorting and several orders ad one energy
 
 \subsection{Drawbacks}
-increased background due ot larger CCD region, less source-free region for background.instrum analysis, coplexity of analysis (numberof spectra!), sprectra coupled (cross-dispersion, read-out streaks for traditional CCDs, wings and energy resolution), does not work for crowded fields, fields with extended emission (galactic center)
+Because the DTRS concept spreads the dispersed photons over a larger detector area than a single-channel design, the background contamination in the spectra is higher, simply because more pixels contribute to the extracted science spectrum.
+This does not matter for bright isolated point-source, but makes observations of faint sources, where the dispersed spectrum has count rates comparable to the background, difficult.
+Similarly, observations where there is extended emission in the field-of-view, for example in the galactic center, are difficult.
+The DTRS concept is also not well-suited for crowded fields, where the dispersed signal of one source overlaps with the dispersed signal of another source.
+
+Depending on the detector technology chosen, the DTRS concept might also lead to undesirable coupling between the spectra of different channels.
+For example, traditional CCDs have a read-out streak, where photons that arrive at the detector during the charge transfer are not associated with the correct position on the detector in the direction of the read-out.
+This means that photons that should be recorded at the zero-order position or at the position of a strong emission line in one channel are assigned to the wrong channel, where they might cause what looks light a weak emission feature.
+This cross-talk can be avoided by using different detector technologies, or be taken into account in the data analysis.
+Similarly, the pure number of channels with different chip gaps and detector positions increases the computational complexity of the data analysis compared to a single-channel design.
+Development of dedicated software tools is needed to help with the analysis of these multichannel data.
 
@@
 
-The Far Ultraviolett Spectroscopic Explorer (FUSE) \cite{2000ApJ...538L...1M} faced a design challenge similar to what we discuss here for X-ray instruments.
+The Far Ultraviolett Spectroscopic Explorer (FUSE) \citep{2000ApJ...538L...1M} faced a design challenge similar to what we discuss here for X-ray instruments.
 FUSE also has four spectroscopic channels with gratings arranged in a Rowland geometry.
 Unlike in X-rays, UV gratings can operate in reflection close to normal incidence.
-The reflection gratings are thus positioned perpendicular to the path of the rays, which means that are almost tangential to the surface of the Rowland torus.
-Thus, a small number of gratings (just one per channel in the extreme case) is sufficient and, even if the grating itself is flat, it never deviates much from the surface of the Rowland torus.
+The reflection gratings are thus positioned almost perpendicular to the path of the rays tangential to the surface of the Rowland torus.
+Thus, a few gratings (just one per channel in the extreme case) is sufficient and, even if the grating itself is flat, it never deviates much from the surface of the Rowland torus.
 Since UV light has a longer wavelength, it can be dispersed to larger angles increasing the spectral resolving power.
@@
 If the pointing position drifts and the optical channels are not fully aligned, this can be seen in the science data.
-In contrast, FUSE has performed observations where the target was not in the field-of-view of all channels, but the amount of the exposure time lost is unclear \cite{2000ApJ...538L...1M}.
+In contrast, FUSE has performed observations where the target was not in the field-of-view of all channels, but the amount of the exposure time lost is unclear \citep{2000ApJ...538L...1M}.
+
+\section{Applications}
+\label{sect:applications}
+In the previous sections, we discussed the concept of a DTRS in general.
+The Arcus X-ray spectrograph \citep{2023SPIE12678E..0ES} adopts this concept with $\alpha=3.6$~deg and $\beta=2\alpha$.
+Arcus has gone through multiple design iterations for different mission calls and the design will continue to be refined.
+As an example, we describe the Arcus setup here as submitted to the NASA probe call in 2023.
+Arcus uses silicon-pore optics (SPO) with a 12~m focal length that are developed for the Athena mission \citep{2023SPIE12679E..05G}; $a$ is slightly smaller than the focal length with $a=11876$~mm.
+SPOs are manufactured as small stacks of about 30 mirror plates with sizes of order 10-20~cm in each direction (the exact size depends on the radius).
+40 SPOs are aligned into one petal and each petal defines one of four optical channels in a DTRS design with offsets.
+The two sub-channels are offset by $\pm2.5$~mm in dispersion direction with respect to the design with just two Rowland tori.
+In cross-dispersion the four channels are located at $-7.5$, $-2.5$, $+2.5$, and $+7.5$~mm from the center of the CCDs.
+Critical-angle transmission (CAT) gratings are manufactured through deep reactive-ion etching of silicon-on-insulator wafers \citep{2021SPIE11822E..15H,2023SPIE12679E..0LH}.
+These gratings are mounted with a blaze angle, i.e.\ they are intentionally tilted by 1.8~deg with respect to the incoming photons, which concentrates the diffracted photons around the ``blaze peak'' at a diffraction angle of $2 * 1.8 = 3.6$~deg, so that only 8~CCDs are needed in each camera covering about 30~cm in the dispersion direction with a gap close to 50~cm between the two cameras.
+Arcus will achieve a spectral resolving power around 3500 for wavelength between 1.5 and 6.0~nm, averaged over all diffraction orders, and an effective area up to 500~cm$^2$ between 1.5 and 2.5~nm, dropping to 100~cm$^2$ at 6.0~nm.
+Detailed ray-tracing simulations for Arcus that optimize the parameters of the design and predict the performance depending on the misalignment for the different components are shown in \citet{2017SPIE10397E..0PG,2018SPIE10699E..6FG,2023SPIE12678E..1DG}.
+
+\begin{figure*}
+    \plotone{EQFullDet}
+    \caption{Simulation of an Arcus observation of an emission lines spectrum (see section~\ref{sect:applications} for details).
+Position of the CCDs are outlined with a white frame and the zeros orders are indicated.
+Dashed arrows mark the dispersion direction of the four orders from the zeroths order.
+Only photons that are detected on a CCD are included.
+Note that the dispersion and cross-dispersion directions are scaled differently.
+        }
+    \label{fig:Arcusfull}
+\end{figure*}
+
+Figure~\ref{fig:Arcusfull} shows a simulation of an Arcus observation of an emission line spectrum.
+The spectral model is taken from the active star EQ~Peg~A \citep{2008A&A...491..859L} and the simulation is run for an exposure time of 100~ks.
+Note that the plot shows an extremely distorted view of the CCD plane.
+The $x$-axis extends over 1.2 m, while the $y$-axis covers only 2.5 cm.
+
+This particular simulation does not include any background.
+The photons that are seen in between the spectra are cross-dispersed photons.
+The CAT gratings disperse photons in one direction.
+However, the main CAT grating bars are held in place by a support structure (L1) that also acts as a grating and disperses some photons perpendicular to the main dispersion direction.
+This is visible in the plot for the zeroth orders and the strongest emission lines.
+
+\begin{figure*}
+    \plotone{EQAllDet}
+    \caption{Images of the 16 individual Arcus CCDs are shown for the same simulation as in figure~\ref{fig:Arcusfull}. Note that the color scale is chosen differently for each chip.
+        }
+    \label{fig:Arcusall}
+\end{figure*}
+
+Figure~\ref{fig:Arcusall} shows the same simulation, but with each CCD to scale.
+The color scale is different for each CCD though, to bring out details on CCDs that do not contain the bright zeroths orders or extremely bright emission lines.
+
+%Each spectral trace is slightly curved.
+This is the result of astigmatism coupled with sub-aperturing.
+If each channel had a full 360 degree mirror, we would see an hour-glass shape that is narrowest at the position of the zeroth order and then widens as the spectral focus diverges from the imaging focus.
+In Arcus' design with a tilted torus, there is a second point where the spectral focus coincides with the imaging focus and thus the spectral trace would become narrower in cross-dispersion direction again.
+However, the SPOs for each channel cover only a very small fraction of the full circle, and thus the spectral trace fills only one "edge" of the hourglass shape, which results in the curved image.
+The dispersion direction is exactly parallel to the $x$ axis at all locations, but the $y$ extend of the extraction region can be chosen to follow the curve to reduce the number of background events in the spatial extraction region.
 
 \section{Summary}
+\label{sect:summary}
+\emph{Chandra}/HETGS and LETGS employ an on-axis Rowland torus geometry.
+We show how this geometry can be modified in several steps to increase the spectral resolving power without compromising the effective area for a specific class of future X-ray observatories.
+Our design is optimized for grating spectroscopy of X-ray point sources.
+We use sub-aperturing to increase the spectral resolving power.
+We tilt the Rowland torus such that the optical axis of the mirror is offset from the center of the Rowland circle.
+That reduces the deviation of flat gratings from the surface of the torus and also allows us to position a second channel next to the first one, doubling the effective area compared to a single, sub-apertured channel.
+Both channels have different Rowland tori (dobule tilted Rowland spectrograph, DTRS), but they overlap in the Rowland circle, and thus both channels can be imaged onto the same set of detectors.
+In this geometry, all zeroths orders are visible for wavelength calibration and to detect hard X-rays in direct imaging.
+Finally, we discuss how small offsets between the channels can mitigate chip gaps and reduce the alignment requirements between optical elements during the assembly of the instrument.
+We illustrate these concepts with sketches and ray-traces designed to illustrate the concept.
 
-A specialized layout for a mission with point source spectroscopy goal, not a general observatory design
-
-Arcus (need ot bring is up finally - or have a section on "Application" obove?)
+We briefly discuss Arcus, one NASA probe class mission concept that implements a DTRS and show realistic ray-traces for a 100~ks observation of an emission line dominated spectrum with Arcus.
 
@@
 
-\software{marxs\cite{}, AstroPy \citep{2013A&A...558A..33A,2018AJ....156..123A}, NumPy \citep{van2011numpy,harris2020array}, Matplotlib \citep{Hunter:2007},IPython\cite{IPython}}
+\software{marxs \citep{2017AJ....154..243G}, AstroPy \citep{2013A&A...558A..33A,2018AJ....156..123A}, NumPy \citep{harris2020array}, Matplotlib \citep{Hunter:2007}, IPython\cite{IPython}}
 

### 2024-03-18 papers/Doubletiltconcept/doubletiltconcept.tex
add authors to paper draft
Address Ralf's and Casey's comments

@@
 \affiliation{MIT Kavli Institute for Astrophysics and Space Research, 77 Massachusetts Avenue, Cambridge, MA 02139, USA}
-
-\author{Add your name behind this, will sort later}
-\affiliation{}
+\author[0000-0002-9184-4561]{Casey DeRoo}
+\affiliation{Dept. of Physics \& Astronomy, University of Iowa, Iowa City, IA 52242, USA}
+\author[0000-0001-9980-5295]{Ralf K. Heilmann}
+\affiliation{MIT Kavli Institute for Astrophysics and Space Research, 77 Massachusetts Avenue, Cambridge, MA 02139, USA}
+\affil{Space Nanotechnology Laboratory, MIT Kavli Institute for Astrophysics and Space Research, Cambridge, MA 02139, USA}
+\author[0000-0002-6747-9648]{Edward Hertz}
+\affiliation{Center for Astrophysics, Harvard-Smithsonian Astrophysical Observatory, Cambridge, MA 02138, USA}
 
@@
 High-resolution spectroscopy in soft X-rays ($<2$~keV) requires diffractive elements to resolve any astrophysically relevant diagnostics, such as closely spaced lines, weak absorption lines, or line profiles.
-For transmission gratings, the Rowland torus geometry describes how gratings and detectors need to be positioned to optimize the spectral resolving power.
+The Rowland torus geometry describes how gratings and detectors need to be positioned to optimize the spectral resolving power.
 We describe how an on-axis Rowland geometry can be tilted to accommodate blazed gratings.
@@
 X-ray spectroscopy can deliver many insights into the astrophysics of the hot and energetic universe that are not otherwise observable.
-In particular, high-resolution spectroscopy (with resolving power $> 500$ or better) can resolve closely-spaced emission lines which offer diagnostics such as density and temperature, reveal the near-edge fine structure in absorption lines, or allow us to detect the presence of faint and narrow absorption lines caused by the warm-hot intergalactic medium in front of bright background quasars to give just a few examples.
+In particular, high-resolution spectroscopy (with resolving power $R=E/\Delta E> 500$ or better) can resolve closely-spaced emission lines which offer diagnostics such as density and temperature, reveal the near-edge fine structure in absorption lines, or allow us to detect the presence of faint and narrow absorption lines caused by the warm-hot intergalactic medium in front of bright background quasars to give just a few examples.
 In the X-ray band, such high resolving powers can be achieved with either microcalorimaters or dispersive elements such as crystals or diffraction gratings.
-Microcalorimeters offer a fixed absolute energy resolution $\Delta E$, which gives them good resolving power $R=E/\Delta E$ at large energies $E$ and consequently a microcalorimeter has been launched on Astro-E/Hitomi/SXS \citep{2014SPIE.9144E..2AM} and XRISM/Resolve \citep{2018JATIS...4a1214K} and this technology is also considered for other large missions in development such as Athena \citep{2014SPIE.9144E..2LR} and Lynx \citep{2019JATIS...5b1017B}.
-\citet{2016Natur.535..117H} used the microcalorimeter data from Astro-E/Hitomi to measure the line-of-sight velocity dispersion of \ion{Fe}{26} in the Perseus cluster and found that the turbulent pressure is only a few percent of the thermodynamic pressure in that region.
-However, diffraction gratings have a distinct advantage for softer X-rays below about 2~keV.
+Microcalorimeters offer a fixed absolute energy resolution $\Delta E$, which gives them good resolving power $R$ at large energies $E$ and consequently a microcalorimeter has been launched on Astro-E/Hitomi/SXS \citep{2014SPIE.9144E..2AM} and XRISM/Resolve \citep{2018JATIS...4a1214K} and this technology is also considered for other large missions in development such as Athena \citep{2014SPIE.9144E..2LR} and Lynx \citep{2019JATIS...5b1017B}.
+Below about 2~keV diffraction gratings have a distinct advantage.
 
@@
 Unlike microcalorimeters, diffraction gratings require an additional dispersive element (a transmission or a reflection grating) in the optical path of the instrument and dedicated detectors to capture the dispersed signal.
-One common layout for a diffraction spectrometer is the Rowlandtorus \citep{Beuermann:78}.
+For a given diffraction order, gratings typically deliver a fixed wavelength resolution $\Delta \lambda$, so they perform better than microcalorimeters at longer wavelengths/lower energies.
+One common layout for a diffraction spectrometer is the Rowland torus \citep{Beuermann:78}.
 In the plane defined by the optical axis and the dispersion direction, the gratings are located at one side of a circle, the detectors on the other.
@@
 In this article, we present an optical layout for an instrument that stacks two or four different channels with distinct optical axes in such a way that each channel follows the Rowland torus prescription, yet at the same time all channels are imaged onto a common set of detectors.
-The general concept could be applied for a large range of focal lengths, grating constants, and other hardware properties.
+The general concept could be applied to a large range of focal lengths, grating constants, and other hardware properties.
 Thus, we discuss advantages and challenges of this layout in general.
-However, we note that Arcus \citep{2023SPIE12678E..0ES}, an X-ray spectroscopy mission proposed for NASA midex (medium explorer) and probe calls, applies this layout.
+However, we note that Arcus \citep{2023SPIE12678E..0ES}, an X-ray spectroscopy mission proposed for NASA MIDEX (medium explorer) and probe calls, applies this layout.
 Specific details of the  Arcus performance, such as predicted effective area and resolving power are shown in \citet{2018SPIE10699E..6FG,2023SPIE12678E..1DG}, but the general concept of a double tilted Rowland spectrograph (DTRS) with its advantages and drawbacks has not been discussed -- a void we attempt to fill with this article.
@@
 \label{sect:elements}
-An X-ray spectrograph using the Rowland layout consists of a focussing mirror, diffractive gratings, and a detector.
+An X-ray spectrograph using the Rowland layout consists of a focusing mirror, diffractive gratings, and a detector.
 
@@
 The concept of a DTRS can be implemented with different mirror technologies.
-The size of the collecting area is to first order proportional to the number of photons that the spectrograph can detect, and the spectral resolving power $\Delta \lambda / \lambda$ increases when the mirror point-spread-function (PSF) decreases.
-Depending on the type of mirror chosen, the PSF can be dominated by the alignment error between mirror shells, in particular for mirrors that consist of hundreds of elements.
+In X-rays, the mirror assembly will typically consist of nested (full or partial) shells that reflect X-rays in grazing incidence.
+The higher the energy of the X-ray, the smaller the angle between the mirror surface and the incoming ray has to be to allow for reflection.
+The size of the collecting area is to first order proportional to the number of photons that the spectrograph can detect, and the spectral resolving power $\lambda / \Delta \lambda$ increases when the mirror point-spread-function (PSF) decreases.
+Depending on the type of mirror chosen, the PSF can be dominated by the alignment error between mirror shells, in particular for mirror assemblies that consist of hundreds of elements.
 In this case, X-rays are offset from the focus point both in the plane of reflection and perpendicular to it.
 On the other hand, mirrors can also have a figure error, which is the deviation of the mirror surface from the ideal shape.
-If the PSF is dominated by the figure error or scattering by particulates on the mirror surface, then scatter in the plane of reflection is typically larger than the out-of-plane scatter.
-In this case, sub-aperturing of the aperture can improve the spectral resolving power.
+If the PSF is dominated by the figure error or scattering by microroughness or particulates, then scattering angles in the plane of reflection are typically larger than the out-of-plane scatter.
+In this case, sampling a limited azimuthal span of the aperture -- subaperturing -- can improve the spectral resolving power \citep{1987ApOpt..26.2915C}.
 This is illustrated in Fig.~\ref{fig:subaperture}.
@@
 All panels show a view from the universe looking down onto the telescope.
-\emph{left}: Each dot is a photon shown as it passes through a diffraction gratings.
+\emph{left}: Each dot is a photon shown as it passes through a diffraction grating.
 Photons are colored according to their polar angle in this panel.
@@
 \subsection{Dispersive elements}
-In a Rowland torus geometry, diffraction gratings are placed on the surface of the Rowland torus.
+In a Rowland torus geometry, diffraction gratings are placed on the surface of the Rowland torus, ideally tangentially for transmission gratings.
 They diffract light according to the grating equation for photons arriving perpendicular to the grating surface:
@@
 where $\lambda$ is the wavelength of the light, $n$ is the order of the diffraction, $d$ is the grating constant, and $\theta$ is the angle between the incoming light and the normal to the grating.
-(The equation can be rewritten for rays hitting the grating at arbitrary angles, but for the purpose of this conceptual discussion that is not needed.)
+(The equation can be rewritten for rays hitting the grating at arbitrary angles, but for the purpose of this conceptual discussion this is not needed.)
 Typically, these transmission gratings are flat and only a few cm$^2$ in size; the spacing between gratings bars $d$ is the same for all gratings and does not vary within a grating.
@@
 \citet{2020SPIE11444E..88G} explored bending the gratings in one dimension or varying the distance between grating bars over a grating to compensate for the deviations from the ideal shape in simulations.
-However, these concepts have not been implemented in a hardware yet.
+However, these concepts have not yet been implemented in X-ray transmission gratings.
 
@@
 Thus, the ideal detector is much longer in one direction than in the other.
-Individual detector elements (e.g.\ CCDs) can be tiled, but to avoid chip gaps where photons are lost and gaps appear in the extracted spectum, it is beneficial to keep the number of elements low.
+Individual detector elements (e.g.\ CCDs) can be tiled, but to avoid chip gaps where photons are lost and gaps appear in the extracted spectrum, it is beneficial to keep the number of elements low.
 In our DTRS design several spectral traces are positioned in parallel (section~\ref{sect:opticallayout}), so the detector has to be position sensitive in 2D to separate individual channels.
@@
 Detectors that might be at risk of pile-up in direct observations can still be used for the dispersed signal.
-On the other hand, the larger regions needed to extract the signal imply that background events (internal background and diffuse astrophysical emission) are more important than in direct imaging observations.
-Similarly, fast timing resolution is typically not needed, because the lower signal means that the detected photons need to be binned up into longer time bins for analysis anyway.
-Intrinsic energy resolution is required to both suppress the background by filtering out events that are incompatible with wavelengths of the dispersed signal at any particular location and to separate grating orders, where photons of different orders are diffracted to the same position (e.g.\ the sixth order for the O~{\sc vii} forbidden line of the He-like triplet at 2.21~nm and the seventh order of 1.94~mn photons, close to the O~{\sc viii} Lyman$\alpha$ line end up at almost the same $\theta$ according to equation~\ref{eqn:diffraction}).
+On the other hand, the larger regions needed to extract the signal imply that background events (internal background and diffuse astrophysical emission) make up a larger fraction of the extracted signal than in direct imaging observations.
+Similarly, fast timing resolution is typically not essential, because the lower signal means that the detected photons need to be binned up into longer time bins for analysis anyway.
+Intrinsic energy resolution is required to both suppress the background by filtering out events that are incompatible with wavelengths of the dispersed signal at any particular location, and to separate grating orders where photons of different orders are diffracted to the same position (e.g.\ the sixth order for the O~{\sc vii} forbidden line of the He-like triplet at 2.21~nm and the seventh order of 1.94~nm photons, close to the O~{\sc viii} Lyman $\alpha$ line end up at almost the same $\theta$ according to equation~\ref{eqn:diffraction}).
 
-In our DTRS design, the direct images of all channels are positioned on detectors to provide an accurate wavelength calibration, but also because the high-energy part of the spectrum, which is not efficiently dispersed by gratings, can provide valuable context for the scientific interpretation of the high-resolution grating spectrum.
+In our DTRS design, the images ($n=0$) of all channels are positioned on detectors to provide an accurate wavelength calibration, but also because the high-energy part of the spectrum, which is not efficiently dispersed by gratings, can provide valuable context for the scientific interpretation of the high-resolution grating spectrum.
 Thus, the detector type chosen should also be able to handle direct imaging, albeit at a lower count rate than in a direct imaging instrument without dispersive elements.
@@
 We start from a single central Rowland torus as in Chandra/HETGS and Chandra/LETGS.
-We then show how this design can be modified step-by-step and what considerations are important when choosing the specific parameters for a design.
+We then show how this design can be modified step-by-step to accommodate blazed gratings and what considerations are important when choosing the specific parameters for a design.
 We illustrate the concept with ray-trace simulations.
@@
 
-Figure~\ref{fig:sketch} shows sketches of the geometry in the plane spanned by the optical axis and the symmetry axis of the torus. Details will be discussed in the following sub-sections.
-
+Figure~\ref{fig:sketch_label} shows sketches of the geometry in the plane spanned by the optical axis and the symmetry axis of the torus. Details will be discussed in the following subsections.
+\begin{figure*}
+    \includegraphics[width=0.5\textwidth]{sketch_label}
+    \includegraphics[width=0.5\textwidth]{raytrace_labelled}
+    \caption{\emph{left:} Sketch of a tilted Rowland torus geometry shown in the plane of the optical axis and the symmetry axis of the torus.
+Photons arrive from the left.
+The location of the mirror is shown as blue box.
+The optical axis (solid blue line) is offset from the center of the Rowland circle (blue square) and is perpendicular to the symmetry axis of the torus (blue long-dashed line).
+The two blue circles show the cut through the 3D torus in this plane; they are connected by blue dotted lines to help visualize the 3D position of the torus.
+$\alpha$ is the angle between the optical axis and the line from the point where the optical axis intersects the Rowland torus (A) to the center of the Rowland circle.
+$\beta$ is the angle from optical axis to the line connecting A to the ``hinge'' -- the other point in this plane where symmetry axis of the torus and the Rowland circle intersect.
+    \emph{right:} Visualization of a ray-trace of the same geometry.
+Colored lines show the path of individual rays starting on the left.
+Rays detected in the zeroth order are shown in purple, green and yellow are higher orders.
+Optical axis, the axis of the Rowland torus etc.\ are indicated in blue and the location of detectors in orange, matching the sketch in the left panel.
+        }
+    \label{fig:sketch_label}
+\end{figure*}
 \begin{figure*}
     \plotone{sketch}
-    \caption{\emph{left:} Sketch of the on-axis Rowland geometry in the plane of the optical axis and the symmetry axis of te torus.
-Photons arrive from the left.
-The location of the mirror is shown as blue box.
-The optical axis (solid blue line) passed through the center of the Rowland circle (blue square) and is perpendicular to the symmetry axis of the torus (blue long-dashed line).
-The two blue circles show the cut through the 3D torus in this plane; they are connected by blue dotted lines to help visualize the 3D position of the torus.
-    \emph{center:} Tilted Rowland torus.
-The optical axis of the blue channel is offset from the center of the Rowland circle, but intersects the symmetry axis of the torus in the focal point $F$.
-$\alpha$ is the angle between the optical axis and the line from the point where the optical axis intersects the Rowland torus to the center of the Rowland circle.
-$\beta$ is the angle from optical axis to the line connecting that intersection to the ``hinge'', the other point where symmetry axis of the torus and the Rowland circle intersect in this plane.
+    \caption{\emph{left:} Sketch of the on-axis Rowland geometry in the plane of the optical axis and the symmetry axis of the torus. Photons arrive from the left.
+    \emph{center:} Tilted Rowland torus. The optical axis of the blue channel is offset from the center of the Rowland circle, but intersects the symmetry axis of the torus in the focal point $F$.
     \emph{right:} Layout with two Rowland tori, tilted in opposite direction.
 In the plane spanned by both optical axes (which are parallel) and the symmetry axis of the torus, the two tori overlap in exactly one circle, the Rowland circle.
+See figure~\ref{fig:sketch_label} for a more detailed explanation.
         }
@@
 \label{sect:onetorus}
-For a single Rowland torus with no tilt, the optical axis passes through the center of the Rowland circle and $r=R$ as shown in the left panel of figure~\ref{fig:sketch}.
+For a single Rowland torus with no tilt, the optical axis passes through the center of the Rowland circle and $R=r$ as shown in the left panel of figure~\ref{fig:sketch}.
 Figure~\ref{fig:3d:single} shows a 3D view of a ray-trace.
@@
 \begin{figure}
-    \plotone{ifigs/single}
-    \caption{3D image of the ray-trace for a source with a continuum spectrum over a limited bandpass.
-Rays are shown from the entrace aperture on (circle on the right).
+  \begin{interactive}{js}{ifigs/on-axis.zip}
+  \includegraphics[width=0.22\textwidth]{ifigs/on-axis_overview}
+  \includegraphics[width=0.22\textwidth]{ifigs/on-axis_aperture}
+  \includegraphics[width=0.22\textwidth]{ifigs/on-axis_gratings}
+  \includegraphics[width=0.22\textwidth]{ifigs/on-axis_detector}
+  \end{interactive}
+\caption{3D image of the ray-trace for a source with a continuum spectrum over a limited bandpass, seen from the side, from the top (where the rays enter the aperture), zoom on the grating positions, and zoom on the detector.
+Rays are shown from the entrance aperture on.
 They pass a focussing mirror represented by a green box and get diffracted by gratings (white squares).
@@
 A section of the Rowland torus is shown as transparent gray.
-The big square at the left indicates the position of the focal plane.
-Rays are traced past the detectors to the focal plane to show how the CCDs on the Rowland circle capture the rays when the dispersion width is smallest - past the CCDs the rays widen in dispersion direction again, but still become narrower in cross-dispersion direction.
+The big square at the right indicates the position of the focal plane.
+Rays are traced past the detectors to the focal plane to show how the CCDs on the Rowland circle capture the rays where the dispersion width is smallest - past the CCDs the rays widen in dispersion direction again, but still become narrower in cross-dispersion direction.
 This figure is interactive in the online version, allowing the reader to pan, zoom, and rotate to see those details and inspect the rays from all angles.
@@
 On other devices, other native controls might be available, e.g.\ two-finger touch to zoom on touchscreens.)
-        }
-    \label{fig:3d:single}
+\label{fig:3d:single}}
+
 \end{figure}
-%\begin{figure}
-%  \begin{interactive}{js}{file_wth_arll_components.tar/zip}
-%  figure call (e.g.\plotone, \includegraphics, etc.)
-%  \end{interactive}
-%  \caption{Description of the example figure and the interactive portion which includes how to use the interactive functionality.}
-%\end{figure}
 
@@
 
-\begin{figure*}
-    \plotone{ondetector}
+\begin{figure}
+    \centering
+    \includegraphics[width=0.6\textwidth]{ondetector}
     \caption{Distribution of photons on the detectors for different designs.
@@
 Note that the dispersion and cross-dispersion directions are not to scale.
+In the lower two panels the zeroths orders of one channel overlap with the dispersed phontons from another channel.
+This could be avoided with larger offsets between channels in cross-dispersion direction.
+Also, in real instruments with focal lengths significantly larger than in the illustrative model we use throughout this article, the bending and the width in cross-dispersion direction are typically smaller, so that this is not a problem in practice.`'
         }
     \label{fig:fish}
-\end{figure*}
+\end{figure}
 
-\citet{doi:10.1117/12.856482} suggested tilting the torus with respect to the optical axis when using CAT gratings.
-This tilt can be described by two angles $\alpha$ and $\beta$ as shown in the middle panel of figure~\ref{fig:sketch}.
-$\alpha$ is the angle between the optical axis and the line from the point where the optical axis intersects the Rowland torus (labelled ``A'' in the figure) and the focal point of the mirror (``F'').
-We call the distance between those two points $a$, which is chosen to be slightly shorter than the focal length $f$ of the mirror to allow for space to mount the gratings behind the mirror.
-In this configuration, the torus has $R < r$ there are two points where the torus overlaps itself in the plane spanned by the optical axis and the symmetry axis of the torus (the plane shown in figure~\ref{fig:sketch}).
+\citet{doi:10.1117/12.856482} suggested tilting the torus with respect to the optical axis when using critical angle transmission (CAT) gratings \citep{2021hai4.book..235H,2022ApJ...934..171H}.
+In CAT gratings, the grating bars have a high aspect ratio, that is the distance between two gratings bars is much smaller than the height of the bars.
+CAT gratings are typically used at a blaze angle, where the grating bars are inclined with respect to the incoming rays.
+CAT gratings preferentially diffract photons to twice the blaze angle and can reach high diffraction efficiencies.
+However, the blaze angle also means that the gratings can deviate significantly from the surface of the  Rowland torus.
+To minimize this deviation, the Rowland torus can be tilted.
+This tilt can be described by two angles $\alpha$ and $\beta$ as shown in figure~\ref{fig:sketch_label}.
+$\alpha$ is the angle between the optical axis and the line from the point where the optical axis intersects the Rowland torus (labelled ``A'' in the figure) and the center of the Rowland circle.
+We choose the distance between point A and the focal point F to be slightly shorter than the focal length $f$ of the mirror to allow for space to mount the gratings behind the mirror and use the symbol ``a'' for the distance between A and F.
+In this configuration, the torus has $R < r$ so there are two points where the torus overlaps itself in the plane spanned by the optical axis and the symmetry axis of the torus (the plane shown in figure~\ref{fig:sketch_label}).
 The first of those intersection points is F.
 We call the second point ``Hinge''.
-$\beta$ is the angle from optical axis to the line connecting A and the Hinge. With that we can derive the radius of the Rowland circle to
-$$r = \frac{a}{2}\cos \alpha.$$
- If the focal point F is the origin of the coordinate system, then the center of the Rowland circle (blue squares in left blue circles in the panels in figure~\ref{fig:sketch}) is at location
+$\beta$ is the angle from optical axis to the line connecting A and the Hinge. From this we can derive the radius of the Rowland circle as
+$$r = \frac{a}{2\cos \alpha}.$$
+ If the focal point F is the origin of the coordinate system, then the center of the Rowland circle (blue square in left blue circle in the panels in figure~\ref{fig:sketch_label}) is at location
  \begin{equation}
     \begin{pmatrix} x \\ y \\ z \end{pmatrix} =
-    \begin{pmatrix} 0 \\ \frac{a}{2} \\  \frac{a}{2}\tan\alpha \end{pmatrix}.
+    \begin{pmatrix} \frac{a}{2} \tan\alpha \\0 \\   \frac{a}{2} \end{pmatrix}.
  \end{equation}
 
+(In this coordinate system the astrophysical source is located at $z=+\infty$.)
 Since A, F, and the Hinge are by definition all on the Rowland circle, their distance to this point is $r$.
-With this symmetry, we can obtain the angle of the axis of the torus with respect to the case where the axis of the torus is perpendicular to the optical axis (left panel of figure~\ref{fig:sketch}).
+With this symmetry, we can obtain the angle of the axis of the torus with respect to the $x$-axis.
 This angle is $\beta-\alpha$.
@@
 \begin{equation}
-    R = r \sin\left(\frac{\pi}{2} - \beta\right),
+    R = r \sin\left(\frac{\pi}{2} - \beta\right) = r \cos\beta\ ,
 \end{equation}
@@
     \begin{pmatrix} x \\ y \\ z \end{pmatrix} =
-    \begin{pmatrix} 0 \\ \frac{a}{2}-R\cos{\beta-\alpha} \\ \frac{a}{2}\tan{\alpha}+R\sin{\beta-\alpha} \end{pmatrix}.
+    \begin{pmatrix} \frac{a}{2}\tan{\alpha}+R\sin{(\beta-\alpha)} \\ 0 \\
+        \frac{a}{2}-R\cos{(\beta-\alpha)} \end{pmatrix}.
 \end{equation}
 
-CAT gratings are blazed and thus deviate more from the Rowland-torus than a grating that is placed perpendicular to the incoming rays.
-Figure~\ref{fig:3d:tilted} shows this layout in 3D, but with a very large blaze angle for clearity; for real CAT gratings typical blaze angles are 1-2~deg.
-Tilting the Rowland torus by about twice the blaze angle reduces this effect.
+Assuming CAT gratings where the gratings bars are perpendicular to the surface of the grating, the gratings need to be installed blazed (tilted) with respect to the incoming rays.
+Thus, they deviate more from the Rowland-torus than a grating that is placed perpendicular to the incoming rays.
+Figure~\ref{fig:3d:tilted} shows this layout in 3D, but with a very large blaze angle for clarity; for real CAT gratings typical blaze angles are 1-2~deg.
+Tilting the Rowland torus by about twice the blaze angle reduces the average deviation of the grating bar position from te surface of the Rowland torus.
 Also, CAT gratings disperse predominantly in one direction.
-However, the Rowland geometry optimizes the spectral resolving power by minimizing the spread of the photons in cross-dispersion direction; detectors are located on the Rowland circle, which represents the position of minimal spread in dispersion direction.
+However, the Rowland geometry optimizes the spectral resolving power by minimizing the spread of the photons in dispersion direction; detectors are located on the Rowland circle, which represents the position of minimal spread in dispersion direction.
 This is not the same location as the imaging focal plane, which is defined by minimizing the total size of the PSF.
 At the focal point, the Rowland circle and the imaging focal plane match.
-Figure~\ref{fig:fish} shows how the dispersed photons spread out in cross-dispersion direction, roughly resembling the outline of a fish.
-With a tilted Rowland torus, there is a second intersection point between the spectral and the imaging focus where the cross-dispersion width becomes small (the point between the body and the tail of the fish, about +300 in dispersion direction in the second panel of figure~\ref{fig:fish}).
+Figure~\ref{fig:fish} shows how the dispersed photons spread out in cross-dispersion direction with increasing distance from F, roughly resembling the outline of a fish.
+With a tilted Rowland torus, there is a second intersection point between the spectral and the imaging focus where the cross-dispersion width becomes small (the point between the body and the tail of the fish, the ``Hinge'', about +300~mm in dispersion direction in the second panel of figure~\ref{fig:fish}).
 This location can be chosen to be roughly in the center of the dispersed spectrum and thus limit the cross-dispersion profile to reduce the area of the detector that contributes to the extracted signal and thus the background.
@@
 \begin{figure}
-    \plotone{ifigs/tilted}
+    \begin{interactive}{js}{ifigs/tilted.zip}
+    \includegraphics[width=0.22\textwidth]{ifigs/tilted_overview}
+    \includegraphics[width=0.22\textwidth]{ifigs/tilted_aperture}
+    \includegraphics[width=0.22\textwidth]{ifigs/tilted_gratings}
+    \includegraphics[width=0.22\textwidth]{ifigs/tilted_detector}
+    \end{interactive}
   \caption{Ray-trace like figure~\ref{fig:3d:single}, but for a tilted Rowland torus setup.
@@
 
-A tilted Rowland can also be combined with sub-aperturing to increase the spectral resolving power as discussed in figure~\ref{fig:subaperture}.
-In this case, not all of the fish shape is filled with photons.
-This can be seen by comparing the dark purple points in the second, third, and forth panel in figure~\ref{fig:fish}.
-The Rowland torus for this channel has the same tilt and dimensions in all three designs, but in the second panel, the mirror fills the full circle, in the third panel it has just two wedges, and in the forth panel just one.
+A tilted Rowland torus can also be combined with sub-aperturing to increase the spectral resolving power as discussed in figure~\ref{fig:subaperture}.
+In this case, not all the fish shape is filled with photons.
+This can be seen by comparing the dark purple points in the second, third, and fourth panel in figure~\ref{fig:fish}.
+The Rowland torus for this channel has the same tilt and dimensions in all three designs, but in the second panel, the mirror fills the full circle, in the third panel it has just two wedges, and in the fourth panel just one.
 In either case, the spectral extraction region used in data analysis should take this shape into account to minimize detector background.
@@
 \subsection{Double tilted Rowland Torus}
+
 \begin{figure}
-    \plotone{ifigs/tilted_double}
+    \begin{interactive}{js}{ifigs/double.zip}
+    \includegraphics[width=0.22\textwidth]{ifigs/double_overview}
+    \includegraphics[width=0.22\textwidth]{ifigs/double_aperture}
+    \includegraphics[width=0.22\textwidth]{ifigs/double_gratings}
+    \includegraphics[width=0.22\textwidth]{ifigs/double_detector}
+    \end{interactive}
   \caption{Ray-trace like figure~\ref{fig:3d:tilted}, but with channels split on two separately tilted Rowland tori.
-See figure~\ref{fig:3d:single} for an explanation of the elements shown in the image and the interactivity in the online version.
-        }
+See figure~\ref{fig:3d:single} for an explanation of the elements shown in the image and the interactivity in the online version.}
   \label{fig:3d:tilted_double}
@@
 This can be compensated by adding a second channel, with mirrors mounted next to the mirrors of the first channel.
-Again, gratings and detectors for this second channel need to be positioned on a Rowland torus, however the two Rowland tori for the two channels need not be identical.
+Again, gratings and detectors for this second channel need to be positioned on a Rowland torus.
+However the two Rowland tori for the two channels need not be identical.
 When $R> r$ the Rowland circle intersects the focal plane twice.
@@
 Figure~\ref{fig:3d:tilted_double_offset} shows a 3D ray-trace image in this configuration; the resulting signal on the detector is shown in figure~\ref{fig:fish} (bottom panel).
+
 \begin{figure}
-    \plotone{ifigs/tilted_double_offset}
+    \begin{interactive}{js}{ifigs/offset.zip}
+    \includegraphics[width=0.22\textwidth]{ifigs/offset_overview}
+    \includegraphics[width=0.22\textwidth]{ifigs/offset_aperture}
+    \includegraphics[width=0.22\textwidth]{ifigs/offset_gratings}
+    \includegraphics[width=0.22\textwidth]{ifigs/offset_detector}
+    \end{interactive}
     \caption{Ray-trace like figure~\ref{fig:3d:tilted_double}, but with offsets in dispersion and cross-dispersion direction.
@@
 \end{figure}
+
 \subsubsection{Channel offsets in the cross-dispersion direction}
@@
 To do that, it is necessary to know which optical channel a photon passed through and thus the setup needs to avoid overlapping photons from both channels on the same location of the detector.
-This can be accomplished by shifting the Rowland tori (and mirrors and gratings positioned on them) in the cross-dispersion direction (into or out of the plane of the plot in figure~\ref{fig:sketch}).
+This can be accomplished by shifting the Rowland tori (and mirrors and gratings positioned on them) in the cross-dispersion direction (into or out of the plane of the plot in figurs~\ref{fig:sketch_label} and~\ref{fig:sketch}).
 The shift should be large enough to clearly separate the two channels, but keep both dispersed spectral traces on the detector with margin (figure~\ref{fig:fish}).
@@
 These can then be arranged closely together (see figure~\ref{fig:3d:tilted_double}) to maximize the area used given the size constraints of the spacecraft.
-In this case, it is an option to split each of the two channels into two sub-channels with their own separate offset, such that there are two parallel traces running in each direction (Figure~\ref{fig:3d:tilted_double_offset}) for a total of four traces (figure~\ref{fig:fish}, right panel).
-Without this split, the opposing mirror and grating wedges have to be carefully aligned to the same optical axis, any misalignment between them would broaden the PSF and thus reduce the spectral resolving power.
-With this split, each wedge defines its own optical axis and the exact location and thus the wavelength scale for each zeroth order can be determined from observations.
+In this case, it is an option to split each of the two channels into two sub-channels with their own separate offset, such that there are two parallel traces running in each direction (Figure~\ref{fig:3d:tilted_double_offset}) for a total of four traces (figure~\ref{fig:fish}, bottom panel).
+Without this split, the opposing mirror and grating wedges have to be carefully aligned to the same optical axis, and any misalignment between them would broaden the PSF and thus reduce the spectral resolving power.
+With this split, each wedge defines its own optical axis and the exact location, and thus the wavelength scale for each zeroth order can be determined from observations.
 
@@
 
-\subsection{Subaperturing space considerations, and resolving power}
-Manufacturing mirrors with a smaller PSF is more difficult and costly than manufacturing mirrors with a larger PSF, but otherwise identical properties.
+\subsection{Advantages of the DTRT concept}
+\subsubsection{Subaperturing space considerations, and resolving power}
+Manufacturing mirrors with a smaller PSF is more difficult and costly than manufacturing mirrors with a larger PSF but otherwise identical properties.
 For a spectrograph, the width of the PSF in dispersion direction is far more important than the width in cross-dispersion direction.
@@
 
-\subsection{Zeroth order is seen}
+\subsubsection{Zeroth order is seen}
 Observing the direct image of all channels on the detector has several advantages.
 First, most of the high-energy photons are not diffracted by typical CAT gratings.
-So, the direct image can be used for analysis of high-energy spectrum using the intrinsic energy resolution of the detector, e.g.\ a CCD or even microcalorimeter.
-For example, the Fe line at 6.7~keV is a widely used diagnostic for many astrophysical objects.
+So, the direct image can be used for analysis of high-energy spectra using the intrinsic energy resolution of the detector, e.g.\ a CCD or even microcalorimeter.
+For example, the Fe line at 6.7~keV is a widely used diagnostic for many astrophysical objects; in a DTRT design this line can be observed with an energy sensitive detector such as a CCD in the zeroth order, while the lower-energy photons are observed in the dispersed orders at the same time.
 Second, accurate knowledge of the position of the direct image is important for the wavelength calibration of the dispersed signal.
@@
 
-\subsection{Number of detector elements}
-An instrument that uses only a single tilted Rowland torus needs to cover a long strip of dispersed light with a detector and a small area around the zeroth order.
-On the other hand, a DTRS needs two long detector strips to cover the dispersed signal on both sides; those cameras will cover the zeroth order of all channels at the same time.
-
-\subsection{Robustness}
+\subsubsection{Robustness}
 The DTRS concept increases the robustness and reduces the technical risk of a mission in several ways.
-Because we can assign each sub-apertured mirror wedge its own channel (in the 4 channel design with offsets), there is no need to align the mirror for each channel to each other to better than the cm scale.
+Because we can assign each sub-apertured mirror wedge its own channel (in the 4 channel design with offsets), there is no need to align the mirrors for each channel to each other to better than a cm or so.
 Also, the design is robust against loosing individual detectors in a camera.
@@
 \subsection{Drawbacks}
+\subsubsection{Number of detector elements}
+An instrument that uses only a single tilted Rowland torus needs to cover a long strip of dispersed light with a detector and a small area around the zeroth order.
+On the other hand, a DTRS needs two long detector strips to cover the dispersed signal on both sides; those cameras will cover the zeroth order of all channels at the same time.
+
+\subsubsection{Background and source confusion}
 Because the DTRS concept spreads the dispersed photons over a larger detector area than a single-channel design, the background contamination in the spectra is higher, simply because more pixels contribute to the extracted science spectrum.
 This does not matter for bright isolated point-source, but makes observations of faint sources, where the dispersed spectrum has count rates comparable to the background, difficult.
-Similarly, observations where there is extended emission in the field-of-view, for example in the galactic center, are difficult.
-The DTRS concept is also not well-suited for crowded fields, where the dispersed signal of one source overlaps with the dispersed signal of another source.
+Similarly, observations are difficult when extended emission or many point sources are in the field-of-view, for example in the galactic center or in star forming regions.
+In this case, spectra from different locations on the sky overlap on the detector.
 
+\subsubsection{Complexity of spectral fitting}
 Depending on the detector technology chosen, the DTRS concept might also lead to undesirable coupling between the spectra of different channels.
@@
 The reflection gratings are thus positioned almost perpendicular to the path of the rays tangential to the surface of the Rowland torus.
-Thus, a few gratings (just one per channel in the extreme case) is sufficient and, even if the grating itself is flat, it never deviates much from the surface of the Rowland torus.
+Thus, a few gratings (just one per channel in the extreme case) are sufficient and, and even if the grating itself is flat, it never deviates much from the surface of the Rowland torus.
 Since UV light has a longer wavelength, it can be dispersed to larger angles increasing the spectral resolving power.
@@
 \label{sect:applications}
-In the previous sections, we discussed the concept of a DTRS in general.
-The Arcus X-ray spectrograph \citep{2023SPIE12678E..0ES} adopts this concept with $\alpha=3.6$~deg and $\beta=2\alpha$.
-Arcus has gone through multiple design iterations for different mission calls and the design will continue to be refined.
-As an example, we describe the Arcus setup here as submitted to the NASA probe call in 2023.
-Arcus uses silicon-pore optics (SPO) with a 12~m focal length that are developed for the Athena mission \citep{2023SPIE12679E..05G}; $a$ is slightly smaller than the focal length with $a=11876$~mm.
+As a concrete example, we describe the Arcus X-ray spectrograph  \citep{2023SPIE12678E..0ES}.
+Arcus has gone through multiple iterations for different mission calls, most recently as a NASA X-ray Probe concept submitted in 2023.
+This mission adopts a DTRS design with $\alpha=3.6$~deg and $\beta=2\alpha$.
+It uses silicon-pore optics (SPO) with a 12~m focal length that are developed for the Athena mission \citep{2023SPIE12679E..05G}; $a$ is slightly smaller than the focal length with $a=11876$~mm.
 SPOs are manufactured as small stacks of about 30 mirror plates with sizes of order 10-20~cm in each direction (the exact size depends on the radius).
@@
 In cross-dispersion the four channels are located at $-7.5$, $-2.5$, $+2.5$, and $+7.5$~mm from the center of the CCDs.
-Critical-angle transmission (CAT) gratings are manufactured through deep reactive-ion etching of silicon-on-insulator wafers \citep{2021SPIE11822E..15H,2023SPIE12679E..0LH}.
+CAT gratings are manufactured through deep reactive-ion etching of silicon-on-insulator wafers \citep{2022ApJ...934..171H,2023SPIE12679E..0LH}.
 These gratings are mounted with a blaze angle, i.e.\ they are intentionally tilted by 1.8~deg with respect to the incoming photons, which concentrates the diffracted photons around the ``blaze peak'' at a diffraction angle of $2 * 1.8 = 3.6$~deg, so that only 8~CCDs are needed in each camera covering about 30~cm in the dispersion direction with a gap close to 50~cm between the two cameras.
@@
     \plotone{EQFullDet}
-    \caption{Simulation of an Arcus observation of an emission lines spectrum (see section~\ref{sect:applications} for details).
-Position of the CCDs are outlined with a white frame and the zeros orders are indicated.
-Dashed arrows mark the dispersion direction of the four orders from the zeroths order.
+    \caption{Simulation of an Arcus observation of an emission line spectrum (see section~\ref{sect:applications} for details).
+Positions of the CCDs are outlined with white frames and the zeros orders are indicated.
+Dashed arrows mark the dispersion direction of the four spectra from the zeroths order.
 Only photons that are detected on a CCD are included.
@@
 That reduces the deviation of flat gratings from the surface of the torus and also allows us to position a second channel next to the first one, doubling the effective area compared to a single, sub-apertured channel.
-Both channels have different Rowland tori (dobule tilted Rowland spectrograph, DTRS), but they overlap in the Rowland circle, and thus both channels can be imaged onto the same set of detectors.
+Both channels have different Rowland tori (double tilted Rowland spectrograph, DTRS), but they overlap in the Rowland circle, and thus both channels can be imaged onto the same set of detectors.
 In this geometry, all zeroths orders are visible for wavelength calibration and to detect hard X-rays in direct imaging.
 Finally, we discuss how small offsets between the channels can mitigate chip gaps and reduce the alignment requirements between optical elements during the assembly of the instrument.
-We illustrate these concepts with sketches and ray-traces designed to illustrate the concept.
+We illustrate these concepts with sketches and ray-traces.
 
-We briefly discuss Arcus, one NASA probe class mission concept that implements a DTRS and show realistic ray-traces for a 100~ks observation of an emission line dominated spectrum with Arcus.
+We briefly discuss Arcus, one NASA probe class mission concept that implements a DTRS and show realistic ray-traces for a 100~ks observation of an emission line dominated spectrum observed by Arcus.
 
 \begin{acknowledgements}
-Support for this work was provided in part through NASA grant NNX17AG43G and Smithsonian Astrophysical Observatory (SAO)
+Support for this work was provided in part through NASA grant 80NSSC22K1904 and Smithsonian Astrophysical Observatory (SAO)
 contract SV3-73016 to MIT for support of the {\em Chandra} X-Ray Center (CXC),
 which is operated by SAO for and on behalf of NASA under contract NAS8-03060.
-
 \end{acknowledgements}
 
-\software{marxs \citep{2017AJ....154..243G}, AstroPy \citep{2013A&A...558A..33A,2018AJ....156..123A}, NumPy \citep{harris2020array}, Matplotlib \citep{Hunter:2007}, IPython\cite{IPython}}
+\software{marxs \citep{2017AJ....154..243G}, AstroPy \citep{2013A&A...558A..33A,2018AJ....156..123A}, NumPy \citep{harris2020array}, Matplotlib \citep{Hunter:2007}, IPython \citep{IPython}}
 

### 2024-03-25 papers/Doubletiltconcept/doubletiltconcept.tex
ApJ double tilt paper ready for submission

@@
 The DTRS concept is especially useful for sub-apertured mirrors, because it allows an effective use of space in the entrance aperture of a spacecraft.
-One mission that applies this concept is the Arcus probe.
+One mission that applies this concept is the Arcus Probe.
 
@@
 Thus, we discuss advantages and challenges of this layout in general.
-However, we note that Arcus \citep{2023SPIE12678E..0ES}, an X-ray spectroscopy mission proposed for NASA MIDEX (medium explorer) and probe calls, applies this layout.
+However, we note that Arcus \citep{2023SPIE12678E..0ES}, an X-ray spectroscopy mission proposed for NASA MIDEX (medium explorer) and Probe calls, applies this layout.
 Specific details of the  Arcus performance, such as predicted effective area and resolving power are shown in \citet{2018SPIE10699E..6FG,2023SPIE12678E..1DG}, but the general concept of a double tilted Rowland spectrograph (DTRS) with its advantages and drawbacks has not been discussed -- a void we attempt to fill with this article.
 
-The structure of this article is as follows: First, we discuss the properties on the optical elements that are most influential for the design of a DTRS, namely mirror properties, gratings, and detectors in Sect.~\ref{sect:elements}.
+The structure of this article is as follows: First, we discuss the properties of the optical elements that are most influential for the design of a DTRS, namely mirror properties, gratings, and detectors in Sect.~\ref{sect:elements}.
 Then, we illustrate the optical layout, modifying the design of a traditional single-channel Rowland torus spectrograph to a DTRS.
@@
 In X-rays, the mirror assembly will typically consist of nested (full or partial) shells that reflect X-rays in grazing incidence.
-The higher the energy of the X-ray, the smaller the angle between the mirror surface and the incoming ray has to be to allow for reflection.
+The higher the energy of the X-ray, the smaller the angle between the mirror surface and the incoming ray has to be to allow for efficient reflection.
 The size of the collecting area is to first order proportional to the number of photons that the spectrograph can detect, and the spectral resolving power $\lambda / \Delta \lambda$ increases when the mirror point-spread-function (PSF) decreases.
@@
 If the PSF is dominated by the figure error or scattering by microroughness or particulates, then scattering angles in the plane of reflection are typically larger than the out-of-plane scatter.
-In this case, sampling a limited azimuthal span of the aperture -- subaperturing -- can improve the spectral resolving power \citep{1987ApOpt..26.2915C}.
+In this case, sampling a limited azimuthal span of the aperture -- subaperturing -- can improve the spectral resolving power \citep{1987ApOpt..26.2915C,2010SPIE.7732E..1JH}.
 This is illustrated in Fig.~\ref{fig:subaperture}.
@@
 \citet{2020SPIE11444E..88G} explored bending the gratings in one dimension or varying the distance between grating bars over a grating to compensate for the deviations from the ideal shape in simulations.
-However, these concepts have not yet been implemented in X-ray transmission gratings.
+\citet{2019SPIE11118E..11G} shows that CAT gratings can be bent to the required radii without degrading their performance.
 
@@
 The location of the mirror is shown as blue box.
-The optical axis (solid blue line) is offset from the center of the Rowland circle (blue square) and is perpendicular to the symmetry axis of the torus (blue long-dashed line).
+The optical axis (solid blue line) is offset from the center of the Rowland circle (blue square).
+The symmetry axis of the torus  is marked as blue long-dashed line.
 The two blue circles show the cut through the 3D torus in this plane; they are connected by blue dotted lines to help visualize the 3D position of the torus.
@@
     \emph{right:} Layout with two Rowland tori, tilted in opposite direction.
-In the plane spanned by both optical axes (which are parallel) and the symmetry axis of the torus, the two tori overlap in exactly one circle, the Rowland circle.
+In the plane spanned by both optical axes (which are parallel) and the symmetry axis of the tori, the two tori overlap in exactly one circle, the Rowland circle.
 See figure~\ref{fig:sketch_label} for a more detailed explanation.
@@
   \end{interactive}
-\caption{3D image of the ray-trace for a source with a continuum spectrum over a limited bandpass, seen from the side, from the top (where the rays enter the aperture), zoom on the grating positions, and zoom on the detector.
+\caption{3D images of the ray-trace for a source with a continuum spectrum over a limited bandpass, seen (from left to right) from the side, from the top (where the rays enter the aperture), zoom on the grating positions, and zoom on the detector.
 Rays are shown from the entrance aperture on.
@@
 They are detected by CCDs (yellow).
-Rays are colored by the diffraction order: Light purple (middle) are zero order photons.
+Rays are colored by the diffraction order: Light purple are zero order photons.
 A section of the Rowland torus is shown as transparent gray.
@@
 This could be avoided with larger offsets between channels in cross-dispersion direction.
-Also, in real instruments with focal lengths significantly larger than in the illustrative model we use throughout this article, the bending and the width in cross-dispersion direction are typically smaller, so that this is not a problem in practice.`'
+Also, in real instruments with focal lengths significantly larger than in the illustrative model we use throughout this article, the bending and the width in cross-dispersion direction are typically smaller, so that this is not a problem in practice.
         }
@@
 
-\citet{doi:10.1117/12.856482} suggested tilting the torus with respect to the optical axis when using critical angle transmission (CAT) gratings \citep{2021hai4.book..235H,2022ApJ...934..171H}.
+\citet{doi:10.1117/12.856482} suggested tilting the torus with respect to the optical axis when using critical-angle transmission (CAT) gratings \citep{2021hai4.book..235H,2022ApJ...934..171H}.
 In CAT gratings, the grating bars have a high aspect ratio, that is the distance between two gratings bars is much smaller than the height of the bars.
@@
 Figure~\ref{fig:3d:tilted} shows this layout in 3D, but with a very large blaze angle for clarity; for real CAT gratings typical blaze angles are 1-2~deg.
-Tilting the Rowland torus by about twice the blaze angle reduces the average deviation of the grating bar position from te surface of the Rowland torus.
+Tilting the Rowland torus by about twice the blaze angle reduces the average deviation of the grating bar position from the surface of the Rowland torus.
 Also, CAT gratings disperse predominantly in one direction.
@@
 For a spectrograph, the width of the PSF in dispersion direction is far more important than the width in cross-dispersion direction.
-If the width in dispersion direction can be reduced by sub-aperturing, that provides a higher spectral resolving power at lower cost and technical complexity compared to reducing the PSF of the mirror itself.
+If the width in dispersion direction can be reduced by sub-aperturing, a higher spectral resolving power is achieved at lower cost and technical complexity compared to reducing the PSF of the mirror itself.
 On the other hand, it reduces the collecting area of the mirror, since only a fraction of the full circle can be used.
-The DTRS design allows us to compensate that by having multiple channels, ultimately delivering better resolving power at the same mirror requirements than a single-channel design.
+The DTRS design allows us to compensate for this by having multiple channels, ultimately delivering better resolving power at the same mirror requirements than a single-channel design.
 
@@
 Observing the direct image of all channels on the detector has several advantages.
-First, most of the high-energy photons are not diffracted by typical CAT gratings.
+First, most of the high-energy photons are neither diffracted nor absorbed by typical CAT gratings.
 So, the direct image can be used for analysis of high-energy spectra using the intrinsic energy resolution of the detector, e.g.\ a CCD or even microcalorimeter.
@@
 The wavelength of the light is determined by measuring the distance $x$ from the detected photon to the zeroth order and the known distance $a$ between the gratings and the focal plane.
-Conceptually, $\tan \theta = x/a$.
+Conceptually, $\tan \theta \approx x/a$.
 With the known grating constant $d$, equation~\ref{eqn:diffraction} then yields $\lambda$.
@@
 The DTRS concept increases the robustness and reduces the technical risk of a mission in several ways.
-Because we can assign each sub-apertured mirror wedge its own channel (in the 4 channel design with offsets), there is no need to align the mirrors for each channel to each other to better than a cm or so.
+Because we can assign each sub-apertured mirror wedge its own channel (in the 4 channel design with offsets), there is no need to align the mirrors for each channel to the other channel to better than a cm or so.
 Also, the design is robust against loosing individual detectors in a camera.
@@
 These gratings are mounted with a blaze angle, i.e.\ they are intentionally tilted by 1.8~deg with respect to the incoming photons, which concentrates the diffracted photons around the ``blaze peak'' at a diffraction angle of $2 * 1.8 = 3.6$~deg, so that only 8~CCDs are needed in each camera covering about 30~cm in the dispersion direction with a gap close to 50~cm between the two cameras.
-Arcus will achieve a spectral resolving power around 3500 for wavelength between 1.5 and 6.0~nm, averaged over all diffraction orders, and an effective area up to 500~cm$^2$ between 1.5 and 2.5~nm, dropping to 100~cm$^2$ at 6.0~nm.
+Arcus will achieve a spectral resolving power around 3500 for wavelengths between 1.5 and 6.0~nm, averaged over all diffraction orders, and an effective area up to 500~cm$^2$ between 1.5 and 2.5~nm, dropping to 100~cm$^2$ at 6.0~nm.
 Detailed ray-tracing simulations for Arcus that optimize the parameters of the design and predict the performance depending on the misalignment for the different components are shown in \citet{2017SPIE10397E..0PG,2018SPIE10699E..6FG,2023SPIE12678E..1DG}.
@@
 We tilt the Rowland torus such that the optical axis of the mirror is offset from the center of the Rowland circle.
-That reduces the deviation of flat gratings from the surface of the torus and also allows us to position a second channel next to the first one, doubling the effective area compared to a single, sub-apertured channel.
+That reduces the deviation of flat, tilted gratings from the surface of the torus and also allows us to position a second channel next to the first one, doubling the effective area compared to a single, sub-apertured channel.
 Both channels have different Rowland tori (double tilted Rowland spectrograph, DTRS), but they overlap in the Rowland circle, and thus both channels can be imaged onto the same set of detectors.
@@
 
-We briefly discuss Arcus, one NASA probe class mission concept that implements a DTRS and show realistic ray-traces for a 100~ks observation of an emission line dominated spectrum observed by Arcus.
+We briefly discuss Arcus, one NASA Probe class mission concept that implements a DTRS and show realistic ray-traces for a 100~ks observation of an emission line dominated spectrum observed by Arcus.
 

### 2024-04-01 papers/special_JATIS_tolerances/tolerances.tex
Special JATIS paper I: First draft

\documentclass[12pt]{spieman}  % 12pt font required by SPIE;
%\documentclass[a4paper,12pt]{spieman}  % use this instead for A4 paper
\usepackage{amsmath,amsfonts,amssymb}
\usepackage{graphicx}
\usepackage{setspace}
\usepackage{tocloft}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage[table]{xcolor}
\include{moritz}

\title{Arcus X-ray telescope performance predictions and alignment requirements}

\author[a]{Hans Moritz G\"unther$^*$}
\author[b]{Peter Cheimets}
\author[c]{Casey T. DeRoo}
\author[a,d]{Ralf K. Heilmann}
\author[a]{Eric D. Miller}
\author[e]{Andrew Ptak}
\author[b]{Randall K. Smith}

\affil[a]{MIT Kavli Institute for Astrophysics and Space Research, Cambridge, MA 02139, USA}
\affil[b]{Center for Astrophysics, Harvard-Smithsonian Astrophysical Observatory, Cambridge, MA 02138, USA}
\affil[c]{Dept. of Physics \& Astronomy, University of Iowa, Iowa City, IA 52242, USA}
\affil[d]{Space Nanotechnology Laboratory, MIT Kavli Institute for Astrophysics and Space Research, Cambridge, MA 02139, USA}
\affil[e]{NASA Goddard Space Flight Center, Greenbelt, MD 20771, USA}

\renewcommand{\cftdotsep}{\cftnodots}
\cftpagenumbersoff{figure}
\cftpagenumbersoff{table}
\begin{document}
\maketitle

\begin{abstract}
Arcus is a concept for a probe class mission to deliver high-resolution FUV and X-ray spectroscopy with two separate instruments. The X-ray spectrograph (XRS) is the focus of this paper. It consists of four spectral channels arranged in a double-tilted Rowland torus geometry.  It combines cost-effective silicon pore optics (SPO) with high-throughput critical-angle transmission (CAT) gratings to achieve $R> 3000$ in a bandpass from 12-50 \AA. We present ray-tracing studies to derive performance characteristics such as the spectral resolving power and effective area, and look at the best positioning of the four channels to improve the resiliency towards misalignments and reduce the overall impact of chip gaps. We study the effect of misalignments on the performance, and present alignment requirements in six degrees of freedom for all optical elements in the XRS. We
conclude that most tolerances can be achieved with mechanical means alone.
\end{abstract}

% Include a list of up to six keywords after the abstract
\keywords{Arcus, ray-tracing, tolerances, alignment, X-ray}

% Include email contact information for corresponding author
{\noindent \footnotesize\textbf{*}Hans Moritz G\"unther,  \linkable{hgunther@mit.edu} }

% FOR SUBMISSION
% \begin{spacing}{2}   % use double spacing for rest of manuscript

% For my own while writing. It's just easier to read this way
\begin{spacing}{1}

\section{Introduction}
\label{sect:intro}  % \label{} allows reference to this section
High-resolution X-ray and UV spectroscopy open a window into the physics of the universe that often cannot be observed with any other technique. X-rays and UV can probe the hottest and most ionized gas that remains invisible in longer wavelengths because the high ionization levels do not produce observable transitions in the radio, infrared or optical. Thus, observations in the UV and X-rays probe the most energetic processes in a number of systems. A few examples are: stellar space weather \cite{2022AN....34320019B}, the accretion onto young stars, where the UV and X-rays come from the energetic infall of material from the disk onto the stellar surface; the innermost regions of the accretion disks around black holes; and the absorption of background (X-ray) light from distant, bright continuum sources by the warm-hot inter-galactic medium to find the missing baryons\cite{10.1117/12.2231193}. High-resolution UV and X-ray spectroscopy that resolves the profiles of individual emission or absorption lines is particularly valuable, because it allows us to address a host of physical questions that cannot be answered by simply measuring the broad-band X-ray flux.
To follow-up with one specific example: in the case of accretion onto young stars, resolving the kinematic line profile of emission in different ions can tell us which part of the emission is formed in the infalling, red-shifted accretion column, which part is related to the stellar corona seen at the rest wavelength of the star (possibly with small velocities from up and downflows in the corona), and which part is formed in (or possibly absorbed by) the blue-shifted outflow.

The Arcus mission is a concept that will address those challenges with two instruments for X-ray and UV high-resolution spectroscopy. The mission evolved through several stages. It was originally
proposed as an instrument mounted on the International Space Station\cite{10.1117/12.2062671} and then redesigned as a satellite\cite{10.1117/12.2231778,10.1117/12.2272818}. Arcus' X-ray spectrograph (XRS) will perform
high-resolution spectroscopy in the soft X-ray range (about 12-50~\AA{}) with a resolving power $R>3000$ and an effective area $A_\textrm{eff}>
300\;\mathrm{cm}^2$ for most of the bandpass with a peak close to $A_\textrm{eff} > 600\;\mathrm{cm}^2$ around 18~\AA{}. The resolving power is a factor of
$3-5$ larger than what existing instruments on Chandra or XMM-Newton can
deliver and the effective area is also significantly higher; the exact number
depends on the bandpass, e.g.\ in the crucial region around the O~{\sc vii}
He-like triplet, which is density and temperature sensitive, Arcus will reach
about two orders of magnitude more effective area than Chandra/HETG currently
has. The UV spectrograph (UVS) is described by ref.~\citenum{10.1117/12.2677948} and the UV science case is discussed by ref.~\citenum{10.1117/12.2677332}. In this work, we concentrate on the XRT.

Beginning in the very early phases of the mission design we performed
ray-tracing to verify the performance characteristics of Arcus and to
improve the design\cite{10.1117/12.2232157}; in particular in \citenum{10.1117/12.2312678} we utilized ray-tracing to determine the allowable misalignments of the mechanical elements that make up Arcus. Ray-tracing is particularly well suited for the problem, because it allows an arbitrary misalignment to be introduced for any element in the optical path and it can quickly iterate over different input parameters. Since the work in \citenum{10.1117/12.2312678}, the Arcus concept has evolved yet again. In its most recent version, Arcus is a NASA Probe-class mission with significantly increased effective area and the addition of the UV instrument. Since the UV telescope has a smaller field-of-view than the X-ray instrument, it will define the fiducial pointing direction; now misalignment between the two separate instruments is a new concern.

\section{Layout of the X-ray Spectrograph on Arcus}
The XRS on Arcus follows the layout of a double-tilted Rowland Spectrograph (DTRS). This concept is described in detail in Ref.~\citenum{DTRS} and summarized in this section. The XRS has four channels, each with their own separate X-ray mirror and grating petal. However, the photons from all channels are imaged onto the same set of detectors. Figure~\ref{fig:arcus} shows a rendering of the optical elements of the XRS and some photon pathways through it.

\begin{figure}
    \centering
    \includegraphics[height=8cm]{Arcus.png}
    \caption {\label{fig:arcus}
    Arcus to scale. On the left are the four channels with SPOs (green) and CAT gratings (white). The CCDs are in the background, where the lines converge. Lines show ray-traces. Different colors indicate different diffraction orders. Only rays that reach the detector are shown.
}
\end{figure}

\subsection{Silicon Pore Optics (SPOs)}
The SPOs are a technology developed and matured for Athena. They provide a large effective area at a relatively low weight and cost\cite{10.1117/12.2188988,10.1117/12.2599339,10.1117/12.2677388}. In Arcus, the SPO petals use ``sub-aperturing'', which means that the SPOs do not cover a full circle, but only a narrow wedge. That provides Arcus with an asymmetric PSF that is only about 1.5\ arcsec wide in the dispersion direction. It also means that the individual petals where the SPOs are mounted are not circular, but roughly rectangular, and four separate panels (one per optical channel) can fit into the front assembly of the telescope; see Figure~\ref{fig:arcus} for a rendering.

\subsection{Critial-Angle Transmission (CAT) gratings}
Critical-angle transmission (CAT) gratings are manufactured at the Space Nanotechnology Labo\-ra\-tory. Progress on the development and manufacturing of the CAT gratings is given in a series of papers stretching back many years\cite{10.1117/12.926827,2022ApJ...934..171H}.
CAT gratings have a very high aspect ratio, where the grating bars are about 50 times higher than they are wide. Arcus will feature CAT grating with a period of 200~nm in the dispersion direction, 140~nm gaps between the bars and a depth of 5.6~$\mu$m. The gratings are manufactured with sizes up to $32\times32$ mm$^2$. Individual grating bars are held in place by a perpendicular support structure of wider Si bars (L1 support) and a hexagonal frame (L2 support) that the gratings bars and L1 supports rest on. The L2 mesh blocks $\sim 19$\% of the area, and the recently fabricated L1 meshes block between 10 and 18\% -- see Fig.~\ref{fig:cell} for an image and a schematic.

\begin{figure} [ht]
    \begin{center}
    \begin{tabular}{c} %% tabular useful for creating an array of images
    \includegraphics[height=5cm]{unitcell+pic.jpg}
    \end{tabular}
    \end{center}
    \caption {\label{fig:cell}
    Left: Schematic depiction of the structural hierarchy of the CAT grating membrane (see text). Right: Photograph of a back-illuminated $32\times32$ mm$^2$ prototype CAT grating membrane, showing the hexagonal L2 mesh and optical diffraction from the L1 mesh.
    }
\end{figure}

CAT gratings are mounted blazed, which means that the grating bars are not perpendicular to incoming rays, but tilted by 1.8~degrees. Together with the high aspect ratio, this leads to absolute diffraction efficiency $> 30$\%  at 2.38~nm wavelength (sum over blazed orders), including absorption by L1 and L2 supports\cite{doi:10.1117/12.2314180}. In this setup, the photons are mostly diffracted into one direction, the so-called ``blaze peak'' at twice the blaze angle.

Six CAT gratings are bonded into a window for each SPO. Those windows are then arranged into a grating petal for each channel.

\subsection{Double-tilted Rowland Torus layout}
Each channel follows a Rowland torus layout\cite{Beuermann:78} where the gratings are mounted on the surface of the torus and the detectors on the opposite side. Detectors follow the curved Rowland circle, which optimizes the width in the dispersion direction to increase the spectral resolving power of the instrument at the cost of a slightly wider photon distribution in cross-dispersion direction. For each channel, the Rowland torus is tilted by 3.6175~deg, about twice the blaze angle\cite{10.1117/12.856482,10.1117/12.2273011,DTRS}. This number is set such that the distance between two channels matches the space required to mount the grating petals next to each other. At the same time, all channels can be imaged on the same set of detectors (Fig.~\ref{fig:geometry_camera}).
The different tilted tori are arranged in such a way, that they overlap in the Rowland circle, so that the optimal detector position for one channel is also the optimal detector position for all other channels.
Two cameras are positioned to capture the zeroth order on one side and the blaze peak on the other with a gap in between where so little signal is found that it is not useful to capture it with CCDs.

\begin{figure} [ht]
    \begin{center}
    \begin{tabular}{c} %% tabular useful for creating an array of images
    \includegraphics[width=\textwidth]{geometry_camera}
    \end{tabular}
    \end{center}
    \caption {\label{fig:geometry_camera}
    Layout of the CCDs (orange rectangles) in the focal plane. Given the resolution of the figure, most chip gaps are not visible. Also note, that the x and the y axes are scaled differently (Figure~\ref{fig:4} shows correct aspect of the CCDS, but not the correct gap between the two detector arrays). Two cameras with eight CCDs each are located in the focal plane. In this plot, 0 is the geometric center of all the Rowland tori involved. Note that the cameras are intentionally not symmetric to 0.
    The position of the four optical axes for the four channels is marked with a blue pentagon.  The dispersion ($x$ - green horizontal arrows) and the cross-dispersion ($y$ - blue vertical arrows) of the different spectra are parallel to each other and point in the same direction. Two channels disperse left-to-right, and two right-to-left.
    }
\end{figure}

The four channels are arranged in pairs. Two of them have the zeroth order on the left in Figure~\ref{fig:geometry_camera} and disperse to the right, two of them have the zeroth order on the right and disperse towards the left. The pair of channels dispersing in the same direction set the CAT gratings on very similar Rowland tori, which are offset from each other in cross-dispersion direction by 10~mm so that the signal is clearly separated on the CCDs. They are also offset from each other in dispersion direction by 5~mm, to ensure that the same wavelength seen in different channels falls onto different x coordinates on the CCDs and thus a wavelength that happens to be lost in a chip gap between two CCDs in one channel can be observed in the other channel.

\section{RAY-TRACING}
\label{sect:raytracing}
\subsection{Optical Design}
\label{sect:design}

Each of the four channels of the XRS has a separate zero-order position and disperses the light to a different location on the detector. This has two main advantages: there is no need to align all channels to the same zero-order position within the width of the point-spread function (PSF, less than 1.5 arcsec in dispersion direction), and any wavelength of light that falls in a chip gap for one channel will be observable in the others; see Figure~\ref{fig:geometry_camera}. At the same time, this layout allows us to use the same detectors for all four channels, which reduces cost and complexity. More details of the optical layout are given in our previous ray-tracing papers\cite{10.1117/12.2273011,10.1117/12.2312678,DTRS}.

\section{Ray-traces}
We study the performance and alignment requirements of Arcus with ray-tracing. We use the ray-tracing code MARXS\cite{marxs1.1,2017AJ....154..243G}, a Python code developed by us under an open source license. MARXS is available at \url{https://github.com/chandra-marx/marxs}; simulations in this article are done with a development version with git hash f4b68dce0957 to include Arcus specific code. MARXS has been used for several mission proposals and is backed by hundreds of units tests and verified against laboratory works and on-orbit Chandra observations where feasible. MARXS does not calculate material properties such as reflectivities or grating efficiencies; instead they are read in from tables that can be based on laboratory data or simulations calculated by other programs.

The SPO model in the ray-trace is simplified and does not take into account vignetting, so the effective area predictions are not valid of off-axis source; but it does consider the geometric opening area (incl.\ ribs) and angle-dependent reflectivity to predict the effective area.

For CAT gratings we use efficiencies that are predicted from numerical simulations and that match the efficiencies measured for CAT gratings in the laboratory. The numbers include the effect of cross-dispersion from L1 support structures, dispersion and blockage by the hexagonal L2 supports, and any coverage by grating frames.

The cameras are covered a contamination blocking filter of 30~nm Al deposited on 45~nm polyimide and held by a thicker mesh with 95\% transmission. The mesh is treated statistically in the ray-trace as a 5\% reduction in intensity. 40~nm Al are deposited directly on the detectors to block optical light. The transmission for Al and polyiamide is taken from Ref.~\citenum{CXRO} with XAFS for C edge from measurements of Chandra ACIS imaging filter\cite{10.1117/12.245111}.
For the quantum efficiency of the CCD, we follow the QE of the Suzaku XIS1 BI CCD assuming depletion depth 42 microns\cite{2007PASJ...59S..23K}.

%;below 0.18 keV this is from ACIS S1 BI QE provided by Herman Marshall in file s1_newqe.qdp"}

\section{Effective area and resolving power}
The main performance characteristics of Arcus are the effective area and the resolving power of the extracted spectra. We run simulations on a wavelength grid from 1.5 to 60 \AA{} in steps of 0.15~\AA{}. Each simulation is run with 100,000 rays. MARXS tracks the probability of survival for each ray; for example when a ray passes a filter that transmits 15\% of the photons at that energy, the probability of a ray passing that filter is multiplied by $0.15$. In the end, the sum of the probabilities of all photons hitting the detector divided by the total number of rays simulated for a given entrance aperture gives the effective area with much lower statistical noise than in codes where photons either pass or do not pass individual elements. The uncertainty on the Arcus performance is dominated by the uncertainty in the material properties such as SPO reflectivity or CAT grating efficiency. These systematic effects are difficult to quantify, but are probably of order 10-20\% or so in sum, while the statistical uncertainty for each run is below a few per cent.

\begin{figure}
    \centering
    \includegraphics[width=0.8\textwidth]{Aeff_JATIS}
    \caption {\label{fig:Aeff}
        Effective area for Arcus based on ray-trace simulations. The effective area of individual dispersion orders is shown. For most wavelength ranges, several dispersed orders contribute to the total effective area.
    }
\end{figure}

The effective area for Arcus is shown in Figure~\ref{fig:Aeff} and the spectral resolving power $R$ per order and averaged over all orders that contribute at a particular wavelength (weighted by the effective area of each order at that wavelength) in Figure~\ref{fig:R}. The figure distinguishes between the photons seen in dispersion and in order zero, i.e.\ in direct light, where the only energy resolution is provided by the CCD. The effective area in direct light is high for high energies, in particular including the 6.7~keV (1.85~\AA{}) iron line, which is a crucial diagnostic in many astrophysical sources. Towards longer wavelengths, around 10~\AA{}, light in order 1 and -1, makes up the bulk of the signal with $R$ of a few hundred. From about 12~\AA{} to 60~\AA{} most photons are dispersed into the blaze peak on the set of detectors opposite of the zeroth order and Arcus achieves an $R$ around 3500, almost independent of wavelength. The resolving power increases with increasing distance between the location of the zeroth other and the dispersed signal. For a given order, that means that $R$ increases with wavelength. However, for most wavelengths, multiple orders contribute to the signal. As the wavelength increases, the order on the far side of the blaze peak becomes weaker, and orders on the near side stronger, such that the average $R$ is approximately constant.

\begin{figure}
    \centering
    \includegraphics[width=0.8\textwidth]{R_JATIS}

    \caption {\label{fig:R}
    Resolving power $R$ for Arcus. Colored lines show $R$ per order, the gray area shows $R$ averaged over all diffracted orders contributing to that wavelength, where the average is calculated weighting each order by the effective area. For low orders, there are gaps in the colored lines. At low wavelength, those orders fall on the same camera as the zeroth order, e.g.\ for order -2 up to 20~\AA{}. In order -2 photons longer than 43~\AA{} are dispersed into the blaze peak and detected on the other camera.
    }
\end{figure}

\section{Alignment tolerances}
We use ray-traces to assess the effect of misalignments on the performance of Arcus. Science requirements put limits on the maximal allowable degradation of spectral resolving power and effective area, and engineering constraints determine how well e.g.\ individual SPOs can be aligned into petals, how well the petals can be aligned to the forward assembly, and how well the forward assembly can be aligned to the detector housing. In general, tighter tolerances require more work, time, and money. We thus need to understand how important each possible degree of freedom is to the total performance of the system to identify those where significant design and work needs to go into the alignment.

The simulations start from a perfectly aligned version of Arcus. Even this does not provide infinite resolving power, because the model includes pointing jitter, a limited PSF, some astigmatism inherent in the design, and finite sizes of CAT gratings and CCD detectors, which means that they deviate from the ideal Rowland geometry. A ray-trace is run with this design and spectral resolving power ($R$) and effective area ($A_\mathrm{eff}$) are calculated.

After running the baseline version, one element of Arcus is shifted in one degree of freedom, e.g.\ all CCDs are shifted by 1~mm in the dispersion direction. The ray-trace is repeated, again $R$ and $A_\mathrm{eff}$ are calculated, then all CCDs are shifted by 2~mm and so on. After testing out the parameter space in dispersion direction, the CCDs will be shifted in cross-dispersion direction. In this way, each element (for example the CCD array), will be misaligned by various amounts in one of 6 degrees of freedom (shift along $x$, $y$, $z$, and rotation around $x$, $y$, $z$).
In the Arcus coordinate system, the $z$-axis is parallel to the optical axis, the $x$-axis is the grating dispersion direction and the $y$-axis is the cross-dispersion direction. Unless noted, rotations are not done around the origin of the coordinate system, but around the center of an element (e.g.\ the center of a SPO petal).

In the first stage, only one degree of freedom is changed at a time, because it is not computationally feasible to explore the entire parameter space. From those results, we identify where the alignment is easily (e.g.\ just from simple machining tolerances) much better than the requirements. In a second step, we can then run ray-traces where all degrees of freedom are varied according to the misalignment budget and thus check if the assumptions going into combining the misalignments in different degrees of freedom hold or if non-linear interactions degrade $R$ and $A_\mathrm{eff}$ more than expected. In this case, we would have to revise the misalignment budget appropriately.

To keep the computational load reasonable, we simulate only one channel of Arcus. Since spectra from each channel will be extracted separately and there is symmetry between the channels, most results apply equally to all channels. We discuss in the text the few instances where the channel symmetry does not apply.
Each simulation consists of 200000 photons and the same source photons are used for each series of ray-traces, e.g.\ all global CCD misalignments use the same photons.

\begin{figure} [ht]
    \begin{center}
    \begin{tabular}{c} %% tabular useful for creating an array of images
    \includegraphics[width=0.7\textwidth]{CAT_global_JATIS}
    \end{tabular}
    \end{center}
    \caption {\label{fig:CAT_global}
    Effect of mislignment of the CAT grating petal for rotation around the center of the petal. (a-c): Resolving power for rotation around $x$, $y$, and $z$, respectively, (d-f): effective area for rotation around $x$, $y$, and $z$, respectively.
    }
\end{figure}
\begin{figure} [ht]
    \begin{center}
    \begin{tabular}{c} %% tabular useful for creating an array of images
    \includegraphics[width=0.7\textwidth]{CAT_individual_JATIS}
    \end{tabular}
    \end{center}
    \caption {\label{fig:CAT_individual}
    Effect of mislignment of individual CAT gratings. The mislignment for each grating is drawn from a normal distribution, the $\sigma$ of that distribution varied from 0 to 3~degrees. (a-c): Resolving power for rotation around $x$, $y$, and $z$, respectively, (d-f): effective area for rotation around $x$, $y$, and $z$, respectively. Note that the $A_\mathrm{eff}$ reported is only for a single channel.
    }
\end{figure}
\begin{figure} [ht]
    \begin{center}
    \begin{tabular}{c} %% tabular useful for creating an array of images
    \includegraphics[width=0.7\textwidth]{detector_global_JATIS}
    \end{tabular}
    \end{center}
    \caption {\label{fig:detector_global}
    Effect of mislignment the cameras with respect to the front assembly. (a-c): Resolving power for translation parallel to $x$, $y$, and $z$, respectively, (d-f): effective area for translation parallel to $x$, $y$, and $z$, respectively.
    }
\end{figure}

Figures~\ref{fig:CAT_global}, \ref{fig:CAT_individual}, and \ref{fig:detector_global} show some examples of the results. In the figures, $A_\mathrm{eff}$ is given summed over all dispersed orders that fall on a CCD (bottom row) and $R$ is the average resolving power, where the resolving power from individual orders is averaged weighted by the number of photons in that particular order (top row). Thus, it is possible, that $R$ in the plot increases with increasing misalignment if $A_\mathrm{eff}$ drops at the same time. This happens when an order with lower-than-average $R$ drops off the CCD (thus reducing the summed $A_\mathrm{eff}$ and increasing the average $R$). There is no scientific benefit from the apparently increased $R$ here - if one required a higher resolving power, the lower orders can be ignored in scientific analysis even if they fall on the CCD.

Figure~\ref{fig:CAT_global} shows the effect of a change in the CAT grating petal position, while the SPO petal and the cameras are fixed. Rotations around either $x$ or $y$ mean that the CAT gratings on one side move up, while the other side moves down, changing the path length of the diffracted photons. Those photons coming from the high CAT gratings travel further along the dispersion direction than those from the low gratings, thus causing the dispersed spot to smear out, which reduces $R$. These rotation need to be kept at the level of a few arcminutes. Rotation around $z$ changes the direction of the dispersed light and, for large angles, the dispersed orders miss the CCD.

Figure~\ref{fig:CAT_individual} examines the rotations of individual gratings. For each grating in the CAT petal, a different misalignment is drawn from a normal distribution. Rotations around $x$ (the dispersion direction, see panels a and d in the figure) have little effect on $R$. Since the facets have finite size, their edges differ from the Rowland torus by a little already, and adding a little more rotation does not change much. Rotations in $y$ quickly reduce $A_\mathrm{eff}$ though, because the CCDs are placed for a certain blaze peak. Rotating the CAT gratings shifts the blaze peak and photons miss the CCDs. Rotation around $z$ causes more and more signal to miss the detector. There is no sharp cut-off, because the CAT gratings in the simulation have a distribution of rotation angles, and the larger the Gaussian $\sigma$ is, the more CAT gratings will send their dispersed photons to positions where they cannot be detected. From this figure, we can determine that the alignment tolerance for rotation around $x$ can be large, while the other two directions are of order a few arcminutes to prevent significant degradation of the Arcus performance.

Figure~\ref{fig:detector_global} shows simulations for translating the detector with respect to the forward assembly (SPOs and CAT gratings). The most important degree of freedom is a change in focus (panel c). A shift along $y$ has no impact, as long as it is small enough to keep the dispersed spectrum on the CCDs. For the particular channel simulated here the spectrum drops off the CCDs for $y$ shifts for about -15~mm on one side and about +5~mm on the other side. A shift of 5~mm or more will drop at least one channel of the detector.

The curves for changing $R$ with shift in $x$ show some up and down when an order hits a chip gap (panel a and d). For example, two orders contribute to the signal in the curve for 37~\AA{} photons. At +2 mm one of them hits a chip gap, causing a drop in $A_\mathrm{eff}$ and also in $R$ ($R$ is averaged over all contributing orders, but only one order, which happens to have a lower $R$, is detected at this position). Note that chip gaps are inevitable. There will always be some wavelengths in a chip gap; however, the overall performance of the instrument, which is averaged over a range of wavelengths, is unaffected by this. While, in principle, a shift along $x$ does matter in principle since the focal plane is curved, we find that shifts up to a few mm have little impact, as if the CCDs move in the dispersion direction, the spectrum will be only slightly out of focus. As described above, the Arcus design takes advantage of this to mitigate the effect of chip gaps by offsetting the different channels by a few mm in dispersion direction so that no two spectra have chip gaps at the same wavelength.

We run simulations for rotations and translations for each possible mechanical misalignment in Arcus, as well as for a few other parameters such as the SPO mirror PSF, the repeatability of the CAT grating period, the surface flatness of the CAT gratings etc, but it is beyond the scope of this article to show each result in a figure. Inspecting all results, and considering how easy or hard it is to improve alignments in each degree of freedom, we set the alignment tolerances listed in Table~\ref{tab:tolerances}.

\begin{table}
    \caption{Arcus alignment tolerances ($1\sigma$)\label{tab:tolerances}}
    \centering
    \begin{tabular}{l|cccccc}
        \hline\hline
    alignment & trans x & trans y & trans z & rot x & rot y & rot z \\
     & $\mathrm{mm}$ & $\mathrm{mm}$ & $\mathrm{mm}$ & $\mathrm{{}^{\prime\prime}}$ & $\mathrm{{}^{\prime\prime}}$ & $\mathrm{{}^{\prime\prime}}$ \\
     \hline
    individual SPO in petal & 0.007 & 0.033 & 0.167 & 100 & 100 & 3 \\
    CAT petal to SPO petal & 0.333 & 0.333 & 0.333 & 100 & 100 & 200 \\
    CAT windows to CAT petal & 0.333 & 0.333 & 0.067 & 100 & 60 & 100 \\
    individual CAT to window & 0.333 & 0.333 & 0.067 & 100 & 60 & 100 \\
    Camera to front assembly & 1.667 & 0.667 & 0.333 & 60 & 60 & 60 \\
    \end{tabular}
\end{table}

The alignment budget assumes that all alignment tolerances contribute independently. The next set of simulations is designed to check this assumption. Some misalignments might cancel out in practice, others might have a multiplicative effect. Full ray-tracing is the best way to check that and to predict final instrument performance. We run a set of 100 ray-traces, and draw a new set of mislaignments from Table~\ref{tab:tolerances} for each of them. Figure~\ref{fig:alignment_budget} shows the cumulative distribution of $R$ and effective area for those simulations relative to a perfectly aligned Arcus.

\begin{figure}
    \centering
    \includegraphics[width=0.75\textwidth]{alignbudget_base}
    \caption {\label{fig:alignment_budget}
    Cumulative distribution function for $R$ (panel a) and effective area (panel b) relative to a perfectly aligned Arcus for a set of simulations where all elements are simultaneously misaligned according to the misalignment budget in Table~\ref{tab:tolerances}.}
\end{figure}

Differences are visible between the different wavelengths. At 37~\AA{} two dispersed orders (order -3 and -4) contribute, but order -3 is close to the edge of a CCD and order -4 is close to a chip gap. If either one is lost, $R$ or $A_\mathrm{eff}$ at that wavelength will suffer, but at the same time a neighboring wavelength will improve because it is no longer inside a chip gap. In contrast, the signal at 25~\AA{} is dominated by order -6, comfortably in the middle of a CCD. Thus, the two plots above should not be interpreted as ``longer wavelength will suffer more''; crucial orders are close to a chip gap in different spots over the Arcus bandpass. Instead, the plots should be read as showing the range of effects that the baseline misalignment can have on $R$ and $A_\mathrm{eff}$, depending on what exactly the random numbers are that are drawn.

Apart from the effects of shifting chip gaps, Figure~\ref{fig:alignment_budget} shows that the alignment tolerances listed in Table~\ref{tab:tolerances} give us 95\% of the nominal $R$ and effective area in more than 90\% of the realizations.

\section{Future Work}
At this point, the main limitation of this work is our simplified mirror model. We expect it to work well for on-axis sources, but it does not reproduce the vignetting which becomes significant at about 2~arcmin. Our simplified mirror model also leads to a flat focal plane, while a Wolter-Type II optic typically has a slightly curved focal surface. Detailed ray-trace models specific to the Athena SPOs are available\cite{10.1117/12.2232230,10.1117/12.2594461,10.1117/12.2628133} and can be coupled with our code for CAT gratings and detectors.

\section{Summary}
\label{sect:summary}
Arcus is a Probe-class mission concept with two high-resolution spectrographs in the UV and in X-rays. The X-ray instrument (XRS) is designed in a double-tilted Rowland torus geometry with four channels with independent mirror and grating petals. All four channels are imaged onto the same set of detectors. We discuss the ray-racing model we set up for Arcus and describe the input data for mirrors, gratings, filters, and CCDs. We predict an effective area of $>200$~cm$^2$ from 8-50~\AA{} with a peak of 550~cm$^2$ around 15~\AA{}. The resolving power $R$ is largely flat over the bandpass with $R>3500$. Arcus will also detect higher-energy X-rays up to 10~keV in low diffraction orders (-1, 0, and +1) at the same time.

We run simulations to determine the alignment tolerances for Arcus. In many degrees of freedom, those are of order 0.5~mm for translations or 1-2~arcmin for rotations, which is easily achievable by mechanical means alone. Notably, the alignment requirements of individual SPOs in a mirror petals  and the translation of the CAT windows and petals along the optical axis are tighter and require dedicated alignment procedures.

% \disclosures
\subsection*{Disclosures}
The authors of this paper are members of the Arcus collaboration; R.~Smith is the PI of Arcus, A.~Ptak is the Co-PI, all other authors are Co-Is. Should NASA select Arcus for implementation, their institutions will receive funding which may be used to fund the author's salaries in full or in part in the future.

\subsection* {Code, Data, and Materials Availability}
We use the ray-tracing code MARXS\cite{marxs1.1,2017AJ....154..243G}, a Python code developed by us under an open source license. MARXS is available at \url{https://github.com/chandra-marx/marxs}; simulations in this article are done with a development version with git hash f4b68dce0957. The database of efficiencies for gratings is considered a trade secret by the Space Nanotechnology Laboratory and cannot be shared in open repositories - contact co-author Ralf Heilmann for access.

%In support of open scientific exchange, SPIE journals require Code, Data, and Materials Availability Statements in all accepted papers. This requirement went into effect on 1 May 2023. These statements should describe how to access any data that would be required to replicate or interpret the findings reported in the paper. Authors are encouraged to make the data and code related to the manuscript publicly available whenever possible, and utilize repositories that are well-known to the field (FigShare, Github, CodeOcean, etc.). If the data or code cannot be made publicly available, the authors should state the reason and explain how it can be obtained. Likewise, if data sharing is not applicable, the statement must say so. Example statements may be found in the Author Guidelines for the journal.

\subsection* {Acknowledgments}
Support
for this work was provided in part through NASA grant NNX17AG43G and
Smithsonian Astrophysical Observatory (SAO) contract SV3-73016 to MIT
for support of the {\em Chandra} X-Ray Center (CXC), which is operated
by SAO for and on behalf of NASA under contract NAS8-03060.  The
simulations make use of Astropy, a community-developed core Python
package for Astronomy\cite{astropy:2022}, numpy\cite{numpy}, and IPython\cite{IPython}.

%%%%% References %%%%%

\bibliography{report_with_doi}   % bibliography data in report.bib
\bibliographystyle{spiejour}   % makes bibtex use spiejour.bst

%%%%% Biographies of authors %%%%%
\subsection*{Biographies}
\vspace{2ex}\noindent\textbf{Hans Moritz G\"unther} is a research scientist at MIT. He received his undergraduate degree (in 2005) and his PhD (in 2009) in physics from the University of Hamburg, Germany. After that, he worked at the Harvard-Smithsonian Center for Astrophysics and came to MIT in 2015. He is currently the lead developer of MARX, the ray-tracing software used for the Chandra X-ray observatory. His science interests are in star formation using data from the radio to X-rays.

\vspace{1ex}
\vspace{2ex}\noindent\textbf{Peter Cheimets}
Please send me your biography (approx 75 words).

\vspace{1ex}
\vspace{2ex}\noindent\textbf{Casey DeRoo}
Please send me your biography (approx 75 words).

\vspace{1ex}
\vspace{2ex}\noindent\textbf{Ralf K.~Heilmann} is a principal research scientist at the MIT Kavli Institute for Astrophysics and Space Research, and the associate director of the Space Nanotechnology Laboratory (SNL).  He received his Diplom in physics from the FAU Erlangen/N\"urnberg (1991), and his MS (1993) and PhD (1996) in physics from Carnegie Mellon University.  After a postdoc at Harvard he joined the SNL and has since focused on advanced lithography and the development of x-ray optics for astronomy.

\vspace{1ex}
\vspace{2ex}\noindent\textbf{Eric D. Miller}
Please send me your biography (approx 75 words).

\vspace{1ex}
\vspace{2ex}\noindent\textbf{Andrew Ptak}
Please send me your biography (approx 75 words).

\vspace{1ex}
\vspace{2ex}\noindent\textbf{Randall K. Smith}
Please send me your biography (approx 75 words).

%\noindent Biographies and photographs of the other authors are not available.

%\subsubsection{Reference linking and DOIs}
%A Digital Object Identifier (DOI) is a unique alphanumeric string assigned to a digital object, such as a journal article or a book chapter, that provides a persistent link to its location on the internet. The use of DOIs allows readers to easily access cited articles. Authors should include the DOI at the end of each reference in brackets if a DOI is available. See examples at the end of this manuscript. A free DOI lookup service is available from CrossRef at \\\linkable{http://www.crossref.org/freeTextQuery/}. The inclusion of DOIs will facilitate reference linking and is highly recommended.

%In the present LaTeX template, the author needs to add the DOI reference by including it in a ``note'' in the bibliography file, as shown in the file {\verb+report.bib+}, for example, \\ {\verb+note = "[doi:10.1117/12.154577]"+}. The DOI may be used by the reader to locate that document with the link: {\verb+http://dx.doi.org10.1117/12.154577+}.

\listoffigures
\listoftables

\end{spacing}
\end{document}

### 2024-04-01 papers/special_JATIS_tolerances/doi_into_bib.py

@@ -1,0 +1,13 @@
+import bibtexparser
+# version < 2.0.0
+
+
+with open("report.bib") as bibtex_file:
+    bib_database = bibtexparser.load(bibtex_file)
+
+for entry in bib_database.entries:
+    if "doi" in entry:
+        entry["note"] = f'[{entry["doi"]}]'
+
+with open("report_with_doi.bib", "w") as bibtex_file:
+    bibtexparser.dump(bib_database, bibtex_file)

### 2024-05-20 papers/special_JATIS_ordersort/ordersort.tex
submitted paper to JATIS

\documentclass[12pt]{spieman}  % 12pt font required by SPIE;
%\documentclass[a4paper,12pt]{spieman}  % use this instead for A4 paper
\usepackage{amsmath,amsfonts,amssymb}
\usepackage{graphicx}
\usepackage{setspace}
\usepackage{tocloft}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage[table]{xcolor}
\include{moritz}

\usepackage{lineno}
\linenumbers

\title{Order sorting and background handling for Arcus}

\author[a]{Hans Moritz G\"unther$^*$}
%\author[b]{Peter Cheimets}
%\author[c]{Casey T. DeRoo}
\author[a]{David Huenemoerder}
%\author[a,d]{Ralf K. Heilmann}
\author[a]{Eric D. Miller}
%\author[e]{Andrew Ptak}
%\author[b]{Randall K. Smith}

\affil[a]{MIT Kavli Institute for Astrophysics and Space Research, Cambridge, MA 02139, USA}
%\affil[b]{Center for Astrophysics, Harvard-Smithsonian Astrophysical Observatory, Cambridge, MA 02138, USA}
%\affil[c]{Dept. of Physics \& Astronomy, University of Iowa, Iowa City, IA 52242, USA}
%\affil[d]{Space Nanotechnology Laboratory, MIT Kavli Institute for Astrophysics and Space Research, Cambridge, MA 02139, USA}
%\affil[e]{NASA Goddard Space Flight Center, Greenbelt, MD 20771, USA}

\renewcommand{\cftdotsep}{\cftnodots}
\cftpagenumbersoff{figure}
\cftpagenumbersoff{table}
\begin{document}
\maketitle

\begin{abstract}
The Arcus X-ray spectrograph (XRS) will deliver high-resolution spectra in four telescope/grating modules (called ``channels'') imaged onto the same set of detectors. We examine two effects (cross-dispersion from support structures in the diffraction gratings and frame-transfer) that can lead to a photon being assigned to the different channel in the data reduction. Even within a channel, photons can be assigned the wrong diffraction order because orders are closely spaced in energy. Based on simulations, we show that both effects are small in the XRS and can be dealt with in data analysis and fitting using existing data formats and algorithms. Finally, we describe a process to construct per-order background spectra for XRS simulations.
\end{abstract}

% Include a list of up to six keywords after the abstract
\keywords{Arcus, ray-tracing, X-ray, background, data analysis}

% Include email contact information for corresponding author
{\noindent \footnotesize\textbf{*}Hans Moritz G\"unther,  \linkable{hgunther@mit.edu} }

% FOR SUBMISSION
\begin{spacing}{2}   % use double spacing for rest of manuscript

% For my own while writing. It's just easier to read this way
%\begin{spacing}{1}

\section{Introduction}
\label{sect:intro}  % \label{} allows reference to this section
In the eye of the public, astrophysics is probably most famous for the beautiful pictures of planetary disks, star forming regions, and galaxies in the deep fields. However, in many cases, spectra are required to derive physical parameters such as densities, temperatures, gravitational redshift, or photon ionization fields. Arcus is a Probe class mission concept that combines two instruments, the X-ray spectrometer (XRS) and the UV spectrometer (UVS), to provide high-resolution spectra over the X-ray and UV band. This article describes certain aspects of the XRS data analysis and background handling. The XRS works in the 20-50~\AA{} band and delivers spectra with a resolving power around 3500 \textbf{Reference to Guenther et al. paper ``Arcus X-rat telescope performance predictions ...'' in the same JATIS issue here.}

Arcus addresses three major science themes but will also address a wide range of other questions in astrophysics through observations proposed by the community. Among the major science themes is detecting the missing baryons in the warm-hot intergalactic medium. Observationally, Arcus will address this by observing many sight lines to strong, bright background objects such as quasars. Those sources emit an X-ray continuum spectrum and if the line of sight passes through clouds of hot, ionized gas, Arcus will see weak absorption lines of O~{\sc vii} and other highly ionized ions superimposed on that continuum. In same cases, a sightline may contain several absorbing clouds at different redshifts\cite{10.1117/12.2231193} and the spectrum will thus have weak absorption lines that are hard to find within the noise of a continuum when the number and location of those lines is unknown a priori.

Other major themes for Arcus are studies of stellar coronae and stellar accretion. For those targets, Arcus XRS will easily resolve the emission lines. For example, in young stars that are still surrounded by an accretion disk, the co-called classical T Tauri stars (CTTS), gas is ionized by the stellar radiation at the inner edge of the disk and then falls in towards  the star following the magnetic field lines. The plasma is accelerated to free-fall velocities around 500~km~s$^{-1}$ and passes through a strong shock on the stellar surface. In the shock, material is heated up to X-ray emitting temperatures, and the Arcus XRS will observe emission lines from ions such as O~{\sc vii}, O~{\sc viii}, Ne~{\sc ix}, and Ne~{\sc x}\cite{2022hxga.book...57S}. Those lines may be shifted by a few hundred km~s$^{-1}$ and will be kinematically resolved, but in general identifying the strong emission lines will be relatively easy. Those two scenarios, bright, known emission lines and weak absorption lines at unknown position represent two extremes in the range of spectral fitting. If both can be successfully addressed, spectral modelling for any other scenario can also be done.

In this article, we analyze certain aspects of the data analysis for Arcus X-ray spectra.
We first summarize the design of the XRS in sect.~\ref{sect:layout}. Arcus consists of four independent telescope/grating modules (which we call ``channels'' from now on) that are imaged onto a single focal plane.
Given the layout of the XRS, spectral extraction might seem daunting at first, due to cross-talk between different channels (section~\ref{sect:crosstalk}) and contamination by neighboring orders (section~\ref{sect:ordersorting}).

We show in section~\ref{sect:dataanalysis} that established techniques and data formats used for the spectra from grating spectroscopy in Chandra and XMM-Newton are already sufficient to achieve Arcus science goals though further improvements in data analysis are possible. We also look at the particle background in the detector (section~\ref{sect:nxb}).

The discussion here is based on ray-traces with the ray-tracing code MARXS\cite{2017AJ....154..243G,marxs2.0}, a Python code developed by us under an open source license. MARXS is available at \url{https://github.com/chandra-marx/marxs}; simulations in this article are done with version 2.0.

\section{Layout of the X-ray Spectrometer on Arcus}
\label{sect:layout}
The Arcus XRS is designed as a double-tilted Rowland Spectrograph (DTRS). This concept is described in detail in Ref.~\citenum{DTRS} and summarized in this section. Every X-ray grating spectrometer consists of three basic elements: A focussing mirror, a set of gratings, and a focal plane for the dispersed photons, usually also covering the direct light that passes through the gratings without getting dispersed. We summarize the layout of the XRS here and refer to our previous ray-tracing papers for details\cite{10.1117/12.2273011,10.1117/12.2312678,DTRS}.

For the XRS, the focussing mirror is made from Silicon Pore Optics (SPOs), which are originally developed for the Athena mission\cite{10.1117/12.2188988,10.1117/12.2599339,10.1117/12.2677388}. The XRS will ``sub-aperture'', which means that the SPOs do not cover a full circle, but only a small wedge about 30~deg wide. While the point-spread function (PSF) of a mirror covering the full circle would be rotationally symmetric, the sub-aperturing will lead to an elongated PSF that is relatively narrow (about 1.5~arcsec) in the dispersion direction and wider in the cross-dispersion direction\cite{1987ApOpt..26.2915C,2010SPIE.7732E..1JH,DTRS}, hence yielding higher spectral resolution than the full aperture.

Beyond the mirror, the photons encounter a set of Critical Angle Transmission (CAT) gratings\cite{10.1117/12.926827,2022ApJ...934..171H}. CAT gratings have a high aspect ratio; in the case of Arcus the grating bars are 5.6~$\mu$m high, but only 60~nm wide. CAT gratings are mounted blazed, such that the incoming photons make a small angle (1.8~deg on average for Arcus) to the grating bar side walls. Conceptually, one can think of this as individual photons being reflected of the grating bar side walls and thus CAT gratings disperse most photons to about twice the blaze angle on just one side. Looking at the grating equation for a grating with a period $d$ (200~nm for the XRS) where $n$ is the dispersion order, $\lambda$ the wavelength of the photon, and $\theta$ the dispersion angle
\begin{equation}
    n \lambda = d \sin{\theta} \label{eqn:grating}
\end{equation}
we see that photons with different $\lambda$ can be dispersed by the same angle, if they are in different dispersion orders. For example, $\lambda=6$~nm photons in dispersion order $n=2$, $\lambda=4$~nm photons with $n=3$, and $\lambda=3$~nm photons with $n=4$ will all be dispersed by the same angle $\theta$ and thus land on the same location of the detector. Since the efficiency of CAT gratings is such that photons are preferably dispersed such that $\theta$ is close to twice the blaze angle, the XRS will detect several dispersion orders with different wavelength at the same location on the detector -- a challenge for the data analysis that we discuss in sect.~\ref{sect:ordersorting}. For higher energies, the CAT gratings will act as phase-shifting gratings, which means that those photons get dispersed into lower orders, but on both sides (typically -2, -1, +1, +2).

\begin{figure}
    \centering
    \includegraphics{EQFullDet}
    \caption {\label{fig:fulldet}
    Simulated observation of an active star with the XRS. Photons are only shown if detected in one of the 16 CCDs. Note that the dispersion and cross-dispersion axes are scaled differently to highlight the offset between channels in cross-dispersion direction. There are four zeroth orders and four dispersed spectra. Two spectra are dispersed from left to right and two from right to left. Spectra also have small (5~mm) offsets in the dispersion direction to avoid the same wavelength falling into a chip gap for multiple channels.
}
\end{figure}

The XRS has four modules (``channels''), each with its own petal of SPOs aligned to its own optical axis and its own set of CAT gratings. However, to reduce cost and power needs, all four channels are arranged to disperse the light on a single focal plane consisting of 2 cameras with 8 CCDs each. The four spectral traces are intentionally offset from each other in the cross-dispersion direction so that they can be separated by the data extraction algorithms (Fig.~\ref{fig:fulldet}).

\section{Cross-talk between channels}
\label{sect:crosstalk}

\subsection{Cross-dispersion}
Despite the spatial separation, a small fraction of photons from one channel will be detected as photons of another channel. Figure~\ref{fig:onechannel} shows just the photons from a single channel for the same simulation as fig.~\ref{fig:fulldet}.
The four dispersed spectra are relatively tight and only a few mm wide. We can define a spectral extraction region for each channel (marked in red in the figure). The spectra appear to be slightly curved, but this only a consequence of subaperturing; without subaperting the spectra would be wider in cross-dispersion direction (as well having a wider line-spread function (LSF) in dispersion direction) and be symmetric. The dispersion direction is not curved, however, given this design, we can make use of the shape we see and define a spatial extraction region that is narrow and also slightly curved for each channel; this way the extraction region is smaller, and thus the contamination and background is lower than for a flat, rectangular extraction region. In figure~\ref{fig:onechannel} four extraction regions for the four channels are shown in red. Each of them covers about 80\% of the photons in that channel.

\begin{figure}
    \centering
    \includegraphics{EQFullDet_onechannel}
    \caption {\label{fig:onechannel}
    This shows the same simulation as in Fig.~\ref{fig:fulldet} but only shows photons from a single channel (in real observations, we will not always know in which channel a photon originated but in the simulation we can track that). The red area marks a spectral extraction region that contains about 80\% of the photons in this channel. In several locations, cross-dispersed photons are visible, in particular at the zeroth order (around +380~mm), as bands reaching up and down to the right of the zeroth order, and about parallel to the main spectrum around the blaze peak (-350~mm).
}
\end{figure}

The figure uses a color scale that highlights faint features and only photons from one channel are shown. The photons fill extraction region for their channel (the top red region) but there are also locations where some photons leak into the extraction regions of other channels. This happens because of cross-dispersion. In the CAT gratings, the primary grating bars are held in place by Si strips with a wider spacing that run perpendicular to the main grating bars (Level 1 or ``L1'' support structure). Those L1 support bars also act as a diffraction grating dispersing perpendicular to the main dispersion direction (up or down in the image). However, since their spacing is much wider and the grating is not blazed in this direction, the efficiency of the L1 bars is only a few percent.
Cross-dispersion from the zeroth order where photons of all wavelengths are found at the same detector location affects the other channels the most. Close to the zeroth order, some $n=1$ photons are also cross-dispersed into cross-dispersion orders (mostly $m =-3, -2, -1, +1, +2, +3$). The distance of cross-dispersion from the main spectrum depends on the wavelengths of the photons according to the grating equation $m\lambda = d_{\mathrm{L1}}\sin\gamma$, where $m$ is the cross-dispersion order, $\gamma$ the angle, and $d_{\mathrm{L1}}=5\;\mu\mathrm{m}$. Therefore, the cross-dispersion orders run diagonal from the zeroth order around the dispersion coordinate 300~mm. We also see cross-dispersion at higher orders around the blaze peak (about -300~mm) where the cross-dispersion orders run almost parallel to the main spectrum because of the larger distance to the zeroth order.

\subsection{Read-out streaks}
Since Arcus uses CCDs as detectors, there is another mechanism that also re-distributes photons from their original channel to a position on the detector that will cause them to be assigned to a different channel in the data reduction: Photons arriving during an exposure cause a charge cloud centered on some pixel of the CCD, e.g.\ $(x,y) = (234, 516)$. During the read-out, charge is transferred line-by-line along the $y$ direction of the CCD (``parallel transfer''). If another photon arrives during the read-out at the same time that line 516 is transferred from, e.g., line 123 to line 122 and that new photon hits at line 122, then it will be read out together with the signal from line 516. Thus, it will be assigned $y=516$, not $y=122$. The faster the read-out, the lower the probability that photons arrive during the frame transfer, but the speed of the read-out is limited by the electronics and the acceptable read noise.

While we do not know the $y$ location of any individual photon in this case, we do know the ratio of integration time to frame transfer time and can calculate the fraction of photons that will be assigned a wrong value of $y$ and that will end up in the wrong spectral extraction region.

\subsection{Accounting for channel cross-talk}
If the dispersion coordinate was the same for all channels, moving a few photons from one channel to another channel would not be a problem. However, in the XRS, two channels disperse left-to-right and two disperse right-to-left and even those channels that are dispersing in the same direction are offset in the dispersion coordinate. Thus, photons will be assigned the wrong dispersion coordinate and thus the wrong wavelength if they are assigned to the wrong channel. As an example, for an emission line spectrum with a strong Ne~{\sc x} line at 1.214~nm, some of those photons leaking into a different channel via readout streaks might appear as an emission line, that, in the dispersion coordinate of the other channel, seems to be at 1.15~nm. The situation is even more difficult for continuum sources where the effect might be less obvious since it does not cause any isolated features, but just increases the apparent level of the continuum, making it harder to find weak absorption lines.

Fortunately, both effects are small for the XRS because the out-of-frame time factions is $<1$\%. The cross-dispersion efficiencies are only a few percent of the dispersion efficiency. Cross-dispersed photons and photons detected during the frame transfer time are both distributed over large regions of the detector and only a very small fraction of them falls into the spatial extraction region of another channel (figure~\ref{fig:onechannel}). Nevertheless, we can measure the efficiency of cross-dispersion and ray-trace the location on which the photons fall and calculate the probability of photons arriving during the frame transfer time. This can than be accounted for in spectral modelling. To follow the example from above, the response matrix file (RMF), that describes in which spectral channel (here: dispersion coordinate) photons of a given energy will be detected, would have a strong peak at 1.214~nm for photons with a true wavelength of 1.214~nm, but also a smaller peak at 1.15~nm. Thus, a spectral model that predicts the line at 1.214~nm will also explain the apparent (misplaced) line at 1.15~nm.

\section{Order sorting}
\label{sect:ordersorting}
Given the $(x, y)$ position on the detector for each detected photon and the geometry of the instrument, we can calculate the angle $\theta$ for each photon. Since we also know $d$, it is often convenient for order-sorting to describe the position of a photon as $n\lambda$. Photons of 60, 30, 20, and 15~\AA{}, diffracted into order 1, 2, 3, and 4 respectively, fall on the same position on the detector, because they have the same value for $n\lambda$. To determine the wavelength of each photon, it must be assigned to an order $n$. This can be done using the intrinsic energy resolution of the CCD. One way to look at the order-sorting is to display the $n\lambda$ coordinate along the x-axis and the CCD energy along the y-axis. Because of their hyperbolic shape, these plots are colloquially known as ``banana plots''. Figure~\ref{fig:banana} shows an example.

\begin{figure}
    \centering
    \includegraphics{banana}
    \caption {\label{fig:banana}
    Lines in a "banana plot", showing the relation between $m\lambda$ and photon energy for orders -1 to -13. Every value of $n\lambda$ corresponds to a specific position on the detector. The black dots are individual photons from an Arcus simulation.
}
\end{figure}

It is easy to analytically calculate the difference in CCD energy between the different orders $n$ and $m$ for any given value of $n\lambda$ (which corresponds to a given location on the CCD): Using eqn.~\ref{eqn:grating} and the photon energy $E = h c / \lambda$, where $h$ is the Planck constant and $c$ the speed of light, we can compare the energies of two photons at the same $d \sin\theta$, but with different orders $m$ and $n$:

$$ d \sin\theta = \frac{m h c}{E_m} = \frac{n h c}{E_n} $$

We can now calculate the difference:

$$\Delta E = E_m - E_n = E_m - E_m \frac{n}{m} =  E_m (1-\frac{n}{m})$$

For a given detector position, the energy difference between two consecutive orders is constant. Re-writing the energy as $n\lambda$ again, we can see that $\Delta E$ is largest for small values of $n\lambda$, closer to the zeroth-order:

\begin{equation}
    \Delta E = \frac{hc (m-n)}{m\lambda_m} \label{eqn:deltae}
\end{equation}

The fact that the orders are further apart on one side of the banana plot than on the other one means that order sorting will work better on one side of the detector, while the orders overlap more on the other side.

In a real CCD the energy of each photon is imperfectly measured. For the Arcus CCDs, we predict a FWHM $<70$~eV for the soft X-ray range based on simulations and laboratory measurements of prototype devices\cite{Miller2023}. That means that not all photons that belong to a specific order land exactly on the line in the order-sorting plot. Instead, they are distributed in energy. The black dots in figure~\ref{fig:banana} are photons in an Arcus simulation. The input spectrum is a powerlaw with an exponent of -2. Because the Arcus CCDs are not sensitive to wavelengths above about 60~\AA{}, there are no photons in the plot for $n=-1$ and for $n=-2$ above $m\lambda=125$~\AA{} (our convention is to use negative order number for the blaze peak in CAT gratings). For all orders, the photons cluster around the colored lines, but because of the finite energy resolution of the CCDs, some fall in between orders. Fewer photons are seen in higher orders. This is due to a combination of decreasing efficiency for the CAT gratings for higher energies and the shape of the input spectrum.

How do we assign photons to a specific order, and thus, given $n\lambda$, to a specific wavelength? Every $n\lambda$ corresponds to a specific distance from the zeroth order. We can make a vertical cut through figure~\ref{fig:banana} for a specific value. Figure~\ref{fig:120} shows this for $|n\lambda|=120$~\AA{}. Instead of individual photons from a simulation, the figure shows as shaded area the probability distribution for photons to be detected at a certain CCD energy, given its true diffraction order and true energy. Note that in this plot, every order is shown on the y-axis at the same height to explain the concepts applied here. In practice, the height of each order depends both on the characteristics of Arcus, e.g.\ the quantum efficiency of the CCD changes with energy, and on the spectrum of the source, e.g.\ if the source does not emit any photons below 0.5~keV, then there will be no signal in order -1 to -5 and therefore also no contamination of the extracted order -6 by order -5 photons.

\begin{figure}
    \centering
    \includegraphics{120}
    \caption {\label{fig:120}
    This plot is a for a specific value of $m\lambda=120$~\AA{}. The vertical lines mark the exact energy that a photon of that order should have at $m\lambda=120$~\AA{}. The distributions show the probability density of detecting a photon of a specific order at a specific energy (from left to right, orders -4, -5, and -6), or in other words, they show the energy distribution that we will observe from a large number of photons in an order. Because these distributions are wider than the distance in energy between two consecutive orders, they overlap. Thus, a photon that arrives at this detector position and has a detected energy of 0.45 keV could be either an order -4 photon or an order -5 photon. See text in section~\ref{sect:ordersorting} for a discussion of the bars on top of the plot.
}
\end{figure}

The bars on the top of figure~\ref{fig:120} mark the energy regions that will be extracted for each order. The three sets of bars at $y=16$, $y=15$, and $y=14$ show different possible ways to define the order sorting regions. The top bars at $y=16$ are for a scenario where the extraction regions for the orders touch. This maximizes the number of photons extracted because each photon will be assigned to exactly one extracted spectrum. The cost for this is that an order -5 spectrum will have significant contributions from order -4 and -6, which can complicate spectral fitting.
On the other hand, a narrower region like the bottom of the three bars (at y=14) is for a scenario where only photons within 30 eV of the expected energy are extracted. One can see that this leads to a considerable loss of photons, because the integrated probability for a photon to land in the extracted regions is less than 60\%; in other words, this choice of extraction region reduces the effective area of Arcus by more than 40\%. However, it also reduces contributions from order -4 and -6.

Other prescriptions for the order-sorting boundaries are possible, for example one could choose to always extract e.g.\ 60\% of the photons. Since the distance between orders depends on $n\lambda$ that means that the width of the extraction region in energy space also depends on $n\lambda$.

So far, the discussion concentrated on the relative importance for each order. In practice, more factors contribute to the effective area. Next, we discuss the actual effective area calculated for Arcus for different order-sorting scenarios.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{osiparf}
    \caption {\label{fig:osiparf}
    (a): Effective area for order -6 (the high, central curve) and the two contaminating orders -5 and -7 (lower curves to the left and right). All three contribute to the total effective area in the extracted spectrum, but their contributions are shown here separately. The x-axis shows the true wavelength of the photons seen in the extracted order -6. Even photons $>30$~\AA{} contribute to the extracted -6 order spectrum. Three different scenarios for order sorting are shown.
    (b) The same plot as on the left, but the x-axis shows the wavelength that will be assigned to the detected photons, not their true wavelength. The central curve is the same, but photons of order -5 and -7 will be assigned the wrong wavelength, since they are erroneously identified as order -6 photons. Those 30~\AA{} photons discussed as example, will be detected around 25~\AA{}, making the identification of any feature more difficult.
    (c) Ratio of the contaminating orders to the main order.
}
\end{figure}

Figure~\ref{fig:osiparf} shows the effective area for order -6 and the contamination from order -5 and -7 for three different scenarios for order sorting. In contrast to the above, the curves shown here now take all contributions to the ARF into account, e.g.\ the chip gaps (which look like small rectangles ``hanging down'' from the curve), the CCD QE, and the filter transmission. Thus, the contribution of order -5 and -7 are no longer the same, unlike in the banana plots above.

All figures here show the effective area for different order-sorting scenarios, where the order sorting regions contain a different fraction of the photons. The scenario with touching extraction regions is called ``osiptouch'', where ``osip'' stands for ``order sorting and integrated probabilities''. In the ``osiptouch'' scenario, the integrated probability to find that a photon of a specific order within the region is maximized by stretching the regions for each order as far as possible without overlapping. This delivers the highest total effective area, but has also the highest contamination; ``osip60'' and ``osip80'' leave a gap in energy space between the extracted orders, such that 60\% or 80\% of all photons will be extracted. This leads to noticeably reduced effective area. However, the contaminating orders are more suppressed than the central order. While the total effective area in the ``osip60'' scenario is 40\% reduced compared to ``osiptouch'', the relative contribution of the contaminating orders drops by a factor of a few.

Fortunately, the energy resolution of the CCDs is good enough that the contamination is low in most cases, but it can be seen in the case of strong emission lines. Figure~\ref{fig:contam} shows a simulation of a continuum spectrum with a strong emission line at 0.5~keV (24.8~\AA{}) that, at its peak, is about two orders of magnitude stronger than the continuum. Indeed, the line is seen strongest at 24.8~\AA{}, but it also appears at $4/5 * 24.8=19.8$~\AA{} and $6/5 * 24.8=29.8$~\AA{} because photons from order -4 and -6 are erroneously allocated to order -5. Photons in order -4 are diffracted only $4/5$ as far as photons of the same energy in order -5, and thus they are found at 19.8~\AA{} in a spectrum that uses the wavelength scale for order -5.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{contam}
    \caption {\label{fig:contam}
    Arcus simulation for CCD order -5 for a model with a continuum and a strong emission line at 0.5~keV (24.8~\AA). The line can be seen three times. It is strongest at 24.8~\AA{}, but the contamination from neighboring orders is strong enough to also cause discernible signal aliased to 19.84 and 29.76~\AA{} (equal $m\lambda$).
}
\end{figure}

\section{Data analysis}
\label{sect:dataanalysis}
Arcus is not the first spectrograph with overlapping orders and multiple extracted spectra and modelling and fitting such spectra is a solved problem. The Chandra/LETGS\cite{10.1117/12.278846,10.1117/12.278845} uses the Chandra/HRC-S detector\cite{10.1117/12.283772}, which essentially has no intrinsic energy resolution, and thus all orders appear in the same extracted count spectrum. In all spectrographs on Chandra, more than one count spectrum is extracted and modeled simultaneously with the same model spectrum. In the LETGS there are two count spectra (positive and negative diffraction) and in the HETGS\cite{2005PASP..117.1144C}, there are five or more (zeroth order with CCD energy resolution, high-energy grating (HEG) positive and negative first order, for deep exposures higher orders can be extracted separately, and medium energy grating (MEG) positive and negative first order, for deep exposures higher orders can be extracted separately). Similarly, in XMM-Newton there are typically five spectra (CCD resolution from the EPN, MOS1, and MOS2 cameras and grating spectra from RGS1 and RGS2\cite{2001A&A...365L...7D}).

In X-ray data analysis, one typically starts with a model spectrum and convolves that with the instrument response to predict the number of observed counts. Model parameters are adjusted to achieve a good fit between the convolved model spectrum and the observed data. The instrument response in this case is the sum of the contributions from the different orders, and all common programs used to model X-ray spectra such as XSPEC\cite{1996ASPC..101...17A}, Sherpa\cite{10.1117/12.447161}, ISIS\cite{2000ASPC..216..591H}, and others implement this using standard OGIP data formats that have been defined over 30 years ago and are common to all X-ray astronomy missions worldwide\cite{1995ASPC...77..219C}.
Arcus data analysis will follow this heritage.

\subsection{Overlapping orders}
Since the CCDs in the XRS have an energy resolution, standard data products for Arcus will be generated per diffraction order. Products will also be split between the camera which contains the zeroth order and the camera that covers the blaze peak, just to avoid dealing with the large spatial gap between the two cameras in the same file, which would require discontinuous wavelength grids or many channels with zero effective area, vastly increasing the file size. Depending on the exposure time and the spectrum of the source, not all orders may contain usable signal in a particular observation but typically, one will extract spectra for order -2, -1, 0, +1, and +2 from the camera which covers the zeroth order (``near'' camera) and spectra from order -2 to -14 in the other camera (``far'' camera). In the ``near'' camera, orders are well separated (see eqn.~\ref{eqn:deltae}), but for the ``far'' camera, we have to consider contamination from neighboring orders in some cases, which means that each spectrum will be modeled using three effective area files (``ARF'') and three response matrix files (``RMF''). So, for each channel, there will be about 18 extracted spectra (``PHA'' files) and 44 ARF and RMF files. That is more than for the spectrographs on Chandra and XMM, but can be handled with today's fitting programs on a current laptop or desktop computer. Wrapper scripts can simplify reading in and setting up all those files at once.

\subsection{Cross-talk between channels}
Once the relative location the spectral traces of the four channels is calibrated in space, we can predict at which location photons from one channel will be seen in another channel. Both cross-dispersion and read-out streaks are deterministic and can be built into the ARF and RMF files such that they are automatically accounted for in spectral modelling.

There are two approaches to deal with the four independent channels. One approach is to add spectra for each order for all channels. That reduces the number of files needed for fitting, but also reduces the information content since the information which channel leaks into which other one is lost. For the most precise fitting, data for each channel should be loaded separately, but be fitted at the same time with the same model as it is done today in Chandra/HETG or XMM/RGS.

\subsection{Future developments}
As shown above, Arcus data can be packaged and fit entirely with existing data formats and modelling codes and this is sufficient to reach the science objectives. However, improved fitting methods can give better results, yet they do require new data formats and software approaches. In the OGIP scheme, every photon is assigned to exactly one order (or a background spectrum).
While for some photons there is a high degree of certainty to which order they belong (those photons detected very close to the lines in the banana plot infigure~\ref{fig:banana}), others fall between orders and might originate from either. In this case, one could assign a fractional photon to each order to account for the probability that it might belong to either order. However, this approach does not allow for fitting using Poisson likelihoods, since the number of detected photons in each order would no longer be an integer, so it can only be used in the case of very high signal, where the distribution of photons per bin can be approximated as a normal distribution.

Another approach is to fit in a multidimensional space. Instead of extracting a number of 1D spectra, one can perform the fit in the 2D space of the banana plot ($n\lambda$ vs. CCD energy) or even in the 3D space of detector coordinates and CCD energy ($x$, $y$, CCD energy). This would make use of the knowledge which photons are detected close to the expected position, and which photons are ambiguously seen between channels or orders and thus should in principle lead to a more precise determination of model parameters from the same number of observed photons.

An entirely different way to address this is to forgo deterministic, hand-designed calibration files altogether. Instead a ray-trace code is used to simulate PHA files, event lists, or detector images for many possible input spectra. A neural net or other machine learning algorithm is trained to deduce spectral parameters from the observed data. Essentially, the neural net learns to perform the deconvolution from examples and thus there is no need to understand how the detector response behaves in detail, as long as the training data covers the possible range of spectra and detector effects.
Tools like this are currently developed for CCD resolution spectra\cite{2021RNAAS...5..113R,2023arXiv231118014R}. However, it is not known at this time if a similar approach can work for high-resolution grating spectra. Given the much higher resolution, one might expect that a much larger number of training spectra are needed, which might be beyond the computational power available in the near future, but given the rapid development in the field of machine learning, it is hard to predict the power of those algorithms when Arcus launches. Another concern is that in high-resolution grating data, we typically will not just look at a few overall parameters of a spectral model such as temperature, abundance, or powerlaw exponents, but will analyze individual line profiles or line shifts. In many cases, we do not have physical models that predict these properties from first principles and thus we cannot generate artificial training data for a machine algorithm to learn on. Workarounds such as training on generically parameterized line profiles could be possible, but it is not clear if machine learning is actually advantageous to the more traditional forward folding approach in these cases.

\section{X-ray instrument particle background}
\label{sect:nxb}

Galactic cosmic ray (GCR) particles interact with the Arcus X-ray instrument camera structure, generating secondary X-rays and electrons that produce signals in the detectors that are indistinguishable from focused celestial X-rays. To first order, this background is uniform across the focal plane due to the isotropic GCR flux, and therefore it has a very different spatial-spectral pattern compared to the diffracted spectra produced by the optics and gratings.

To simulate the effect of this non-X-ray background (NXB) on per-order spectral analysis of Arcus X-ray observations, we produced a model spectrum using in-flight data from CCD detectors on Chandra (ACIS-S3) and Suzaku (XIS1). These heritage detectors share many features with the Arcus X-ray CCDs: identical pixel size (24 $\mu$m), backside illumination with similar depletion depth (45 $\mu$m for ACIS-S3 and XIS1 vs.\ 60 $\mu$m for Arcus), and similar operational and on-board processing parameters. The Chandra ACIS-S3 NXB spectrum was derived from ACIS stowed data taken in 2005--2009, during a time of intermediate Solar activity. The Suzaku XIS1 spectrum was derived from observations of the dark (night) Earth limb taken over the course of the full 2005--2015 mission. Both datasets were filtered in standards ways, with no Cut-Off Rigidity (COR) filtering applied for Suzaku\cite{Tawaetal2008}.

As expected from the detector similarities, the spectra in Figure~\ref{fig:nxb} (left panel) show very similar features: a general power-law slope; instrumental fluorescence lines of O K (0.5~keV), Al K (1.5~keV), Si K (1.7~keV), and Au M (2.3~keV); and upturns at low and high energies. The Chandra NXB is about a factor of four higher than Suzaku when scaled to physical detector area, expected due to the different particle environments in high- and low-Earth orbits. Arcus should have a very similar overall unrejected background event rate to Chandra ACIS, as it will also be outside the Earth's protective magnetic field. There are some important changes we implemented to model the Arcus NXB spectrum:

\begin{figure} [t]
    \begin{center}
    \begin{tabular}{c} %% tabular useful for creating an array of images
    \includegraphics[width=\textwidth]{Fig7}
    \end{tabular}
    \end{center}
    \caption {\label{fig:nxb}
    (panel a) Observed non-X-ray background (NXB) spectra for the Chandra ACIS-S3 and Suzaku XIS1 CCD detectors, shown with the adopted Arcus model spectrum. XIS1 has a lower background due to Suzaku's low-Earth orbit, compared to the high-Earth orbits of Chandra and Arcus. Other features are discussed in the text. (panel b) Example simulated per-order NXB spectrum, shown along with the number of CCDs and effective area (both with arbitrary scaling) as a function of wavelength. The broad feature peaking near 24 \AA\ in the NXB spectrum is due to the O K fluorescence feature seen near 0.5 keV in panel (a). This demonstrates the subtle effects that the NXB and detector energy response can impose on extracted per-order spectra.
    }
\end{figure}

\begin{enumerate}
    \item The upturn at high energy is removed. This upturn is due to cosmic ray minimum-ionizing particles (MIPs) that hit the detector at normal incidence and traverse the full detector thickness, confined to a single pixel. The distribution of energy each MIP deposits is governed by the detector thickness, but it is quite broad, extending to low energies. Since the Arcus CCD is somewhat thicker (60 $\mu$m) than ACIS-S3 (45 $\mu$m), this feature will move to higher energies above the Arcus band.
    \item The upturn at low energies from Suzaku is used instead of Chandra. It is likely that the gain scale at low energies for ACIS-S3 is non-linear due to high charge-transfer inefficiency (CTI), and we suspect events here suffer incorrect energy assignment. The Arcus CCD performance at low energies will be much closer to XIS1.
    \item The instrumental fluorescence lines will likely differ in amplitude for Arcus, but we have kept the Suzaku line strengths to be aware of their presence.
\end{enumerate}

In practice, we have simply fit an empirical model to the Suzaku spectrum, ignoring the high-energy upturn, and scaled that to match the ACIS-S3 rate at 1 keV.

To create per-order NXB spectra, the CCD NXB model was first converted to cts s$^{-1}$ keV$^{-1}$ \AA$^{-1}$ by multiplying by the spatial cross-dispersion width of the extraction window (50 pixels = 1.2 mm) and by the dispersion relation (in \AA\ mm$^{-1}$) from eqn~\ref{eqn:grating}.  At each response channel wavelength, we integrate the CCD NXB model over a window 70 eV wide, centered on that energy. This results in a spectral count rate in units of cts/s in bins equivalent to those in the per-order response. Finally, we multiply the rate by the number of CCDs contributing to the sensitivity at each wavelength. Example NXB spectra are shown in the right panel of Figure \ref{fig:nxb}; these can be used with per-order Arcus response files to simulate the effects of the NXB on the analysis of faint science targets.

\section{Summary}
\label{sect:summary}
The Arcus XRS is an X-ray spectrograph with four separate channels in a double-tilted Rowland torus configuration. All found channels are imaged onto the set of 16 CCDs, which are split over two cameras with 8 CCDs each. We discuss cross-contamination between channels due to cross-dispersion on the grating support structures and CCD frame transfer. For each channel, several orders will be diffracted towards the blaze peak and will be detected at the same physical detector location. Those orders need to be separated using the intrinsic energy resolution of the CCDs. We show that there will be a slight overlap between orders, but it is mostly relevant for very bright emission lines. All those effects can be handled in the OGIP scheme of PHA, ARF, and RMF files and thus the spectra can be fit with existing and well-tested software packages. We briefly discuss potential improvements of the fitting process by fitting in 2D or 3D space or using machine-learning techniques. Last, we calculate the expected non X-ray background rate for the XRS, which turns out to be so low that it is not relevant except for the faintest sources that we expect the XRS to observe.

% \disclosures
\subsection*{Disclosures}
The authors declare no conflicts of interest.

\subsection* {Code, Data, and Materials Availability}
We use the ray-tracing code MARXS\cite{2017AJ....154..243G,marxs2.0}, a Python code developed by us under an open source license. MARXS is available at \url{https://github.com/chandra-marx/marxs}; simulations in this article are done with version 2.0. The database of grating efficiencies and material properties is available upon request.

%In support of open scientific exchange, SPIE journals require Code, Data, and Materials Availability Statements in all accepted papers. This requirement went into effect on 1 May 2023. These statements should describe how to access any data that would be required to replicate or interpret the findings reported in the paper. Authors are encouraged to make the data and code related to the manuscript publicly available whenever possible, and utilize repositories that are well-known to the field (FigShare, Github, CodeOcean, etc.). If the data or code cannot be made publicly available, the authors should state the reason and explain how it can be obtained. Likewise, if data sharing is not applicable, the statement must say so. Example statements may be found in the Author Guidelines for the journal.

\subsection* {Acknowledgments}
Support for this work was provided in part through NASA grant NNX17AG43G and
Smithsonian Astrophysical Observatory (SAO) contract SV3-73016 to MIT
for support of the {\em Chandra} X-Ray Center (CXC), which is operated
by SAO for and on behalf of NASA under contract NAS8-03060.  The
simulations make use of Astropy, a community-developed core Python
package for Astronomy\cite{astropy:2022}, numpy\cite{numpy}, and IPython\cite{IPython}.

%%%%% References %%%%%

\bibliography{report_with_doi}   % bibliography data in report.bib
\bibliographystyle{spiejour}   % makes bibtex use spiejour.bst

%%%%% Biographies of authors %%%%%
\subsection*{Biographies}
\vspace{2ex}\noindent\textbf{Hans Moritz G\"unther} is a research scientist at MIT. He received his undergraduate degree (in 2005) and his PhD (in 2009) in physics from the University of Hamburg, Germany. After that, he worked at the Harvard-Smithsonian Center for Astrophysics and came to MIT in 2015. He is currently the lead developer of MARX, the ray-tracing software used for the Chandra X-ray observatory. His science interests are in star formation using data from the radio to X-rays.

\vspace{2ex}\noindent\textbf{David Huenemoerder} (Ph.D. Astronomy, 1982, University of
Wisconsin-Madison) joined the MIT Kavli Institute in 1992 as a
member of the Chandra X-ray Center, for which he is responsible
for Chandra grating processing and analysis.  He has overseen the
the development of the Chandra grating data catalog and
archive (tgcat.mit.edu).  His recent research is on the winds of
massive stars, diagnosed through high resolution X-ray
spectroscopy with the Chandra High Energy Transmission Grating
Spectrometer.

\vspace{2ex}\noindent\textbf{Eric Miller} is a Research Scientist at the MIT Kavli Institute for Astrophysics and Space Research. He received his BA in physics from Oberlin College in 1996, and his PhD in Astronomy and Astrophysics from the University of Michigan in 2003. He develops X-ray imaging detectors for future missions, leads the XRISM in-flight calibration team, and studies galaxy clusters and the diffuse intergalactic medium.

%\noindent Biographies and photographs of the other authors are not available.

%\subsubsection{Reference linking and DOIs}
%A Digital Object Identifier (DOI) is a unique alphanumeric string assigned to a digital object, such as a journal article or a book chapter, that provides a persistent link to its location on the internet. The use of DOIs allows readers to easily access cited articles. Authors should include the DOI at the end of each reference in brackets if a DOI is available. See examples at the end of this manuscript. A free DOI lookup service is available from CrossRef at \\\linkable{http://www.crossref.org/freeTextQuery/}. The inclusion of DOIs will facilitate reference linking and is highly recommended.

%In the present LaTeX template, the author needs to add the DOI reference by including it in a ``note'' in the bibliography file, as shown in the file {\verb+report.bib+}, for example, \\ {\verb+note = "[doi:10.1117/12.154577]"+}. The DOI may be used by the reader to locate that document with the link: {\verb+http://dx.doi.org10.1117/12.154577+}.

\listoffigures
\listoftables

\end{spacing}
\end{document}

### 2024-05-20 papers/special_JATIS_tolerances/tolerances.tex
submitted paper to JATIS

@@
 
+\usepackage{lineno}
+\linenumbers
+
 \title{Arcus X-ray telescope performance predictions and alignment requirements}
@@
 \author[a,d]{Ralf K. Heilmann}
-\author[a]{Eric D. Miller}
-\author[e]{Andrew Ptak}
-\author[b]{Randall K. Smith}
 
@@
 \affil[d]{Space Nanotechnology Laboratory, MIT Kavli Institute for Astrophysics and Space Research, Cambridge, MA 02139, USA}
-\affil[e]{NASA Goddard Space Flight Center, Greenbelt, MD 20771, USA}
 
@@
 \begin{abstract}
-Arcus is a concept for a probe class mission to deliver high-resolution FUV and X-ray spectroscopy with two separate instruments.
+Arcus is a concept for a Probe-class X-ray mission to deliver high-resolution FUV and X-ray spectroscopy with two separate instruments.
 The X-ray spectrograph (XRS) is the focus of this paper.
@@
 % FOR SUBMISSION
-% \begin{spacing}{2}   % use double spacing for rest of manuscript
+\begin{spacing}{2}   % use double spacing for rest of manuscript
 
 % For my own while writing. It's just easier to read this way
-\begin{spacing}{1}
+% \begin{spacing}{1}
 
@@
 Thus, observations in the UV and X-rays probe the most energetic processes in a number of systems.
-A few examples are: stellar space weather \cite{2022AN....34320019B}, the accretion onto young stars, where the UV and X-rays come from the energetic infall of material from the disk onto the stellar surface; the innermost regions of the accretion disks around black holes; and the absorption of background (X-ray) light from distant, bright continuum sources by the warm-hot inter-galactic medium to find the missing baryons\cite{10.1117/12.2231193}.
+A few examples are: stellar space weather \cite{2022AN....34320019B}, the accretion onto young stars, where the UV and X-rays come from the energetic infall of material from the disk onto the stellar surface; the innermost regions of the accretion disks around black holes; and the absorption of background (X-ray) light from distant, bright continuum sources by the warm-hot inter-galactic medium to find the missing baryons\cite{10.1117/12.2231193,10.1117/12.2529499}.
 High-resolution UV and X-ray spectroscopy that resolves the profiles of individual emission or absorption lines is particularly valuable, because it allows us to address a host of physical questions that cannot be answered by simply measuring the broad-band X-ray flux.
-To follow-up with one specific example: in the case of accretion onto young stars, resolving the kinematic line profile of emission in different ions can tell us which part of the emission is formed in the infalling, red-shifted accretion column, which part is related to the stellar corona seen at the rest wavelength of the star (possibly with small velocities from up and downflows in the corona), and which part is formed in (or possibly absorbed by) the blue-shifted outflow.
+To follow-up with one specific example: in the case of accretion onto young stars, resolving the kinematic line profile of emission in different ions can tell us which part of the emission is formed in the infalling, red-shifted accretion column, which part is related to the stellar corona seen at the rest wavelength of the star (possibly with small velocities from up and downflows in the corona), and which part is formed in (or possibly absorbed by) the blue-shifted outflow \cite{10.1117/12.2529499}.
 
@@
 ray-tracing to verify the performance characteristics of Arcus and to
-improve the design\cite{10.1117/12.2232157}; in particular in \citenum{10.1117/12.2312678} we utilized ray-tracing to determine the allowable misalignments of the mechanical elements that make up Arcus.
+improve the design\cite{10.1117/12.2232157}; in particular in \citenum{10.1117/12.2312678,10.1117/12.2677455} we utilized ray-tracing to determine the allowable misalignments of the mechanical elements that make up Arcus.
 Ray-tracing is particularly well suited for the problem, because it allows an arbitrary misalignment to be introduced for any element in the optical path and it can quickly iterate over different input parameters.
 Since the work in \citenum{10.1117/12.2312678}, the Arcus concept has evolved yet again.
-In its most recent version, Arcus is a NASA Probe-class mission with significantly increased effective area and the addition of the UV instrument.
-Since the UV telescope has a smaller field-of-view than the X-ray instrument, it will define the fiducial pointing direction; now misalignment between the two separate instruments is a new concern.
 
@@
 They provide a large effective area at a relatively low weight and cost\cite{10.1117/12.2188988,10.1117/12.2599339,10.1117/12.2677388}.
-In Arcus, the SPO petals use ``sub-aperturing'', which means that the SPOs do not cover a full circle, but only a narrow wedge.
-That provides Arcus with an asymmetric PSF that is only about 1.5\ arcsec wide in the dispersion direction.
+In Arcus, the SPO petals use ``sub-aperturing'', which means that the SPO do not cover a full circle, but only a narrow wedge.
+That provides Arcus with an asymmetric point-spread function (PSF) that is only about 1.5\ arcsec wide in the dispersion direction.
 It also means that the individual petals where the SPOs are mounted are not circular, but roughly rectangular, and four separate panels (one per optical channel) can fit into the front assembly of the telescope; see Figure~\ref{fig:arcus} for a rendering.
@@
 Critical-angle transmission (CAT) gratings are manufactured at the Space Nanotechnology Labo\-ra\-tory.
-Progress on the development and manufacturing of the CAT gratings is given in a series of papers stretching back many years\cite{10.1117/12.926827,2022ApJ...934..171H}.
+Progress on the development, manufacturing and x-ray performance of CAT gratings is given in a series of papers stretching back many years\cite{OE2008,AO2011,EIPBN2016,AO2019,2022ApJ...934..171H}.
 CAT gratings have a very high aspect ratio, where the grating bars are about 50 times higher than they are wide.
@@
     \begin{tabular}{c} %% tabular useful for creating an array of images
-    \includegraphics[height=5cm]{unitcell+pic.jpg}
+    \includegraphics[height=5cm]{unitcell+pic.png}
     \end{tabular}
@@
 The different tilted tori are arranged in such a way, that they overlap in the Rowland circle, so that the optimal detector position for one channel is also the optimal detector position for all other channels.
-Two cameras are positioned to capture the zeroth order on one side and the blaze peak on the other with a gap in between where so little signal is found that it is not useful to capture it with CCDs.
+Two cameras are positioned to capture the zeroth order on one side and the blaze peak on the other.
+There is a gap between the two cameras because so little signal is found in that region that it is not useful to capture it with CCDs.
 
@@
 Given the resolution of the figure, most chip gaps are not visible.
-Also note, that the x and the y axes are scaled differently (Figure~\ref{fig:4} shows correct aspect of the CCDS, but not the correct gap between the two detector arrays).
+Also note, that the x and the y axes are scaled differently.
 Two cameras with eight CCDs each are located in the focal plane.
@@
 Each of the four channels of the XRS has a separate zero-order position and disperses the light to a different location on the detector.
-This has two main advantages: there is no need to align all channels to the same zero-order position within the width of the point-spread function (PSF, less than 1.5 arcsec in dispersion direction), and any wavelength of light that falls in a chip gap for one channel will be observable in the others; see Figure~\ref{fig:geometry_camera}.
+This has two main advantages: there is no need to align all channels to the same zero-order position within the width of the PSF (less than 1.5 arcsec in dispersion direction), and any wavelength of light that falls in a chip gap for one channel will be observable in the others; see Figure~\ref{fig:geometry_camera}.
 At the same time, this layout allows us to use the same detectors for all four channels, which reduces cost and complexity.
@@
 We study the performance and alignment requirements of Arcus with ray-tracing.
-We use the ray-tracing code MARXS\cite{marxs1.1,2017AJ....154..243G}, a Python code developed by us under an open source license.
-MARXS is available at \url{https://github.com/chandra-marx/marxs}; simulations in this article are done with a development version with git hash f4b68dce0957 to include Arcus specific code.
+We use the ray-tracing code MARXS\cite{2017AJ....154..243G,marxs2.0}, a Python code developed by us under an open source license.
+MARXS is available at \url{https://github.com/chandra-marx/marxs}; simulations in this article are done with version 2.0.
 MARXS has been used for several mission proposals and is backed by hundreds of units tests and verified against laboratory works and on-orbit Chandra observations where feasible.
@@
 After testing out the parameter space in dispersion direction, the CCDs will be shifted in cross-dispersion direction.
-In this way, each element (for example the CCD array), will be misaligned by various amounts in one of 6 degrees of freedom (shift along $x$, $y$, $z$, and rotation around $x$, $y$, $z$).
+In this way, each element (for example, the CCD array), will be misaligned by various amounts in one of 6 degrees of freedom (shift along $x$, $y$, $z$, and rotation around $x$, $y$, $z$).
 In the Arcus coordinate system, the $z$-axis is parallel to the optical axis, the $x$-axis is the grating dispersion direction and the $y$-axis is the cross-dispersion direction.
@@
 In a second step, we can then run ray-traces where all degrees of freedom are varied according to the misalignment budget and thus check if the assumptions going into combining the misalignments in different degrees of freedom hold or if non-linear interactions degrade $R$ and $A_\mathrm{eff}$ more than expected.
-In this case, we would have to revise the misalignment budget appropriately.
+Should this be the case, we revise the misalignment budget appropriately.
 
@@
 Those photons coming from the high CAT gratings travel further along the dispersion direction than those from the low gratings, thus causing the dispersed spot to smear out, which reduces $R$.
-These rotation need to be kept at the level of a few arcminutes.
+These rotations need to be kept at the level of a few arcminutes.
 Rotation around $z$ changes the direction of the dispersed light and, for large angles, the dispersed orders miss the CCD.
@@
 Rotations around $x$ (the dispersion direction, see panels a and d in the figure) have little effect on $R$.
-Since the facets have finite size, their edges differ from the Rowland torus by a little already, and adding a little more rotation does not change much.
+Since the flat facets have finite size, their edges differ from the Rowland torus by a little already, and adding a little more rotation does not change much.
 Rotations in $y$ quickly reduce $A_\mathrm{eff}$ though, because the CCDs are placed for a certain blaze peak.
@@
 A shift along $y$ has no impact, as long as it is small enough to keep the dispersed spectrum on the CCDs.
-For the particular channel simulated here the spectrum drops off the CCDs for $y$ shifts for about -15~mm on one side and about +5~mm on the other side.
+For the particular channel simulated here, the spectrum drops off the CCDs for $y$ shifts for about -15~mm on one side and about +5~mm on the other side.
 A shift of 5~mm or more will drop at least one channel of the detector.
@@
 
-We run simulations for rotations and translations for each possible mechanical misalignment in Arcus, as well as for a few other parameters such as the SPO mirror PSF, the repeatability of the CAT grating period, the surface flatness of the CAT gratings etc, but it is beyond the scope of this article to show each result in a figure.
+We run simulations for rotations and translations for each possible mechanical misalignment in Arcus, as well as for a few other parameters such as the SPO mirror PSF, the repeatability of the CAT grating period, the surface flatness of the CAT gratings etc.
 Inspecting all results, and considering how easy or hard it is to improve alignments in each degree of freedom, we set the alignment tolerances listed in Table~\ref{tab:tolerances}.
@@
     individual CAT to window & 0.333 & 0.333 & 0.067 & 100 & 60 & 100 \\
-    Camera to front assembly & 1.667 & 0.667 & 0.333 & 60 & 60 & 60 \\
+    Camera to front assembly & 1.667 & 0.667 & 0.333 & 60 & 60 & 200 \\
     \end{tabular}
@@
 
+We use the feedback we gain from our ray traces to revisit the error budget and look for places where we can relax it without impacting performance.
+For example, we have been able to relaxe rotational alignment tolerances for CAT grating membranes and/or facets to their grating window by a factor of three relative to Table 1 with acceptable minimal impact on XRS performance.
+
 \section{Future Work}
@@
 We expect it to work well for on-axis sources, but it does not reproduce the vignetting which becomes significant at about 2~arcmin.
-Our simplified mirror model also leads to a flat focal plane, while a Wolter-Type II optic typically has a slightly curved focal surface.
+Our simplified mirror model also leads to a flat focal plane, while a Wolter-Schwarzschild optic typically has a slightly curved focal surface.
 Detailed ray-trace models specific to the Athena SPOs are available\cite{10.1117/12.2232230,10.1117/12.2594461,10.1117/12.2628133} and can be coupled with our code for CAT gratings and detectors.
@@
 \subsection* {Code, Data, and Materials Availability}
-We use the ray-tracing code MARXS\cite{marxs1.1,2017AJ....154..243G}, a Python code developed by us under an open source license.
-MARXS is available at \url{https://github.com/chandra-marx/marxs}; simulations in this article are done with a development version with git hash f4b68dce0957.
+We use the ray-tracing code MARXS\cite{2017AJ....154..243G,marxs2.0}, a Python code developed by us under an open source license.
+MARXS is available at \url{https://github.com/chandra-marx/marxs}; simulations in this article are done with version 2.0.
 The database of efficiencies for gratings is considered a trade secret by the Space Nanotechnology Laboratory and cannot be shared in open repositories - contact co-author Ralf Heilmann for access.
@@
 \vspace{2ex}\noindent\textbf{Peter Cheimets}
-Please send me your biography (approx 75 words).
+Peter Cheimets received a BS in Mechanical Engineering in 1978 from MIT, and a MS in Mechanical Engineering with concentrations in dynamics and control systems from Stanford in 1981.
+He received a management of engineering degree from Tuft in 2007.
+He has developed astrophysical instrumentation for the Smithsonian Astrophysical Observatory since 1983, placing instruments on TRACE, Hinode, SDO, and IRIS, and is leading the effort to develop an instrument for MUSE.
 
@@
 \vspace{2ex}\noindent\textbf{Casey DeRoo}
-Please send me your biography (approx 75 words).
+Casey DeRoo received his BA degree in physics and classical studies in 2011 and his PhD in physics from University of Iowa in 2016.
+He served as an Astrophysicist at the Smithsonian Astrophysical Observatory.
+He joined the faculty of the University of Iowa as an assistant professor in 2018, and was named a NASA Astrophysics Nancy Grace Roman Technology Fellow in 2020.
+His research interests include X-ray optics technologies, nanofabrication, and metrology.
 
@@
 
-\vspace{1ex}
-\vspace{2ex}\noindent\textbf{Eric D. Miller}
-Please send me your biography (approx 75 words).
+%\vspace{1ex}
+%\vspace{2ex}\noindent\textbf{Eric D. %Miller}
+%Please send me your biography (approx 75 words).
 
-\vspace{1ex}
-\vspace{2ex}\noindent\textbf{Andrew Ptak}
-Please send me your biography (approx 75 words).
+%\vspace{1ex}
+%\vspace{2ex}\noindent\textbf{Andrew Ptak}
+%Please send me your biography (approx 75 words).
 
-\vspace{1ex}
-\vspace{2ex}\noindent\textbf{Randall K. Smith}
-Please send me your biography (approx 75 words).
+%\vspace{1ex}
+%\vspace{2ex}\noindent\textbf{Randall K. Smith}
+%Please send me your biography (approx 75 words).
 

### 2024-08-20 papers/Doubletiltconcept/referee_reply_1.txt
second revision of Double tilted concept paper

We thank the referee for their time spent reviewing the manuscript. The quality of the feedback expressed in the review makes it apparent that the referee is an expert in the Rowland geometry. However, as we note in our response below, we disagree that the general principles of the geometry are as widely known by the broader astronomical community as claimed by the referee. Moreover, none of the prior discussion of the DTRS concept appears in peer-reviewed publications. This paper is intended to address these shortcomings of the literature.

The referee has made some specific suggestion with figures to combine or remove and we address those below (note that this means that figures in the new manuscript are renumbered and might not match the figure numbers in the previous referee's comments any more). We highlighted new or changed text in bolt font for the re-submission, but we also shortened several of the discussions in the paper, consistent with the referee’s suggestion to focus on information not presented elsewhere. Given the difficulty of reading through the substantial amount of strike-through figures and strike-through text, we opted to fully remove (instead of strike-through) text that we no longer need.

Below, our specific responses to the referee's comments are keyed in between the referee's comments.

Reviewer:
The paper gives a full description of the 'concept of a Double Tilted Rowland Spectrograph for X-rays' but the level of new information is rather limited given the extensive literature by the authors and the reference to FUSE (moose 2000) and many others .

>>> We want to point out that the FUSE design (Moose 2000) is similar in some aspects, but very different on a few key points:
- FUSE uses reflective gratings, not transmission gratings.
- In FUSE, the gratings are illuminated at almost normal incidence and thus entrance aperture and detectors are on the same side of the Rowland circle, not on opposite sides as in our design.
- FUSE does not detect the zeroth orders.
- FUSE does not place all diffracted spectra next to each other on the same read-out array.

For publishing it in ApJ my recommendation is to limit the paper to the really new information and not describe the details (and even repeat them in the paper itself). From my understanding the really new information is probably:

- The use of an offset in the cross dispersion direction
- The discussion about alignment accuracies but, these are only described qualitatively whereas the new information (using the tools) is probably the quantitative information
- Correctly one of the identified drawback is the background (in section 4.2.2). Quantifying under which conditions DTRS or a more classical Rowland angle design is better would be a really new contribution
- Various aspects are not discussed in the paper including order separation and mirror scattering in the cross dispersion direction. Probably this is nicely included on the available tools but how this impacts the performance is not quantified

Hence I would recommend to ask the authors to rewrite the paper significantly, limiting it to the real new information compared to previous articles.

>>> We believe that the comments above are partially related to a misunderstanding about what has been published in "previous articles". It is true that we have written about the concept of the Double Tilted Rowland Spectrograph (DTRS) before specifically for the Arcus mission. However, that work so far has only appeared in conference proceedings (Proceedings of the SPIE), not in refereed journals. We believe that a prior publication in a conference proceeding should not preclude publication of a proper article. Those previous proceedings are also specific to Arcus: They use the Arcus focal length, the specific size in mm for the gratings that Arcus is proposed to have etc. On the one hand, that allows us to be very specific about the resolving power and effective area of the instrument in those conference proceedings, but does not include a general discussion of underlying CONCEPT in general. In this article, the discussion is necessarily more qualitative (because it is not specific to a specific number for focal length, grating size etc.), but more general, so we believe that it adds some value even for those readers who are familiar with with our previous proceedings; we thus regard content as "new" even if it has been described specifically in the Arcus proceedings before.
We have added sentences to the introduction to make this goal and the quality of prior work explicit.
We have also gone through the text to remove duplication and improve the flow by shortening, in particular in section 2, which describes the background for our work. In total, the re-submitted work is about 1.5 pages shorter than our original submission.

For previous papers you can also look also the following papers although the paper by Moose describes most of the technology.

Varied Line-Space Gratings: Past, Present and Future by Michael C. Hettrick, Proc. Soc. Photo-Opt. Instr. Eng. vol. 560, pp. 96-108 (1986)
http://dx.doi.org/10.1117/12.949620

>>> We have added this reference to the section about "dispersive elements" (sect 2.2).

Space Sci Rev (2010) 157: 15-24 DOI 10.1007/s11214-010-9719-y, X-ray Diffraction Gratings for Astrophysics by Frits Paerels

>>> Given the length of the paper (which the referee suggested to shorten significantly), we had chosen to not discuss those review articles. However, we agree that this is a useful article and added a reference to this paper.

Draft version June 22, 2022 Typeset using LATEX default style in AASTeX631 X-ray performance of critical-angle transmission grating prototypes for the Arcus mission by Heillman

>>> This article was already cited and discussed in the version we initially submitted. That work discusses the details of how the gratings are made and how efficient they are, but not how they are used in a Double-Tilted Rowland Spectrograph.

All figures: add coordinates to the plots (X, Y, Z (is the direction of the X-rays?)

>>> Fig 1 and 2 already had x,y,z. For the 3D plots actually adding a floating coordinate system that stays visible and renders well in the webbrowser is surprisingly difficult and required significant effort, but it has now been added to the 3D renderings. In Fig 9, we just describe the relation to the global X,Y,Z in the figure caption, since the dispersion coordinate is read along the curved Rowland circle. In our layout, not even at the focal points does this exactly coincide with the x axis of the coordinate system.

Datailed comments (apart of the length of the paper):

Section 2.1 end: figure 1 is well known and does not need to be illustrated and can be summarized shortly.

>>> When we presented this or similar work in talks, we've found that even people who were involved in the design of Chandra are surprised when we point out the LETG and HETG would have higher resolving power, had Chandra chosen to sub-aperture. It thus seems that the community to which this is "well known" is limited to instrument designers. Since ApJ is a general purpose astronomy journal we've chosen to include this as a figure to make it clear. We admit that this figure does not contain "new information" but merely illustrates previous knowledge, but given how central this is this to the rest of the article, we suggest that it is worthwhile to show in a figure and have thus left it in this version - if this argument does not convince the referee, we will remove it in the next resubmission.

Section 2.2: a correct statement about 'larger gratings (line 7) but this is not the topic of this paper)

>>> Grating size is not the main topic of this paper, but the main motivation for Heilmann et al. to introduce a tilted Rowland torus to transmission gratings was to reduce the average deviation between grating and Rowland torus. For small gratings, that distance is always small, so it seems relevant to at least bring up the topic of grating sizes.

Section 3 Figure 3: in figure 2 the suggestion is given that the mirror covers a circular sub-aperture of the full mirror plane while in the ARCUS paper this is much clearer described (two 'rectangular' subsections. This would also make figure 4 more understandable.

>>> We like the referee's suggestion (in a point below) to also add the aperture to the figure that shows the distribution of photons on the focal plane (currently Fig 5). That clearly shows the area covered by mirror+gratings.
In Fig 2, we have shown a tilted, but not sub-apertured scenario. In our setup for this scenario, the mirror does cover the full circle, not a rectangular subsection.
We also want to note that, in our previous version, we had chosen to use wedges, not rectangles for layout of the gratings in the sub-apertures. For the purpose of this article though the exact shape does not matter - our simulations are meant to illustrate the concept, not to match any particular mission. However, since our use of wedges seems to be unclear, we have redone this now with rectangles for the grating layout to better match the shape of the green boxes we use for the mirrors in our visualization. (Arcus uses something in between that is not quite a rectangle, but is wider at the tip than at the bottom.)

Section 3.2 Figure 4: what is the added value compared to fig 2

>>> In the electronic version, Fig 4 is interactive and Fig 2 is not. For the static version of the article, Fig 4 shows four views including a view of the focal plane, and a zoom-in on the gratings, which is not visible in Fig 2.

Section 3.2 just below figure 4: The use of tilted Rowland circles is used already for a long time in space missions (see RGS on XMM-Newton). The new step introduced by Heilman is that this can also be achieved by Critical Angle Transmission gratings. In this sense using tilted Rowland cicles is a well known concept.

>>> We re-worded that sentence.

Frig 5: blue and black colors are not easy to distinguish. Also the figure becomes more clear if at the rich side a panel is shown with the mirror aperture used for each of the sub-panels.

>>> There were no black points in Fig 5, maybe the referee meant dark purple and blue? We understand the point, but there are are only so many colors we can choose from, also taking into account common combinations to avoid for accessibility (e.g. red/green blindness). We've re-done the figure with different colors. We thank the referee for the suggestion to add the mirror aperture to the figure.

In the same paragraph (5 lines lower) high diffraction efficiencies are mentioned. Omn the various simulatiopns/descriptions this is further neglected. It wopuld be good to quantify this.

>>> We added a number and a reference.

Equation 3: I do not believe anywhere the y component is used (but probably it is included as it is part of the software tools (but then y not equals zero).

>>> We want to be explicit here that the focal point is indeed in the plane that we show in the sketches - the sketches are not a projection onto the z=0 plane, but a cut at the z=0 plane and the focal point is in that plane. (With the addition of the x,y,z vectors to all images, we standarized the coordinate system so that z is now the cross-dispersion direction, so whatever was previously called y is now z.)

Section 3.3: Figure 6 and 7 are another set of similar figures, reducing it to fig 7 should be considered as in fig 2 you describe the basic principle and in fig 7 you give the final concept.

>>> Going from one Rowland Torus to two is a pretty big conceptual step. We have explained the development of the double-tilted Rowland torus in several talks with different audiences (both observing astronomers and instrument builders) and always found that it is conceptually challenging, thus we want to make sure that there is enough detail in the article to explain to the reader how we get there, not just "what the answer is". Thus, our attempt to break it in four steps:
1. Chandra-like (and thus well known to many in the community) - old Fig 4
2. tilted torus (old Fig 6)
3. two tilted tori (old Fig 7)
4. two tilted tori, with the two parts of the two tori offset from each other (old Fig 8).

However, we also understand the referee's concern that the article is long and, for the expert reader, contains too much material that is repeated. So, in order to reduce the number of figures, we've removed the old Fig 7 since Fig 7 and 8 look much more similar than Fig 6 and 7.
(Note however, that by reordering material and removing figures, the figure numbers have changed in the resubmitted version.)

Section 3.4.1 is probably the most relevant extension of earlier papers although I did not check if this is not already in the published ARCUS proposal

>>> We want to be explicit that there is no such thing as a "published ARCUS proposal". Like all proposals, the Arcus proposal submitted to NASA is strictly confidential; even if the referee has access to this proposal in some capacity, it is not relevant to the review of this article.
The goal of this article is to describe the general concept in the refereed literature, as opposed to previous conference proceedings where we always wrote about one specific mission concept.

Section 4.1.1: The advantage of subaperturing is presented at the cost of the technical compelkxity achieving a better PSF. As such it is true but one limits the usable open area provided by a satellite (with a circular shaoe). Alternatively one populate the unsused open area at the cost of an ectra detector to collect the dispersed X-rays in a 90 degree rotated system

>>> We have looked at extra detectors before, but one goal of the DTRS design is to avoid this. We added a short discussion of this alternative.
The question of what is a better use of the open area depends, as the referee points out, on the shape of the opening and the shape of the mirror. For example, in Arcus, the open area is constrained by the shape of the boom (square) while in Chandra the open area is a circle. For Arcus or Athena where the mirror is made out of stacks of SPOs the "mirror shape" can be arranged in different ways for a sub-aperture (e.g. a square or a wedge), while mirrors using full shells like Chandra cover a circle. Thus, there can be no general answer - it depends on the specific mission design and is not an advantage of disadvantage of the DTRS as a general concept.
We discussed a setup like the referee suggested (with a 90 degree rotated system) in a previous paper focussed on the Lynx mission. We have added a reference to that paper.

Figure 9: it could be good to identify which parts of the recorded data is to to the dispersion of the photons perpendicular to the main dispersion direction. How does this affect the dispersed spectram?

>>> While this is an interesting discussion, we believe that it is beyond the scope of this article. The referee already pointed out that the length of the article is a concern, and thus we intentionally left out the problem of order-sorting and a quantitative analysis of cross-dispersion. Both depend sensitively on the specific design and this this is hard to generalize without discussing a specific mission concept, which is not in scope for ApJ instrumentation papers. For example, the fraction of photons that are cross-dispersed depends almost linearly on the spacing of cross-dispersion structure and the dimensions of the support structure. Prototype gratings for Arcus use bars with 0.9 mu width for the support structures in cross-dispersion direction, but e.g. for the gratings Lynx the bars would have been thinner and thus the fraction of photons that are cross-dispersed would be lower; one could also test if larger spacing of the grating support structures in the cross-dispersion direction are sufficient for structural stability and reduce the cross-dispersion that way. So it really depends on the specific gratings and cannot be answered in general.

Fig 10: contains essentially the same as figure 9, I would reduce this to a single figure

>>> We originally included Fig 10 because some details are better seen in the larger scale compared to Fig 9, but the referee is correct: The same data is shown and thus we removed Fig 10 and the associated text to reduce the length of the article.

### 2024-08-20 papers/Doubletiltconcept/doubletiltconcept.tex
second revision of Double tilted concept paper

@@
 \affiliation{MIT Kavli Institute for Astrophysics and Space Research, 77 Massachusetts Avenue, Cambridge, MA 02139, USA}
-\author[0000-0002-9184-4561]{Casey DeRoo}
+\author[0000-0002-9184-4561]{Casey T. DeRoo}
 \affiliation{Dept. of Physics \& Astronomy, University of Iowa, Iowa City, IA 52242, USA}
@@
 
-%% From the front matter, we move on to the body of the paper.
-%% Sections are demarcated by \section and \subsection, respectively.
-%% Observe the use of the LaTeX \label
-%% command after the \subsection to give a symbolic KEY to the
-%% subsection for cross-referencing in a \ref command.
-%% You can use LaTeX's \ref and \label commands to keep track of
-%% cross-references to sections, equations, tables, and figures.
-%% That way, if you change the order of any elements, LaTeX will
-%% automatically renumber them.
-%%
-%% We recommend that authors also use the natbib \citep
-%% and \citet commands to identify citations.  The citations are
-%% tied to the reference list via symbolic KEYs. The KEY corresponds
-%% to the KEY in the \bibitem in the reference list below.
-
 \section{Introduction} \label{sec:intro}
@@
 
-The two largest X-ray observatories in history both field diffraction gratings: The Reflection Grating Spectrometer (RGS) on XMM-Newton \citep{2001A&A...365L...7D} and both the High-Energy Transmission Grating  \citep[HETG,][]{2005PASP..117.1144C} and the Low-Energy Transmission Grating \citep[LETG,][]{1997SPIE.3113..172P} on Chandra.
-Unlike microcalorimeters, diffraction gratings require an additional dispersive element (a transmission or a reflection grating) in the optical path of the instrument and dedicated detectors to capture the dispersed signal.
+The two largest X-ray observatories in history both field diffraction gratings \citep[see][for a review]{2010SSRv..157...15P}: The Reflection Grating Spectrometer (RGS) on XMM-Newton \citep{2001A&A...365L...7D} and both the High-Energy Transmission Grating  \citep[HETG,][]{2005PASP..117.1144C} and the Low-Energy Transmission Grating \citep[LETG,][]{1997SPIE.3113..172P} on Chandra.
+Unlike microcalorimeters, grating spectrographs require an additional dispersive element (a transmission or a reflection grating) in the optical path of the instrument and dedicated detectors to capture the dispersed signal.
 For a given diffraction order, gratings typically deliver a fixed wavelength resolution $\Delta \lambda$, so they perform better than microcalorimeters at longer wavelengths/lower energies.
@@
 In this article, we present an optical layout for an instrument that stacks two or four different channels with distinct optical axes in such a way that each channel follows the Rowland torus prescription, yet at the same time all channels are imaged onto a common set of detectors.
-The general concept could be applied to a large range of focal lengths, grating constants, and other hardware properties.
-Thus, we discuss advantages and challenges of this layout in general.
-However, we note that Arcus \citep{2023SPIE12678E..0ES}, an X-ray spectroscopy mission proposed for NASA MIDEX (medium explorer) and Probe calls, applies this layout.
-Specific details of the  Arcus performance, such as predicted effective area and resolving power are shown in \citet{2018SPIE10699E..6FG,2023SPIE12678E..1DG}, but the general concept of a double tilted Rowland spectrograph (DTRS) with its advantages and drawbacks has not been discussed -- a void we attempt to fill with this article.
+
+This concept has been developed and described in the context of the Arcus mission, which was originally proposed as a NASA Medium Explorer \citep{2019SPIE11118E..0WS} and  has since been proposed to the NASA Astrophysics Probe call \citep{2023SPIE12678E..0ES}.
+However, the DTRS concept is applicable to mission architectures beyond that of Arcus, and could be applied to a broad range of focal lengths, grating parameters, etc.
+This paper aims to discuss the advantages and challenges of this layout from a general perspective.
+We note that this discussion will be necessarily qualitative (rather than quantitative) as performance metrics such as resolving power and effective area depend on the design parameters that will be specific to a given instrument (e.g., focal length, efficiency).
 
 The structure of this article is as follows: First, we discuss the properties of the optical elements that are most influential for the design of a DTRS, namely mirror properties, gratings, and detectors in Sect.~\ref{sect:elements}.
-Then, we illustrate the optical layout, modifying the design of a traditional single-channel Rowland torus spectrograph to a DTRS.
+Then, we illustrate the optical layout, starting by modifying the well-known single-channel Rowland torus spectrograph design step-by-step to a DTRS.
 In section~\ref{sect:discussion}, we discuss advantages and disadvantages of our DTRS concept in general and for the example of Arcus in particular (section~\ref{sect:applications}).
@@
 The higher the energy of the X-ray, the smaller the angle between the mirror surface and the incoming ray has to be to allow for efficient reflection.
-The size of the collecting area is to first order proportional to the number of photons that the spectrograph can detect, and the spectral resolving power $\lambda / \Delta \lambda$ increases when the mirror point-spread-function (PSF) decreases.
+The size of the collecting area is to first order proportional to the number of photons that the spectrograph can detect, and the spectral resolving power $R=\lambda / \Delta \lambda$ increases when the mirror point-spread-function (PSF) decreases.
 Depending on the type of mirror chosen, the PSF can be dominated by the alignment error between mirror shells, in particular for mirror assemblies that consist of hundreds of elements.
@@
 Photons passing mirror and grating close to $z=0$ (dark purple) are spread out along the dispersion direction ($y$ coordinate); photons passing mirror and gratings close to $y=0$ are more concentrated along the dispersion direction.
-Sub-aperturing means that the mirror and gratings do not cover a full circle, but only regions marked in yellow and green in the left panel, so that the resulting photon distributions on the detector are narrower in dispersion direction.
-Sub-aperturing thus reduces the PSF in the dispersion direction, which increases the spectral resolving power.}
+Sub-aperturing means that the mirror and gratings do not cover a full circle, but only regions marked in yellow and green in the left panel, so that the resulting photon distributions on the detector are narrower in dispersion direction, which increases the spectral resolving power.}
 \label{fig:subaperture}
@@
 \end{equation}
-where $\lambda$ is the wavelength of the light, $n$ is the order of the diffraction, $d$ is the grating constant, and $\theta$ is the angle between the incoming light and the normal to the grating.
+where $\lambda$ is the wavelength of the light, $n$ is the order of the diffraction, $d$ is the grating constant, and $\theta$ is the angle between the outgoing light and the normal to the grating.
 (The equation can be rewritten for rays hitting the grating at arbitrary angles, but for the purpose of this conceptual discussion this is not needed.)
-Typically, these transmission gratings are flat and only a few cm$^2$ in size; the spacing between gratings bars $d$ is the same for all gratings and does not vary within a grating.
+Typically, these transmission gratings are flat and only a few cm$^2$ in size; the spacing between gratings bars $d$ is the same for all gratings.
+% and does not vary within a grating.
 While a flat grating necessarily deviates from the curved surface of a torus, in today's instruments those deviations are small enough that they do not impact the resolving power of the spectrograph significantly.
-Larger gratings would increase the effective area of the instrument because less space is needed for frames and other mounting structures that hold the gratings in place, but they would also deviate more from the surface of the torus.
-\citet{2020SPIE11444E..88G} explored bending the gratings in one dimension or varying the distance between grating bars over a grating to compensate for the deviations from the ideal shape in simulations.
-\citet{2019SPIE11118E..11G} shows that CAT gratings can be bent to the required radii without degrading their performance.
+Larger gratings would increase the effective area of the instrument because less space is needed for frames and other mounting structures that hold the gratings in place, but they would also deviate more from the surface of the torus thus reducing $R$.
+To prevent a loss of $R$ with larger gratings one can use gratings where $d$ varies across the grating \citep{1986SPIE..560...96H,2020SPIE11444E..88G} or bend the gratings \citep{2020SPIE11444E..88G}.
+\citet{2019SPIE11118E..11G} shows that gratings can be bent to the required radii without degrading their performance.
 
 \subsection{Detectors}
-The dispersed spectrum is spread out far in the dispersion direction.
-Thus, the ideal detector is much longer in one direction than in the other.
-Individual detector elements (e.g.\ CCDs) can be tiled, but to avoid chip gaps where photons are lost and gaps appear in the extracted spectrum, it is beneficial to keep the number of elements low.
-In our DTRS design several spectral traces are positioned in parallel (section~\ref{sect:opticallayout}), so the detector has to be position sensitive in 2D to separate individual channels.
+%The dispersed spectrum is spread out far in the dispersion direction. Thus, the ideal detector is much longer in one direction than in the other.
+%Individual detector elements (e.g.\ CCDs) can be tiled, but to avoid chip gaps where photons are lost and gaps appear in the extracted spectrum, it is beneficial to keep the number of elements low.
+In our DTRS design several spectral traces are positioned in parallel (section~\ref{sect:opticallayout}), so the detector has to be position sensitive to separate individual channels.
 Because it covers a larger area than in direct imaging observations, a detector with moderate cost and power requirements per surface area is well-suited for this design.
 The detector should follow the shape of the Rowland circle.
-This can be achieved by using curved detectors or by tiling smaller flat detector elements where each element is tangential to the Rowland circle.
+This can be achieved by using curved detectors or by tiling smaller flat detector elements.
+% where each element is tangential to the Rowland circle.
 
-The gratings spread out the signal over a large physical region on the detector.
-Thus, the counts rates per pixel are much lower than they would be in an imaging observation.
-Detectors that might be at risk of pile-up in direct observations can still be used for the dispersed signal.
+The gratings spread out the signal over a large region on the detector.
+Thus, the count rates per pixel are lower than they would be in an imaging observation.
+Detectors that might be at risk of pile-up in direct observations can be used for the dispersed signal.
 On the other hand, the larger regions needed to extract the signal imply that background events (internal background and diffuse astrophysical emission) make up a larger fraction of the extracted signal than in direct imaging observations.
-Similarly, fast timing resolution is typically not essential, because the lower signal means that the detected photons need to be binned up into longer time bins for analysis anyway.
+%Similarly, fast timing resolution is typically not essential, because the lower signal means that the detected photons need to be binned up into longer time bins for analysis anyway.
 Intrinsic energy resolution is required to both suppress the background by filtering out events that are incompatible with wavelengths of the dispersed signal at any particular location, and to separate grating orders where photons of different orders are diffracted to the same position (e.g.\ the sixth order for the O~{\sc vii} forbidden line of the He-like triplet at 2.21~nm and the seventh order of 1.94~nm photons, close to the O~{\sc viii} Lyman $\alpha$ line end up at almost the same $\theta$ according to equation~\ref{eqn:diffraction}).
@@
 We start from a single central Rowland torus as in Chandra/HETGS and Chandra/LETGS.
-We then show how this design can be modified step-by-step to accommodate blazed gratings and what considerations are important when choosing the specific parameters for a design.
-We illustrate the concept with ray-trace simulations.
-Those simulations are performed with the marxs code \citep{2017AJ....154..243G}.
-However, the setup is chosen to illustrate the concept of a DTRS, not to represent any particular instrument.
+We then show how this design can be modified step-by-step.
+The general concept is shown using ray-traces with the marxs code \citep{2017AJ....154..243G} using a simple setup for illustration.
 For example, the position of the diffractive gratings shown leaves valuable mirror area uncovered and is done without considering the size and location of support structures that the gratings can be mounted on.
 The grating constant $d$ is chosen to be smaller than feasible for a real instrument.
-Static images of the 3D ray-traces are embedded in the article, but we encourage the reader explore the interactive 3D models embedded in the electronic version of the publication.
 
@@
 \begin{equation}
-\vec p(\varphi, \theta) = \vec c + R \; \vec e_R(\varphi) + r \; (\vec e_y \sin \theta + \vec e_R(\varphi) \cos \theta)
+\vec p(\varphi, \theta) = \vec c + R \; \vec e_R(\varphi) + r \; (\vec e_z \sin \theta + \vec e_R(\varphi) \cos \theta)
 \end{equation}
 where $\vec c$ points to the center of the torus.
-The axis of symmetry is given by a unit vector $\vec e_y$.
-We define a vector $\vec e_R(\varphi) = \vec e_x \cos\varphi + \vec e_z \sin\varphi$.
+The axis of symmetry is given by a unit vector $\vec e_z$.
+We define a vector $\vec e_R(\varphi) = \vec e_x \cos\varphi + \vec e_y \sin\varphi$.
 $R \vec e_R$ points from the center of the torus to the center of a circle with radius $r$.
@@
 The symmetry axis of the torus  is marked as blue long-dashed line.
-The two blue circles show the cut through the 3D torus in this plane; they are connected by blue dotted lines to help visualize the 3D position of the torus.
-$\alpha$ is the angle between the optical axis and the line from the point where the optical axis intersects the Rowland torus (A) to the center of the Rowland circle.
+\textbf{The torus is formed by rotating the blue circle around this line.
+To help visualize the 3D position of the torus blue dotted lines are shown connecting the two circles.} $\alpha$ is the angle between the optical axis and the line from the point where the optical axis intersects the Rowland torus (A) to the center of the Rowland circle.
 $\beta$ is the angle from optical axis to the line connecting A to the ``hinge'' -- the other point in this plane where symmetry axis of the torus and the Rowland circle intersect.
@@
 Colored lines show the path of individual rays starting on the left.
-Rays detected in the zeroth order are shown in purple, green and yellow are higher orders.
-Optical axis, the axis of the Rowland torus etc.\ are indicated in blue and the location of detectors in orange, matching the sketch in the left panel.
+Rays detected in the zeroth order are shown in green, blue and purple are higher orders.
+Optical axis and the axis of the Rowland torus are indicated in blue and the location of detectors in orange, matching the sketch in the left panel.
         }
@@
 Figure~\ref{fig:3d:single} shows a 3D view of a ray-trace.
-Since the diffraction gratings are symmetric with respect to the optical axis, positive or negative diffraction orders (up or down in figure~\ref{fig:sketch}) are equally likely, and the diffracted photons will be symmetric with respect to the focal point.
+Since the diffraction gratings are symmetric with respect to the optical axis, positive or negative diffraction orders (up or down in figure~\ref{fig:sketch}) are equally likely.
+%, and the diffracted photons will be symmetric with respect to the focal point.
 To avoid the same wavelength to fall in a chip gap on the detector for both the positive and the negative diffraction orders, it is useful to position the CCDs with an offset, so that the chip gaps are not symmetric.
@@
 They pass a focussing mirror represented by a green box and get diffracted by gratings (white squares).
-They are detected by CCDs (yellow).
-Rays are colored by the diffraction order: Light purple are zero order photons.
+They are detected by CCDs (orange).
+Rays are colored by the diffraction order: Light green are zero order photons.
 A section of the Rowland torus is shown as transparent gray.
@@
 This figure is interactive in the online version, allowing the reader to pan, zoom, and rotate to see those details and inspect the rays from all angles.
-(Use the mouse to rotate, ``ctrl'' and mouse to pan, mouse wheel to zoom, and double-click to set the center of rotation.
+(Use the mouse to rotate, mouse wheel to zoom, and double-click to focus on a particular element.
+Right click to navigate to the pre-defined viewpoints shown in the static figure.
 On other devices, other native controls might be available, e.g.\ two-finger touch to zoom on touchscreens.)
@@
     \centering
-    \includegraphics[width=0.6\textwidth]{ondetector}
+    \includegraphics[width=0.9\textwidth]{ondetector}
     \caption{Distribution of photons on the detectors for different designs.
+\emph{left:} Location where photons pass the aperture and SPOs.
+\emph{right:} Location photons on a curved detector projected onto the focal plane \textbf{for a monoenergetic source}.
 Photons are colored by channel and the position of the zero(s) order(s) is highlighted with red boxes.
-Note that the dispersion and cross-dispersion directions are not to scale.
-In the lower two panels the zeroths orders of one channel overlap with the dispersed phontons from another channel.
+Note that the dispersion (along the $y$ axis) and cross-dispersion ($z$ axis) directions are not to scale.
+\textbf{Our toy example detects order -1 and +1 for the on-axis Rowland torus and orders -3, -4, and -5 for the tilted Rowland tori.} In the lower two panels the zeroths orders of one channel overlap with the dispersed photons from another channel.
 This could be avoided with larger offsets between channels in cross-dispersion direction.
 Also, in real instruments with focal lengths significantly larger than in the illustrative model we use throughout this article, the bending and the width in cross-dispersion direction are typically smaller, so that this is not a problem in practice.
+    \textbf{\emph{right:} Same as in the middle row, but for a continuum source.
+Again, photons for a specific energy are detected in several orders, but photons from other energies are detected in different locations, so the orders cannot be separated just based on their spatial location.}
         }
@@
 
-\citet{doi:10.1117/12.856482} suggested tilting the torus with respect to the optical axis when using critical-angle transmission (CAT) gratings \citep{2021hai4.book..235H,2022ApJ...934..171H}.
-In CAT gratings, the grating bars have a high aspect ratio, that is the distance between two gratings bars is much smaller than the height of the bars.
-CAT gratings are typically used at a blaze angle, where the grating bars are inclined with respect to the incoming rays.
-CAT gratings preferentially diffract photons to twice the blaze angle and can reach high diffraction efficiencies.
+\begin{figure}
+    \centering
+    \includegraphics[width=0.4\textwidth]{blazed}
+    \caption{\textbf{Concept of a blazed grating.
+The black grating is perpendicular to the incoming rays, the black grating is blazed by and angle $\gamma$ with respect to the incoming rays.
+The most likely direction for the diffracted photons is near twice the blaze angle.}}
+    \label{fig:blazed}
+\end{figure}
+
+XMM-Newton uses a Rowland torus tilted with respect to the optical axis \citep{2010SSRv..157...15P}; \citet{2010SPIE.7732E..1JH} suggested using the same geometry for Critical Angle Transmission (CAT) gratings.
+In CAT gratings \citep{2021hai4.book..235H,2022ApJ...934..171H}, the grating bars have a high aspect ratio, that is the distance between two gratings bars is much smaller than the height of the bars.
+CAT gratings are used at a blaze angle, where the grating bars are inclined with respect to the incoming rays \textbf{(see figure~\ref{fig:blazed})}.
+CAT gratings preferentially diffract photons to twice the blaze angle and can reach high diffraction efficiencies: \citet{2022ApJ...934..171H} measured $>20$\% summed over orders 4-7 at 1.8~keV.
 However, the blaze angle also means that the gratings can deviate significantly from the surface of the  Rowland torus.
@@
 $$r = \frac{a}{2\cos \alpha}.$$
- If the focal point F is the origin of the coordinate system, then the center of the Rowland circle (blue square in left blue circle in the panels in figure~\ref{fig:sketch_label}) is at location
+ If the focal point F is the origin of the coordinate system, then the center of the Rowland circle (blue square in left blue circle in figure~\ref{fig:sketch_label} and \ref{fig:sketch}) is at location
  \begin{equation}
     \begin{pmatrix} x \\ y \\ z \end{pmatrix} =
-    \begin{pmatrix} \frac{a}{2} \tan\alpha \\0 \\   \frac{a}{2} \end{pmatrix}.
+    \begin{pmatrix} \frac{a}{2} \\ \frac{a}{2}  \tan\alpha \\ 0\end{pmatrix}.
  \end{equation}
 
-(In this coordinate system the astrophysical source is located at $z=+\infty$.)
+(In this coordinate system the astrophysical source is located at $x=+\infty$.)
 Since A, F, and the Hinge are by definition all on the Rowland circle, their distance to this point is $r$.
-With this symmetry, we can obtain the angle of the axis of the torus with respect to the $x$-axis.
+With this symmetry, we can obtain the angle of the axis of the torus with respect to the $y$-axis.
 This angle is $\beta-\alpha$.
@@
     \begin{pmatrix} x \\ y \\ z \end{pmatrix} =
-    \begin{pmatrix} \frac{a}{2}\tan{\alpha}+R\sin{(\beta-\alpha)} \\ 0 \\
-        \frac{a}{2}-R\cos{(\beta-\alpha)} \end{pmatrix}.
+    \begin{pmatrix}
+        \frac{a}{2}-R\cos{(\beta-\alpha)} \\
+        \frac{a}{2}\tan{\alpha}+R\sin{(\beta-\alpha)}
+        \\ 0
+        \end{pmatrix}.
 \end{equation}
 
-Assuming CAT gratings where the gratings bars are perpendicular to the surface of the grating, the gratings need to be installed blazed (tilted) with respect to the incoming rays.
+%Assuming CAT gratings where the gratings bars are perpendicular to the surface of the grating, the gratings need to be installed blazed (tilted) with respect to the incoming rays.
 Thus, they deviate more from the Rowland-torus than a grating that is placed perpendicular to the incoming rays.
-Figure~\ref{fig:3d:tilted} shows this layout in 3D, but with a very large blaze angle for clarity; for real CAT gratings typical blaze angles are 1-2~deg.
-Tilting the Rowland torus by about twice the blaze angle reduces the average deviation of the grating bar position from the surface of the Rowland torus.
-Also, CAT gratings disperse predominantly in one direction.
-However, the Rowland geometry optimizes the spectral resolving power by minimizing the spread of the photons in dispersion direction; detectors are located on the Rowland circle, which represents the position of minimal spread in dispersion direction.
-This is not the same location as the imaging focal plane, which is defined by minimizing the total size of the PSF.
+For real CAT gratings typical blaze angles are 1-2~deg.
+Figure~\ref{fig:3d:tilted} shows this layout in 3D, but with a very large blaze angle for clarity.
+Tilting the Rowland torus by about twice the blaze angle reduces the average deviation of the grating position from the surface of the Rowland torus.
+%Also, CAT gratings disperse predominantly in one direction. ]
+The Rowland geometry optimizes the spectral resolving power by minimizing the spread of the photons in dispersion direction; detectors are located on the Rowland circle, which represents the position of minimal spread in dispersion direction.
+This is not the same location as the imaging focal plane, which is defined by minimizing the total size of the 2D PSF.
 At the focal point, the Rowland circle and the imaging focal plane match.
-Figure~\ref{fig:fish} shows how the dispersed photons spread out in cross-dispersion direction with increasing distance from F, roughly resembling the outline of a fish.
+Figure~\ref{fig:fish} \textbf{(right column)} shows how the dispersed photons spread out in cross-dispersion direction with increasing distance from F, roughly resembling the outline of a fish.
 With a tilted Rowland torus, there is a second intersection point between the spectral and the imaging focus where the cross-dispersion width becomes small (the point between the body and the tail of the fish, the ``Hinge'', about +300~mm in dispersion direction in the second panel of figure~\ref{fig:fish}).
@@
 
-A tilted Rowland torus can also be combined with sub-aperturing to increase the spectral resolving power as discussed in figure~\ref{fig:subaperture}.
+A tilted Rowland torus can also be combined with sub-aperturing. % to increase the spectral resolving power as discussed in figure~\ref{fig:subaperture}.
 In this case, not all the fish shape is filled with photons.
-This can be seen by comparing the dark purple points in the second, third, and fourth panel in figure~\ref{fig:fish}.
-The Rowland torus for this channel has the same tilt and dimensions in all three designs, but in the second panel, the mirror fills the full circle, in the third panel it has just two wedges, and in the fourth panel just one.
+This can be seen by comparing the dark purple points in the second, third, and fourth panel in figure~\ref{fig:fish}, \textbf{right column}.
+The Rowland torus for this channel has the same tilt and dimensions in all three designs, but in the second panel, the mirror fills the full circle, in the third panel it has just two rectangles, and in the fourth panel just one.
 In either case, the spectral extraction region used in data analysis should take this shape into account to minimize detector background.
@@
 
-\begin{figure}
-    \begin{interactive}{js}{ifigs/double.zip}
-    \includegraphics[width=0.22\textwidth]{ifigs/double_overview}
-    \includegraphics[width=0.22\textwidth]{ifigs/double_aperture}
-    \includegraphics[width=0.22\textwidth]{ifigs/double_gratings}
-    \includegraphics[width=0.22\textwidth]{ifigs/double_detector}
-    \end{interactive}
-  \caption{Ray-trace like figure~\ref{fig:3d:tilted}, but with channels split on two separately tilted Rowland tori.
+%\begin{figure}
+%    \begin{interactive}{js}{ifigs/double.zip}
+%    \includegraphics[width=0.22\textwidth]{ifigs/double_overview}
+%    \includegraphics[width=0.22\textwidth]{ifigs/double_aperture}
+%    \includegraphics[width=0.22\textwidth]{ifigs/double_gratings}
+%    \includegraphics[width=0.22\textwidth]{ifigs/double_detector}
+%    \end{interactive}
+%  \caption{Ray-trace like figure~\ref{fig:3d:tilted}, but with channels split on two separately tilted Rowland tori.
 See figure~\ref{fig:3d:single} for an explanation of the elements shown in the image and the interactivity in the online version.}
-  \label{fig:3d:tilted_double}
-\end{figure}
+%  \label{fig:3d:tilted_double}
+%\end{figure}
 
@@
 The second channel also adds redundancy in case of detector failure, and can compensate for chip gaps if the position of the detector elements on one channels is moved along the Rowland circle a little bit with respect to the other channel.
-Figure~\ref{fig:3d:tilted_double} shows a ray-trace for this configuration.
+%Figure~\ref{fig:3d:tilted_double} shows a ray-trace for this configuration.
 
@@
 The next two subsections explain how the channels can be offset in both dispersion and cross-dispersion direction to reduce alignment requirements and mitigate chip gaps.
-Figure~\ref{fig:3d:tilted_double_offset} shows a 3D ray-trace image in this configuration; the resulting signal on the detector is shown in figure~\ref{fig:fish} (bottom panel).
+Figure~\ref{fig:3d:tilted_double_offset} shows a 3D ray-trace image in this configuration; the resulting signal on the detector is shown in figure~\ref{fig:fish} (bottom row).
 
@@
 To do that, it is necessary to know which optical channel a photon passed through and thus the setup needs to avoid overlapping photons from both channels on the same location of the detector.
-This can be accomplished by shifting the Rowland tori (and mirrors and gratings positioned on them) in the cross-dispersion direction (into or out of the plane of the plot in figurs~\ref{fig:sketch_label} and~\ref{fig:sketch}).
+This can be accomplished by shifting the Rowland tori (and mirrors and gratings positioned on them) in the cross-dispersion direction (into or out of the plane of the plot in figures~\ref{fig:sketch_label} and~\ref{fig:sketch}, i.e.\ the $z$ direction of our coordinate system).
 The shift should be large enough to clearly separate the two channels, but keep both dispersed spectral traces on the detector with margin (figure~\ref{fig:fish}).
@@
 
-While it is possible to have a mirror that spans the entire circle for each of the two channels, the DTRS layout is particularly well-suited for sub-aperturing.
-In this case, each channel will have two opposing wedges filled with mirror elements and gratings leading to a total of four mirror and grating wedges.
-These can then be arranged closely together (see figure~\ref{fig:3d:tilted_double}) to maximize the area used given the size constraints of the spacecraft.
+%While it is possible to have a mirror that spans the entire circle for each of the two channels,
+The DTRS layout is particularly well-suited for sub-aperturing.
+In this case, each channel will have two opposing rectangles filled with mirror elements and gratings leading to a total of four mirror and grating rectangles.
+These can then be arranged closely together (see figure~\ref{fig:fish}) to maximize the area used given the size constraints of the spacecraft.
 In this case, it is an option to split each of the two channels into two sub-channels with their own separate offset, such that there are two parallel traces running in each direction (Figure~\ref{fig:3d:tilted_double_offset}) for a total of four traces (figure~\ref{fig:fish}, bottom panel).
@@
 \subsubsection{Channel offsets in the dispersion direction}
-When each channel is split into two sub-channels as described above and each of the sub-channels has its own Rowland torus (not shown in the figures because four overlapping tori are hard to visually separate), the two sub-channels can also be offset in the dispersion direction.
+When each channel is split into two sub-channels as described above and each of the sub-channels has its own Rowland torus, %(not shown in the figures because four overlapping tori are hard to visually separate),
+the two sub-channels can also be offset in the dispersion direction.
 This causes the Rowland torus that defines the grating positions to deviate from the torus that sets the position of the detectors, but if the offset is small, the impact on the spectral resolving power is negligible.
-A benefit of such an offset in dispersion direction is that the dispersed spectra of the two sub-channels do not hit chip gaps at the same wavelength, spreading out the effect of chip gaps on the extracted spectrum without the need to dither the spacecraft.
+A benefit of such an offset in dispersion direction is that the dispersed spectra of the two sub-channels do not hit chip gaps at the same wavelength.
+%, spreading out the effect of chip gaps on the extracted spectrum without the need to dither the spacecraft.
 
@@
 The DTRS design allows us to compensate for this by having multiple channels, ultimately delivering better resolving power at the same mirror requirements than a single-channel design.
+The independent mirrors in a DTRS design can be arranged on the front-end of the spacecraft such that they take up a similar geometric area as a mirror filling the full circle in a single channel design.
 
-The independent mirror wedges in a DTRS design can be arranged on the front-end of the spacecraft such that they take up a similar geometric area as a mirror filling the full circle in a single channel design.
+An alternative design is to use a single, circular mirror and sub-aperture by dispersing light from different parts of the mirror to different, independent detectors rotated in the focal plane \citep{2019JATIS...5b1003G}.
 
@@
 Because the DTRS concept spreads the dispersed photons over a larger detector area than a single-channel design, the background contamination in the spectra is higher, simply because more pixels contribute to the extracted science spectrum.
-This does not matter for bright isolated point-source, but makes observations of faint sources, where the dispersed spectrum has count rates comparable to the background, difficult.
+This does not matter for a bright isolated point-source, but makes observations of faint sources, where the dispersed spectrum has count rates comparable to the background, difficult.
 Similarly, observations are difficult when extended emission or many point sources are in the field-of-view, for example in the galactic center or in star forming regions.
@@
 Note that the dispersion and cross-dispersion directions are scaled differently.
+The cross-dispersion direction is parallel to the coordinate system $z$ axis.
+The dispersion direction is measured along the curved Rowland circle; at any given location on this plot is within 5~deg of the coordinate system $y$ axis.
         }
@@
 Note that the plot shows an extremely distorted view of the CCD plane.
-The $x$-axis extends over 1.2 m, while the $y$-axis covers only 2.5 cm.
+The dispersion axis extends over 1.2~m, while the cross-dispersion axis covers only 2.5 cm.
 
@@
 
-\begin{figure*}
-    \plotone{EQAllDet}
-    \caption{Images of the 16 individual Arcus CCDs are shown for the same simulation as in figure~\ref{fig:Arcusfull}. Note that the color scale is chosen differently for each chip.
-        }
-    \label{fig:Arcusall}
-\end{figure*}
+%\begin{figure*}
+%    \plotone{EQAllDet}
+%    \caption{Images of the 16 individual Arcus CCDs are shown for the same %simulation as in figure~\ref{fig:Arcusfull}. Note that the color scale is %chosen differently for each chip.
+%        }
+%    \label{fig:Arcusall}
+%\end{figure*}
 
-Figure~\ref{fig:Arcusall} shows the same simulation, but with each CCD to scale.
+% Figure~\ref{fig:Arcusall} shows the same simulation, but with each CCD to scale.
 The color scale is different for each CCD though, to bring out details on CCDs that do not contain the bright zeroths orders or extremely bright emission lines.

### 2024-08-20 papers/Doubletiltconcept/doubletiltconcept.tex
doubletiltconcept - astroph submission

@@
-%% Beginning of file 'sample631.tex'
-%%
-%% Modified 2021 March
-%%
-%% This is a sample manuscript marked up using the
-%% AASTeX v6.31 LaTeX 2e macros.
-%%
-%%
-%% using aastex version 6.3
-\documentclass[linenumbers]{aastex631}
+%\documentclass[linenumbers]{aastex631}
+\documentclass[twocolumn]{aastex631}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\graphicspath{{./}{figures/}}
+\graphicspath{{./}{figures/}{ifigs/}}
 %% This is the end of the preamble.  Indicate the beginning of the
@@
 The symmetry axis of the torus  is marked as blue long-dashed line.
-\textbf{The torus is formed by rotating the blue circle around this line.
-To help visualize the 3D position of the torus blue dotted lines are shown connecting the two circles.} $\alpha$ is the angle between the optical axis and the line from the point where the optical axis intersects the Rowland torus (A) to the center of the Rowland circle.
+The torus is formed by rotating the blue circle around this line.
+To help visualize the 3D position of the torus blue dotted lines are shown connecting the two circles.
+$\alpha$ is the angle between the optical axis and the line from the point where the optical axis intersects the Rowland torus (A) to the center of the Rowland circle.
 $\beta$ is the angle from optical axis to the line connecting A to the ``hinge'' -- the other point in this plane where symmetry axis of the torus and the Rowland circle intersect.
@@
   \begin{interactive}{js}{ifigs/on-axis.zip}
-  \includegraphics[width=0.22\textwidth]{ifigs/on-axis_overview}
-  \includegraphics[width=0.22\textwidth]{ifigs/on-axis_aperture}
-  \includegraphics[width=0.22\textwidth]{ifigs/on-axis_gratings}
-  \includegraphics[width=0.22\textwidth]{ifigs/on-axis_detector}
+  \includegraphics[width=0.22\textwidth]{on-axis_overview}
+  \includegraphics[width=0.22\textwidth]{on-axis_aperture}
+  \includegraphics[width=0.22\textwidth]{on-axis_gratings}
+  \includegraphics[width=0.22\textwidth]{on-axis_detector}
   \end{interactive}
@@
 
-\begin{figure}
+\begin{figure*}
     \centering
@@
 \emph{left:} Location where photons pass the aperture and SPOs.
-\emph{right:} Location photons on a curved detector projected onto the focal plane \textbf{for a monoenergetic source}.
+\emph{right:} Location photons on a curved detector projected onto the focal plane for a monoenergetic source.
 Photons are colored by channel and the position of the zero(s) order(s) is highlighted with red boxes.
 Note that the dispersion (along the $y$ axis) and cross-dispersion ($z$ axis) directions are not to scale.
-\textbf{Our toy example detects order -1 and +1 for the on-axis Rowland torus and orders -3, -4, and -5 for the tilted Rowland tori.} In the lower two panels the zeroths orders of one channel overlap with the dispersed photons from another channel.
+Our toy example detects order -1 and +1 for the on-axis Rowland torus and orders -3, -4, and -5 for the tilted Rowland tori.
+In the lower two panels the zeroths orders of one channel overlap with the dispersed photons from another channel.
 This could be avoided with larger offsets between channels in cross-dispersion direction.
 Also, in real instruments with focal lengths significantly larger than in the illustrative model we use throughout this article, the bending and the width in cross-dispersion direction are typically smaller, so that this is not a problem in practice.
-    \textbf{\emph{right:} Same as in the middle row, but for a continuum source.
-Again, photons for a specific energy are detected in several orders, but photons from other energies are detected in different locations, so the orders cannot be separated just based on their spatial location.}
+    \emph{right:} Same as in the middle row, but for a continuum source.
+Again, photons for a specific energy are detected in several orders, but photons from other energies are detected in different locations, so the orders cannot be separated just based on their spatial location.
         }
     \label{fig:fish}
-\end{figure}
+\end{figure*}
 
@@
     \includegraphics[width=0.4\textwidth]{blazed}
-    \caption{\textbf{Concept of a blazed grating.
+    \caption{Concept of a blazed grating.
 The black grating is perpendicular to the incoming rays, the black grating is blazed by and angle $\gamma$ with respect to the incoming rays.
-The most likely direction for the diffracted photons is near twice the blaze angle.}}
+The most likely direction for the diffracted photons is near twice the blaze angle.}
     \label{fig:blazed}
@@
 In CAT gratings \citep{2021hai4.book..235H,2022ApJ...934..171H}, the grating bars have a high aspect ratio, that is the distance between two gratings bars is much smaller than the height of the bars.
-CAT gratings are used at a blaze angle, where the grating bars are inclined with respect to the incoming rays \textbf{(see figure~\ref{fig:blazed})}.
+CAT gratings are used at a blaze angle, where the grating bars are inclined with respect to the incoming rays (see figure~\ref{fig:blazed}).
 CAT gratings preferentially diffract photons to twice the blaze angle and can reach high diffraction efficiencies: \citet{2022ApJ...934..171H} measured $>20$\% summed over orders 4-7 at 1.8~keV.
@@
 At the focal point, the Rowland circle and the imaging focal plane match.
-Figure~\ref{fig:fish} \textbf{(right column)} shows how the dispersed photons spread out in cross-dispersion direction with increasing distance from F, roughly resembling the outline of a fish.
+Figure~\ref{fig:fish} (right column) shows how the dispersed photons spread out in cross-dispersion direction with increasing distance from F, roughly resembling the outline of a fish.
 With a tilted Rowland torus, there is a second intersection point between the spectral and the imaging focus where the cross-dispersion width becomes small (the point between the body and the tail of the fish, the ``Hinge'', about +300~mm in dispersion direction in the second panel of figure~\ref{fig:fish}).
@@
     \begin{interactive}{js}{ifigs/tilted.zip}
-    \includegraphics[width=0.22\textwidth]{ifigs/tilted_overview}
-    \includegraphics[width=0.22\textwidth]{ifigs/tilted_aperture}
-    \includegraphics[width=0.22\textwidth]{ifigs/tilted_gratings}
-    \includegraphics[width=0.22\textwidth]{ifigs/tilted_detector}
+    \includegraphics[width=0.22\textwidth]{tilted_overview}
+    \includegraphics[width=0.22\textwidth]{tilted_aperture}
+    \includegraphics[width=0.22\textwidth]{tilted_gratings}
+    \includegraphics[width=0.22\textwidth]{tilted_detector}
     \end{interactive}
@@
 In this case, not all the fish shape is filled with photons.
-This can be seen by comparing the dark purple points in the second, third, and fourth panel in figure~\ref{fig:fish}, \textbf{right column}.
+This can be seen by comparing the dark purple points in the second, third, and fourth panel in figure~\ref{fig:fish}, right column.
 The Rowland torus for this channel has the same tilt and dimensions in all three designs, but in the second panel, the mirror fills the full circle, in the third panel it has just two rectangles, and in the fourth panel just one.
@@
 \subsection{Double tilted Rowland Torus}
-
-%\begin{figure}
-%    \begin{interactive}{js}{ifigs/double.zip}
-%    \includegraphics[width=0.22\textwidth]{ifigs/double_overview}
-%    \includegraphics[width=0.22\textwidth]{ifigs/double_aperture}
-%    \includegraphics[width=0.22\textwidth]{ifigs/double_gratings}
-%    \includegraphics[width=0.22\textwidth]{ifigs/double_detector}
-%    \end{interactive}
-%  \caption{Ray-trace like figure~\ref{fig:3d:tilted}, but with channels split on two separately tilted Rowland tori.
-See figure~\ref{fig:3d:single} for an explanation of the elements shown in the image and the interactivity in the online version.}
-%  \label{fig:3d:tilted_double}
-%\end{figure}
 
@@
     \begin{interactive}{js}{ifigs/offset.zip}
-    \includegraphics[width=0.22\textwidth]{ifigs/offset_overview}
-    \includegraphics[width=0.22\textwidth]{ifigs/offset_aperture}
-    \includegraphics[width=0.22\textwidth]{ifigs/offset_gratings}
-    \includegraphics[width=0.22\textwidth]{ifigs/offset_detector}
+    \includegraphics[width=0.22\textwidth]{offset_overview}
+    \includegraphics[width=0.22\textwidth]{offset_aperture}
+    \includegraphics[width=0.22\textwidth]{offset_gratings}
+    \includegraphics[width=0.22\textwidth]{offset_detector}
     \end{interactive}
